---
title: "Gene level differential expression analysis on IO360: MUT vs WT (part 2 of 2)"
---

# TALAVE IO360 expression analysis

## ---- Setup ----

### Notebook Options
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Project Bootstrap
```{r bootstrap}
source(file.path(root_dir, "functions", "bootstrap.R"))
initialize_project()

source(file.path(root_dir, "functions", "plotting.R"))
source(file.path(source_dir, "analysis_io360.R"))
source(file.path(source_dir, "stats_utils.R"))
source(file.path(source_dir, "io.R"))

pt.cols <- load_pt_colors()

for (d in c("gene_plot", "cg/volcano", "cg/heatmap", "cg/barplot",
            "ddr/volcano", "ddr/heatmap", "all/heatmap", "all/volcano")) {
  dir.create(file.path(io360_fig_dir, d), recursive = TRUE, showWarnings = FALSE)
}

pdf <- function(...){
  Cairo::CairoPDF(...)
}

ggsave <- function(filename, plot = ggplot2::last_plot(), device = NULL, ...) {

  if (is.null(device)) {
    if (grepl("\\.pdf$", filename, ignore.case = TRUE)) {
      device <- grDevices::cairo_pdf
    }
  }

  ggplot2::ggsave(filename = filename, plot = plot, device = device, ...)
}
```

### Load normalized data

```{r}
load_io360_expression_data() # df.io360, sum.df.io360, all.gns (and other io360_expression_input_data.rda objects)
```

## ---- Gene-Level Modeling ----
### Per-Gene Expression Plots
```{r}
for(gn in all.gns){
  tmp <- sum.df.io360 %>%
    select(PT,Patient.ID,TimePoint,bt3,BOR,!!sym(gn)) %>%
    rename(expr = all_of(gn))
  p <- plot_ct(tmp, y_var = "expr", 
               y_label = "log2(expression)",
    title = paste0(gn, " (IO360)"),
    pt.cols = pt.cols)$plot
  
  ggsave(file.path(io360_fig_dir,"gene_plot",paste0(gsub("/","-",gn),".pdf")),p,width=5,height=7,units="in",dpi=300)
}
```
### Mixed-Effects Modeling
#### Long-Format Input Table
```{r}
expr_long <- sum.df.io360 %>%
  select(-Subtype,-PFS_days) %>%
  pivot_longer(
    cols = -c(PT, Patient.ID, TimePoint, bt3, BOR, BRCA),
    names_to = "gene",
    values_to = "expr",
    values_drop_na = TRUE
  ) %>%
  mutate(
    Patient.ID = factor(Patient.ID),
    TimePoint  = factor(TimePoint, levels = c("BS", "BX2", "BX3")),
    bt3        = factor(bt3),
    BOR        = factor(BOR)
  )

```

Fits a per-BRCA-group model (not joint), since every contrast is within-group. Precomputed
and saved to `io360_gene_contrast_results.rds`; delete the file to force recomputation.
```{r}
io360_gene_contrast_results_file <- file.path(io360_obj_dir, "io360_gene_contrast_results.rds")

if (file.exists(io360_gene_contrast_results_file)) {
  all_res <- readRDS(io360_gene_contrast_results_file)
} else {
  all_res <- expr_long %>%
    group_by(gene) %>%
    group_modify(~ {
      df_gene <- prep_tmp(.x)
      bind_rows(
        fit_lmer_contrasts(df_gene, "WT"),
        fit_lmer_contrasts(df_gene, "MUT")
      )
    }) %>%
    ungroup() %>%
    group_by(BRCA, contrast) %>%
    mutate(FDR = p.adjust(p.value, method = "BH")) %>%
    ungroup()

  saveRDS(all_res, io360_gene_contrast_results_file)
}

volcano_BX2 <- all_res %>%
  filter(contrast == "BX2 - BS") %>%
  mutate(neglog10p = -log10(p.value))

volcano_BX3 <- all_res %>%
  filter(contrast == "BX3 - BS") %>%
  mutate(neglog10p = -log10(p.value))
```

### Model Diagnostics: Normality Check (pooled across all genes, standardized)

Residuals are standardized per gene/BRCA-group before pooling, to avoid a scale-mixing
artifact that would otherwise look leptokurtic regardless of true normality.
```{r}
diag_genes <- unique(expr_long$gene)

diag_residuals_list <- lapply(diag_genes, function(g) {
  d <- expr_long %>% filter(gene == g) %>% prep_tmp()
  list(
    get_gene_lmer_residuals(d, "WT"),
    get_gene_lmer_residuals(d, "MUT")
  )
}) %>% unlist(recursive = FALSE)

diag_residuals_std <- lapply(diag_residuals_list, function(r) {
  if (is.null(r) || length(r) < 2) return(NULL)
  s <- sd(r)
  if (!is.finite(s) || s == 0) return(NULL)
  r / s
}) %>% unlist(use.names = FALSE)

cat("n genes:", length(diag_genes), " | n pooled standardized residuals:", length(diag_residuals_std), "\n")

qqnorm(diag_residuals_std, main = "Q-Q plot of pooled residuals: IO360 (all genes, standardized per gene)")
qqline(diag_residuals_std)
```

Shapiro-Wilk run across 20 subsamples of n=1000 rather than once on the full set, since the
test becomes oversensitive at large n; W is the primary statistic.
```{r}
set.seed(1)
n_draws <- 20
draw_size <- 1000

shapiro_summary <- run_shapiro_draws(diag_residuals_std, draw_size, n_draws)

cat(sprintf(
  "Shapiro-Wilk across %d draws of n=%d:\n  W: median=%.4f, range=[%.4f, %.4f]\n  p: median=%.2e, range=[%.2e, %.2e]\n",
  n_draws, draw_size,
  median(shapiro_summary$W), min(shapiro_summary$W), max(shapiro_summary$W),
  median(shapiro_summary$p), min(shapiro_summary$p), max(shapiro_summary$p)
))

shapiro_summary
```

```{r}
eff_mat_BX2 <- all_res %>%
  filter(contrast == "BX2 - BS") %>%
  select(gene, BRCA, estimate) %>%
  pivot_wider(names_from = BRCA, values_from = estimate) %>%
  tibble::column_to_rownames("gene") %>%
  as.matrix()

eff_mat_BX3 <- all_res %>%
  filter(contrast == "BX3 - BS") %>%
  select(gene, BRCA, estimate) %>%
  pivot_wider(names_from = BRCA, values_from = estimate) %>%
  tibble::column_to_rownames("gene") %>%
  as.matrix()
```

#### Effect Matrices
```{r}
common_genes <- intersect(rownames(eff_mat_BX2), rownames(eff_mat_BX3))

m2 <- eff_mat_BX2[common_genes, , drop = FALSE]
m3 <- eff_mat_BX3[common_genes, , drop = FALSE]

colnames(m2) <- paste0(colnames(m2), "-BX2")
colnames(m3) <- paste0(colnames(m3), "-BX3")

eff_mat_all <- cbind(m2, m3)

eff_mat_all <- eff_mat_all[, c("MUT-BX2", "MUT-BX3", "WT-BX2", "WT-BX3")]
```

```{r}
p2 <- all_res %>%
  filter(contrast == "BX2 - BS") %>%
  select(gene, BRCA, p.value) %>%
  pivot_wider(names_from = BRCA, values_from = p.value) %>%
  column_to_rownames("gene") %>%
  as.matrix()

p3 <- all_res %>%
  filter(contrast == "BX3 - BS") %>%
  select(gene, BRCA, p.value) %>%
  pivot_wider(names_from = BRCA, values_from = p.value) %>%
  column_to_rownames("gene") %>%
  as.matrix()

p2 <- p2[common_genes, , drop = FALSE]
p3 <- p3[common_genes, , drop = FALSE]

colnames(p2) <- paste0(colnames(p2), "-BX2")
colnames(p3) <- paste0(colnames(p3), "-BX3")

pval_mat_all <- cbind(p2, p3)
pval_mat_all <- pval_mat_all[, c("MUT-BX2", "MUT-BX3", "WT-BX2", "WT-BX3")]
```

```{r}
f2 <- all_res %>%
  filter(contrast == "BX2 - BS") %>%
  select(gene, BRCA, FDR) %>%
  pivot_wider(names_from = BRCA, values_from = FDR) %>%
  column_to_rownames("gene") %>%
  as.matrix()

f3 <- all_res %>%
  filter(contrast == "BX3 - BS") %>%
  select(gene, BRCA, FDR) %>%
  pivot_wider(names_from = BRCA, values_from = FDR) %>%
  column_to_rownames("gene") %>%
  as.matrix()

f2 <- f2[common_genes, , drop = FALSE]
f3 <- f3[common_genes, , drop = FALSE]

colnames(f2) <- paste0(colnames(f2), "-BX2")
colnames(f3) <- paste0(colnames(f3), "-BX3")

FDR_mat_all <- cbind(f2, f3)
FDR_mat_all <- FDR_mat_all[, c("MUT-BX2", "MUT-BX3", "WT-BX2", "WT-BX3")]
```

## ---- Targeted Gene Programs ----
### cGAS-STING Genes
```{r}
sig_df <- readRDS(file.path(io360_obj_dir, "sting_from_literature.rds"))

gene_col <- names(sig_df)[str_detect(tolower(names(sig_df)), "gene")][1]
set_col  <- names(sig_df)[str_detect(tolower(names(sig_df)), "ref")][1]

stopifnot(!is.na(gene_col), !is.na(set_col))

sig_df2 <- sig_df %>%
  transmute(
    gene = as.character(.data[[gene_col]]),
    set  = as.character(.data[[set_col]])
  ) %>%
  filter(!is.na(gene), gene != "", !is.na(set), set != "") %>%
  mutate(
    gene = str_trim(gene),
    set  = str_trim(set)
  )

cg_sting_genes <- sig_df2 %>%
  pull(gene) %>% unique()

cg_genes_use <- intersect(cg_sting_genes, rownames(eff_mat_all))

length(cg_genes_use)  # sanity check

eff_mat_cg <- eff_mat_all[cg_genes_use, , drop = FALSE]

max_abs <- max(abs(eff_mat_cg), na.rm = TRUE)
breaks_seq <- seq(-max_abs, max_abs, length.out = 101)
```

#### [Fig 3AB] cGAS-STING Volcano Plot

Plotted against nominal p.
```{r}
volcano_BX2_cg <- volcano_BX2 %>% filter(gene %in% cg_genes_use)
volcano_BX3_cg <- volcano_BX3 %>% filter(gene %in% cg_genes_use)

p_v_BX2 <- make_volcano(volcano_BX2_cg, title = "BX2 (vs BS)", n_label = 20, metric = "p.value")
p_v_BX3 <- make_volcano(volcano_BX3_cg, title = "BX3 (vs BS)", n_label = 20, metric = "p.value")

ymax <- max(p_v_BX2$data$neglog10p, p_v_BX3$data$neglog10p, na.rm = TRUE)
xmin <- min(p_v_BX2$data$estimate, p_v_BX3$data$estimate, na.rm = TRUE)
xmax <- max(p_v_BX2$data$estimate, p_v_BX3$data$estimate, na.rm = TRUE)

p_v_BX2 <- p_v_BX2 + coord_cartesian(xlim = c(xmin, xmax), ylim = c(0, ymax))
p_v_BX3 <- p_v_BX3 + coord_cartesian(xlim = c(xmin, xmax), ylim = c(0, ymax))

ggsave(file.path(io360_fig_dir, "cg", "volcano", "volcano_BX2_cg_pval.pdf"), p_v_BX2, width = 8, height = 3.8)
ggsave(file.path(io360_fig_dir, "cg", "volcano", "volcano_BX3_cg_pval.pdf"), p_v_BX3, width = 8, height = 3.8)

p_v_BX2_cg <- p_v_BX2
p_v_BX3_cg <- p_v_BX3
```

#### [Fig 3C] cGAS-STING Heatmap
```{r}
genes_hm <- rownames(eff_mat_cg)

memb_long <- sig_df2 %>%
  mutate(set_lower = tolower(set)) %>%
  filter(gene %in% genes_hm) %>%
  transmute(
    gene,
    Abe    = str_detect(set_lower, "abe"),
    Dunphy = str_detect(set_lower, "dunphy"),
    Xia    = str_detect(set_lower, "xia")
  ) %>%
  group_by(gene) %>%
  summarise(
    Abe    = any(Abe),
    Dunphy = any(Dunphy),
    Xia    = any(Xia),
    .groups = "drop"
  )

memb_wide <- tibble(gene = genes_hm) %>%
  left_join(memb_long, by = "gene") %>%
  mutate(across(c(Abe, Dunphy, Xia), ~ tidyr::replace_na(.x, FALSE))) %>%
  mutate(across(c(Abe, Dunphy, Xia), ~ factor(ifelse(.x, "Yes", "No"), levels = c("No","Yes")))) %>%
  tibble::column_to_rownames("gene")

FDR_mat_cg  <- FDR_mat_all[cg_genes_use, , drop = FALSE]
pval_mat_cg <- pval_mat_all[cg_genes_use, , drop = FALSE]
```

Two-tier significance: FDR<0.05 (black) vs nominal p<0.05 only (grey).
```{r}
sig_cut <- 0.05
mat_tier_cg <- matrix("none", nrow = nrow(FDR_mat_cg), ncol = ncol(FDR_mat_cg),
                       dimnames = dimnames(FDR_mat_cg))
mat_tier_cg[pval_mat_cg < sig_cut] <- "nominal"
mat_tier_cg[FDR_mat_cg < sig_cut] <- "fdr"
mat_tier_cg[is.na(FDR_mat_cg) | is.na(pval_mat_cg)] <- "none"

mat_val_filt <- eff_mat_cg
mat_tier_filt <- mat_tier_cg

cap <- 2.5
mat_val_filt[mat_val_filt >  cap] <-  cap
mat_val_filt[mat_val_filt < -cap] <- -cap

rownames(mat_val_filt) <- gsub("\\.", "-", rownames(mat_val_filt))

ann_colors <- list(
  Abe    = c("No" = "grey80", "Yes" = "black"),
  Dunphy = c("No" = "grey80", "Yes" = "black"),
  Xia    = c("No" = "grey80", "Yes" = "black"),
  BRCA   = c(MUT="#1b9e77", WT="#7570b3"),
  Time   = c(BX2="#fc8d62", BX3="#8da0cb")
)

max_abs <- max(abs(eff_mat_cg), na.rm = TRUE)

col_fun <- circlize::colorRamp2(
  c(-cap, 0, cap),
  c("#2166ac", "white", "#b2182b")
)

col_annot <- data.frame(
  BRCA = c("MUT","MUT","WT","WT"),
  Time = c("BX2","BX3","BX2","BX3"),
  stringsAsFactors = FALSE
)
rownames(col_annot) <- colnames(eff_mat_cg)

ha_top <- HeatmapAnnotation(
  df  = col_annot[, c("BRCA","Time"), drop = FALSE],
  col = ann_colors[c("BRCA","Time")],
  which = "column",
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 10),
  simple_anno_size = unit(3.0, "mm")
)

row_annot <- memb_wide[, c("Abe","Dunphy","Xia"), drop = FALSE]
row_annot <- row_annot[rownames(eff_mat_cg), , drop = FALSE]

ha_left <- rowAnnotation(
  df  = row_annot,
  col = ann_colors[c("Abe","Dunphy","Xia")],
  annotation_name_side = "bottom",
  annotation_name_gp = gpar(fontsize = 10),
  simple_anno_size = unit(3.0, "mm")
)

ht <- Heatmap(
  mat_val_filt,
  name = "Effect",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  top_annotation = ha_top,
  left_annotation = ha_left,
  rect_gp = gpar(col = NA),
  cell_fun = function(j, i, x, y, w, h, fill) {
    tier <- mat_tier_filt[i, j]
    ## Grey drawn first, black on top for the FDR tier.
    if (tier %in% c("nominal", "fdr")) {
      grid.rect(x = x, y = y, width = w, height = h, gp = gpar(fill = NA, col = "grey30", lwd = 1.5))
    }
    if (tier == "fdr") {
      grid.rect(x = x, y = y, width = w, height = h, gp = gpar(fill = NA, col = "black", lwd = 2))
    }
  }
)

lgd_sig_cg <- Legend(
  labels = c("FDR < 0.05", "nominal p < 0.05"),
  title = "Significance",
  type = "lines",
  legend_gp = gpar(col = c("black", "grey30"), lwd = c(2, 1.5))
)

pdf(file.path(io360_fig_dir, "cg", "heatmap", "heatmap_cg_sting.pdf"), width = 5, height = 7)
ComplexHeatmap::draw(ht, heatmap_legend_side = "right", annotation_legend_side = "right", annotation_legend_list = list(lgd_sig_cg))
dev.off()
```

### DDR Genes
```{r}
ddr_genes_use <- unique(unlist(gene.sigs[40:47]))
mmej_genes <- c("FEN1","POLQ","PARP1")

eff_mat_ddr <- eff_mat_all[ddr_genes_use, , drop = FALSE]

max_abs <- max(abs(eff_mat_ddr), na.rm = TRUE)
breaks_seq <- seq(-max_abs, max_abs, length.out = 101)
```

#### [EDF3AB] DDR Volcano Plot

Plotted against nominal p.
```{r}
volcano_BX2_ddr <- volcano_BX2 %>% filter(gene %in% ddr_genes_use)
volcano_BX3_ddr <- volcano_BX3 %>% filter(gene %in% ddr_genes_use)

highlight_genes <- c("PARP1", "FEN1", "POLQ")

p_v_BX2 <- make_volcano(
  volcano_BX2_ddr,
  title = "BX2 (vs BS)",
  highlight_genes = highlight_genes,
  n_label = 20,
  metric = "p.value"
)

p_v_BX3 <- make_volcano(
  volcano_BX3_ddr,
  title = "BX3 (vs BS)",
  highlight_genes = highlight_genes,
  n_label = 20,
  metric = "p.value"
)

ymax <- max(p_v_BX2$data$neglog10p, p_v_BX3$data$neglog10p, na.rm = TRUE)
xmin <- min(p_v_BX2$data$estimate, p_v_BX3$data$estimate, na.rm = TRUE)
xmax <- max(p_v_BX2$data$estimate, p_v_BX3$data$estimate, na.rm = TRUE)

p_v_BX2 <- p_v_BX2 + coord_cartesian(xlim = c(xmin, xmax), ylim = c(0, ymax))
p_v_BX3 <- p_v_BX3 + coord_cartesian(xlim = c(xmin, xmax), ylim = c(0, ymax))

ggsave(file.path(io360_fig_dir, "ddr", "volcano", "volcano_BX2_ddr_pval.pdf"), p_v_BX2, width = 8, height = 3.8)
ggsave(file.path(io360_fig_dir, "ddr", "volcano", "volcano_BX3_ddr_pval.pdf"), p_v_BX3, width = 8, height = 3.8)

p_v_BX2_ddr <- p_v_BX2
p_v_BX3_ddr <- p_v_BX3
```

#### [EDF3C] DDR Heatmap
```{r}
FDR_mat_ddr  <- FDR_mat_all[ddr_genes_use, , drop = FALSE]
pval_mat_ddr <- pval_mat_all[ddr_genes_use, , drop = FALSE]
```

Two-tier significance: FDR<0.05 (black) vs nominal p<0.05 only (grey).
```{r}
sig_cut <- 0.05
mat_tier_ddr <- matrix("none", nrow = nrow(FDR_mat_ddr), ncol = ncol(FDR_mat_ddr),
                        dimnames = dimnames(FDR_mat_ddr))
mat_tier_ddr[pval_mat_ddr < sig_cut] <- "nominal"
mat_tier_ddr[FDR_mat_ddr < sig_cut] <- "fdr"
mat_tier_ddr[is.na(FDR_mat_ddr) | is.na(pval_mat_ddr)] <- "none"

mat_val_filt <- eff_mat_ddr
mat_tier_filt <- mat_tier_ddr

cap <- max_abs <- max(abs(eff_mat_ddr), na.rm = TRUE)

mat_val_filt[mat_val_filt >  cap] <-  cap
mat_val_filt[mat_val_filt < -cap] <- -cap

col_fun <- circlize::colorRamp2(
  c(-cap, 0, cap),
  c("#2166ac", "white", "#b2182b")  # blue-white-red (can swap)
)

rownames(mat_val_filt) <- gsub("\\.", "-", rownames(mat_val_filt))

ann_colors <- list(
  MMEJ    = c("Other" = "grey80", "MMEJ" = "black"),
  BRCA   = c(MUT="#1b9e77", WT="#7570b3"),
  Time   = c(BX2="#fc8d62", BX3="#8da0cb")
)

col_annot <- data.frame(
  BRCA = c("MUT","MUT","WT","WT"),
  Time = c("BX2","BX3","BX2","BX3"),
  stringsAsFactors = FALSE
)
rownames(col_annot) <- colnames(eff_mat_ddr)

ha_top <- HeatmapAnnotation(
  df  = col_annot[, c("BRCA","Time"), drop = FALSE],
  col = ann_colors[c("BRCA","Time")],
  which = "column",
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 10),
  simple_anno_size = unit(3.0, "mm")
)

mmej_genes <- c("PARP1","FEN1","POLQ")

row_annot <- data.frame(
  MMEJ = ifelse(rownames(eff_mat_ddr) %in% mmej_genes, "MMEJ", "Other")
)
rownames(row_annot) <- rownames(eff_mat_ddr)

ha_left <- rowAnnotation(
  df  = row_annot,
  col = ann_colors["MMEJ"],
  annotation_name_side = "bottom",
  annotation_name_gp = gpar(fontsize = 10),
  simple_anno_size = unit(3.0, "mm")
)

ht <- Heatmap(
  mat_val_filt,
  name = "Effect",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  top_annotation = ha_top,
  left_annotation = ha_left,
  rect_gp = gpar(col = NA),
  cell_fun = function(j, i, x, y, w, h, fill) {
    tier <- mat_tier_filt[i, j]
    ## Grey drawn first, black on top for the FDR tier.
    if (tier %in% c("nominal", "fdr")) {
      grid.rect(x = x, y = y, width = w, height = h, gp = gpar(fill = NA, col = "grey30", lwd = 1.5))
    }
    if (tier == "fdr") {
      grid.rect(x = x, y = y, width = w, height = h, gp = gpar(fill = NA, col = "black", lwd = 2))
    }
  }
)

lgd_sig_ddr <- Legend(
  labels = c("FDR < 0.05", "nominal p < 0.05"),
  title = "Significance",
  type = "lines",
  legend_gp = gpar(col = c("black", "grey30"), lwd = c(2, 1.5))
)

pdf(file.path(io360_fig_dir, "ddr", "heatmap", "heatmap_ddr.pdf"), width = 5, height = 7)
ComplexHeatmap::draw(ht, heatmap_legend_side = "right", annotation_legend_side = "right", annotation_legend_list = list(lgd_sig_ddr))
dev.off()
```

## ---- Pathway Enrichment ----
### MSigDB Pathway Preparation
```{r}
gns_all <- union(volcano_BX2$gene, volcano_BX3$gene) %>% unique()

msigdb <- readRDS(file.path(io360_obj_dir,"msigdb_v7.5.1_almac.rds"))

gene_mapping <- AnnotationDbi::select(org.Hs.eg.db, 
                                      keys = unique(unlist(msigdb)), 
                                      keytype = "ENTREZID", 
                                      columns = "SYMBOL")

msigdb_symbols <- lapply(msigdb, function(gene_ids) {
  gene_symbols <- gene_mapping$SYMBOL[match(gene_ids, gene_mapping$ENTREZID)]
  gene_symbols <- gene_symbols[!is.na(gene_symbols)]
  return(unique(gene_symbols))
})

db.types <- sub("_.+","",names(msigdb_symbols))
msigdb_symbols <- msigdb_symbols[db.types %in% c("HALLMARK","ALMAC","GOBP","GOCC","GOMF","KEGG","REACTOME")]

cg_sig_lists <- with(
  sig_df,
  tapply(gene, ref, function(x) unique(x[x %in% gns_all]))
)
cg_sig_lists <- c(cg_sig_lists,all=list(unique(unlist(cg_sig_lists))))

msigdb_symbols <- c(msigdb_symbols, gene.sigs, CUSTOM_CGAS_STING = cg_sig_lists)
msigdb_symbols <- lapply(msigdb_symbols, function(x) unique(x[x %in% gns_all]))
msigdb_symbols <- msigdb_symbols[sapply(msigdb_symbols, length) >= 5 & sapply(msigdb_symbols, length) <= 500]

msigdb_symbols_all <- msigdb_symbols
msigdb_symbols <- msigdb_symbols_all[sub("_.+","",names(msigdb_symbols_all)) %in%  c("NANOSTRING","HALLMARK","CUSTOM")]
```

### FGSEA Runs
#### BX2
```{r}
rank_MUT_BX2 <- make_rank_signed_logp(volcano_BX2, brca = "MUT", use = "p.value")
rank_WT_BX2  <- make_rank_signed_logp(volcano_BX2, brca = "WT",  use = "p.value")

fgsea_MUT_BX2 <- run_fgsea_rank(rank_MUT_BX2, msigdb_symbols) %>%
  mutate(Time = "BX2", BRCA = "MUT")

fgsea_WT_BX2 <- run_fgsea_rank(rank_WT_BX2, msigdb_symbols) %>%
  mutate(Time = "BX2", BRCA = "WT")
```
#### BX3
```{r}
rank_MUT_BX3 <- make_rank_signed_logp(volcano_BX3, brca = "MUT", use = "p.value")
rank_WT_BX3  <- make_rank_signed_logp(volcano_BX3, brca = "WT",  use = "p.value")

fgsea_MUT_BX3 <- run_fgsea_rank(rank_MUT_BX3, msigdb_symbols) %>%
  mutate(Time = "BX3", BRCA = "MUT")

fgsea_WT_BX3 <- run_fgsea_rank(rank_WT_BX3, msigdb_symbols) %>%
  mutate(Time = "BX3", BRCA = "WT")
```
### Top Enriched Pathways
```{r}
fgsea_BX2_combined <- combine_fgsea(fgsea_MUT_BX2, fgsea_WT_BX2, padj_cut = 0.05) %>%
  mutate(Time = "BX2")

fgsea_BX3_combined <- combine_fgsea(fgsea_MUT_BX3, fgsea_WT_BX3, padj_cut = 0.05) %>%
  mutate(Time = "BX3")

fgsea_all <- bind_rows(
  fgsea_BX2_combined %>% mutate(Time = "BX2"),
  fgsea_BX3_combined %>% mutate(Time = "BX3")
)

fgsea_long <- fgsea_all %>%
  select(Time, pathway,
         mut_NES, mut_pval, mut_padj,
         wt_NES,  wt_pval, wt_padj) %>%
  pivot_longer(
    cols = c(mut_NES, mut_pval, wt_NES, wt_pval, mut_padj, wt_padj),
    names_to = c("Group", ".value"),
    names_pattern = "(mut|wt)_(NES|pval|padj)"
  ) %>%
  mutate(
    Group = recode(Group, mut = "MUT", wt = "WT"),
    Direction = case_when(
      NES > 0 ~ "up",
      NES < 0 ~ "down",
      TRUE ~ "zero"
    ),
    Sig = pval < 0.05 & padj < 0.2
  )

pathway_df <- fgsea_long %>%
  filter(pval < 0.05,padj < 0.2, Direction %in% c("up","down")) %>%
  mutate(Class = paste(Group, Time, sep = "_")) %>%
  select(Class, Direction, pathway, NES, pval,padj)
```
### [EDF3D] cGAS-STING Custom Pathway Enrichment Barplot
```{r}
cg_paths <- unique(grep("CUSTOM_CGAS_STING", pathway_df$pathway, value = TRUE))

cg_fgsea <- fgsea_long %>%
  filter(pathway %in% cg_paths) %>%
  mutate(
    signed_logp = sign(NES) * -log10(padj),
    condition = paste(Group, Time, sep = "_")
  )

cg_fgsea$pathway <- cg_fgsea$pathway |>
  gsub("CUSTOM_CGAS_STING\\.", "", x = _) |>
  gsub(" et al", "", x = _)

cg_fgsea <- cg_fgsea %>%
  mutate(pathway = factor(pathway, levels = sort(unique(pathway))[c(2,1,3,4)]))
  

p <- ggplot(cg_fgsea,
            aes(x = pathway,
                y = signed_logp,
                fill = condition)) +
  geom_col(position = position_dodge(width = .8)) +
  geom_hline(yintercept = 0, colour = "black") +
  geom_hline(yintercept = c(-log10(0.05), log10(0.05)), linetype = "dashed", color = "grey50") +
  theme_bw() +
  labs(
    title = "cGAS–STING pathway activity across reference signatures",
    x = NULL,                     # remove x-axis title
    y = "signed -log10(FDR)",
    fill = "Condition"
  ) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.text.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )

ggsave(file.path(io360_fig_dir, "cg", "barplot", "barplot_cgas_sting.pdf"), p, width = 8, height = 4)
```

### [Fig 4A] Pathway Heatmap
```{r}
fgsea_long_all <- bind_rows(
  fgsea_MUT_BX2 %>% mutate(Group="MUT", Time="BX2"),
  fgsea_MUT_BX3 %>% mutate(Group="MUT", Time="BX3"),
  fgsea_WT_BX2  %>% mutate(Group="WT",  Time="BX2"),
  fgsea_WT_BX3  %>% mutate(Group="WT",  Time="BX3")
) %>%
  mutate(
    Class = paste(Group, Time, sep = "_"),
    signed_slp = sign(NES) * -log10(pval)
  ) %>%
  select(pathway, Class, NES, pval, padj, signed_slp)

mat_val <- fgsea_long_all %>%
  select(pathway, Class, signed_slp) %>%
  pivot_wider(names_from = Class, values_from = signed_slp) %>%
  column_to_rownames("pathway") %>%
  as.matrix()
```

Two-tier significance: FDR<0.05 (black) vs nominal p<0.05 only (grey).
```{r}
sig_cut <- 0.05
mat_tier <- fgsea_long_all %>%
  mutate(tier = case_when(
    padj < sig_cut ~ "fdr",
    pval < sig_cut ~ "nominal",
    TRUE ~ "none"
  )) %>%
  select(pathway, Class, tier) %>%
  pivot_wider(names_from = Class, values_from = tier) %>%
  column_to_rownames("pathway") %>%
  as.matrix()

col_order <- c("MUT_BX2","MUT_BX3","WT_BX2","WT_BX3")
mat_val <- mat_val[, col_order, drop = FALSE]
mat_tier <- mat_tier[, col_order, drop = FALSE]

mat_val[is.na(mat_val)] <- 0
mat_tier[is.na(mat_tier)] <- "none"

keep_rows <- rowSums(mat_tier != "none") > 0

mat_val_filt <- mat_val[keep_rows, , drop = FALSE]
mat_tier_filt <- mat_tier[keep_rows, , drop = FALSE]

cap <- 3.5
mat_val_filt[mat_val_filt >  cap] <-  cap
mat_val_filt[mat_val_filt < -cap] <- -cap

rownames(mat_val_filt) <- format_pathway_label_ascii(rownames(mat_val_filt))

ann_colors <- list(
  BRCA   = c(MUT="#1b9e77", WT="#7570b3"),
  Time   = c(BX2="#fc8d62", BX3="#8da0cb")
)

col_annot <- data.frame(
  BRCA = c("MUT","MUT","WT","WT"),
  Time = c("BX2","BX3","BX2","BX3"),
  stringsAsFactors = FALSE
)
rownames(col_annot) <- colnames(eff_mat_ddr)

ha_top <- HeatmapAnnotation(
  df  = col_annot[, c("BRCA","Time"), drop = FALSE],
  col = ann_colors[c("BRCA","Time")],
  which = "column",
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 10),
  simple_anno_size = unit(3.0, "mm")
)

row_annot <- data.frame(
  MMEJ = ifelse(rownames(eff_mat_ddr) %in% mmej_genes, "MMEJ", "Other")
)
rownames(row_annot) <- rownames(eff_mat_ddr)

col_fun <- circlize::colorRamp2(
  c(-cap, 0, cap),
  c("#2166ac", "white", "#b2182b")
)

ht <- Heatmap(
  mat_val_filt,
  name = "signed\n-log10(FDR)",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  rect_gp = gpar(col = NA),
  top_annotation = ha_top,
  cell_fun = function(j, i, x, y, w, h, fill) {
    tier <- mat_tier_filt[i, j]
    ## Grey drawn first, black on top for the FDR tier.
    if (tier %in% c("nominal", "fdr")) {
      grid.rect(x = x, y = y, width = w, height = h, gp = gpar(fill = NA, col = "grey30", lwd = 1.5))
    }
    if (tier == "fdr") {
      grid.rect(x = x, y = y, width = w, height = h, gp = gpar(fill = NA, col = "black", lwd = 2))
    }
  }
)

lgd_sig <- Legend(
  labels = c("FDR < 0.05", "nominal p < 0.05"),
  title = "Significance",
  type = "lines",
  legend_gp = gpar(col = c("black", "grey30"), lwd = c(2, 1.5))
)

pdf(file.path(io360_fig_dir, "all", "heatmap", "heatmap_pathways_signedslp_sigborder_v2.pdf"), width = 6, height = 8)
ComplexHeatmap::draw(ht, annotation_legend_list = list(lgd_sig))
dev.off()
```

## ---- Focused Pathway Follow-Up ----
```{r}
lym_genes  <- unique(msigdb_symbols$NANOSTRING_Lymphoid)
allo_genes <- unique(msigdb_symbols$HALLMARK_ALLOGRAFT_REJECTION)

mutup_genes_use <- unique(c(lym_genes, allo_genes))

eff_mat_mutup <- eff_mat_all[mutup_genes_use, , drop = FALSE]

max_abs <- max(abs(eff_mat_mutup), na.rm = TRUE)
breaks_seq <- seq(-max_abs, max_abs, length.out = 101)
```

### [Fig 4BC] MUT-Upregulated Pathway Volcano Plot
```{r}
volcano_BX2_mutup <- volcano_BX2 %>%
  filter(gene %in% mutup_genes_use)

volcano_BX3_mutup <- volcano_BX3 %>%
  filter(gene %in% mutup_genes_use)

## Same gene set as before (union of Lymphoid + Allograft Rejection), but point color now
## uses make_volcano()'s three-tier FDR/nominal/ns scheme (as in Fig.3A-B) instead of gene-set
## membership -- axis stays nominal -log10(p) (metric = "p.value"), color reflects FDR.
p_v_BX2_mutup <- make_volcano(
  volcano_BX2_mutup,
  title = "Lymphoid/Allograft: BX2 (vs BS)",
  n_label = 20,
  metric = "p.value"
)

p_v_BX3_mutup <- make_volcano(
  volcano_BX3_mutup,
  title = "Lymphoid/Allograft: BX3 (vs BS)",
  n_label = 20,
  metric = "p.value"
)
ymax <- max(
  p_v_BX2_mutup$data$neglog10p,
  p_v_BX3_mutup$data$neglog10p,
  na.rm = TRUE
)

xmin <- min(
  p_v_BX2_mutup$data$estimate,
  p_v_BX3_mutup$data$estimate,
  na.rm = TRUE
)

xmax <- max(
  p_v_BX2_mutup$data$estimate,
  p_v_BX3_mutup$data$estimate,
  na.rm = TRUE
)

p_v_BX2_mutup <- p_v_BX2_mutup +  coord_cartesian(xlim = c(xmin, xmax), ylim = c(0, ymax))
p_v_BX3_mutup <- p_v_BX3_mutup +  coord_cartesian(xlim = c(xmin, xmax), ylim = c(0, ymax))


ggsave(file.path(io360_fig_dir, "all", "volcano", "volcano_BX2_mutup_fdr.pdf"), p_v_BX2_mutup, width = 7, height = 3.8)
ggsave(file.path(io360_fig_dir, "all", "volcano", "volcano_BX3_mutup_fdr.pdf"), p_v_BX3_mutup, width = 7, height = 3.8)
```

#### Overlap Between cGAS-STING and Other Genes
```{r}
universe <- unique(unlist(msigdb_symbols))

pathway <- "HALLMARK_ALLOGRAFT_REJECTION"
cgas <- unique(msigdb_symbols[[pathway]])

gene_sets_to_test <- msigdb_symbols[names(msigdb_symbols) != pathway]

fisher_results <- map_dfr(
  names(gene_sets_to_test),
  function(gs_name) {
    
    gs <- unique(gene_sets_to_test[[gs_name]])
    
    a <- length(intersect(cgas, gs))
    b <- length(cgas) - a
    c <- length(gs) - a
    d <- length(universe) - (a + b + c)
    
    mat <- matrix(c(a, b, c, d), nrow = 2)
    
    ft <- fisher.test(mat)
    
    tibble(
      pathway = gs_name,
      overlap = a,
      odds_ratio = ft$estimate,
      pval = ft$p.value
    )
  }
)

fisher_results <- fisher_results %>%
  mutate(padj = p.adjust(pval, method = "BH")) %>%
  filter(overlap >= 3) %>%
  arrange(pval)
```

### [Fig 3D] cGAS-STING and IFNa Barplot
```{r}
path_sting <- "CUSTOM_CGAS_STING"
path_ifna  <- "HALLMARK_INTERFERON_ALPHA_RESPONSE"
path_allo <- "HALLMARK_ALLOGRAFT_REJECTION"

use_adj <- TRUE

df_sig <- bind_rows(
  extract_signed_logp(fgsea_MUT_BX2, path_sting, "MUT_BX2", use_adj),
  extract_signed_logp(fgsea_MUT_BX3, path_sting, "MUT_BX3", use_adj),
  extract_signed_logp(fgsea_WT_BX2,  path_sting, "WT_BX2",  use_adj),
  extract_signed_logp(fgsea_WT_BX3,  path_sting, "WT_BX3",  use_adj),

  extract_signed_logp(fgsea_MUT_BX2, path_ifna,  "MUT_BX2", use_adj),
  extract_signed_logp(fgsea_MUT_BX3, path_ifna,  "MUT_BX3", use_adj),
  extract_signed_logp(fgsea_WT_BX2,  path_ifna,  "WT_BX2",  use_adj),
  extract_signed_logp(fgsea_WT_BX3,  path_ifna,  "WT_BX3",  use_adj),
  
  extract_signed_logp(fgsea_MUT_BX2, path_allo,  "MUT_BX2", use_adj),
  extract_signed_logp(fgsea_MUT_BX3, path_allo,  "MUT_BX3", use_adj),
  extract_signed_logp(fgsea_WT_BX2,  path_allo,  "WT_BX2",  use_adj),
  extract_signed_logp(fgsea_WT_BX3,  path_allo,  "WT_BX3",  use_adj)
  
) %>%
  mutate(
    Signature = case_when(
      pathway == path_sting ~ "cGAS–STING signature",
      pathway == path_ifna  ~ "IFNα response (Hallmark)",
      pathway == path_allo  ~ "Allograft rejection (Hallmark)",
      TRUE ~ pathway
    ),
    Condition = factor(Condition, levels = c("MUT_BX2","MUT_BX3","WT_BX2","WT_BX3"))
  )

df_sig
```

```{r}
p_thresh <- -log10(0.05)

p_sig <- ggplot(df_sig, aes(x = Condition, y = signed_logp)) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_hline(yintercept = c(p_thresh, -p_thresh),
             linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_col(width = 0.85) +
  facet_wrap(~ Signature, ncol = 2) +
  scale_x_discrete(expand = expansion(mult = c(0.3, 0.3))) +
  labs(
    x = NULL,
    y = "Signed -log10(FDR)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 15),
    axis.title.y = element_text(size = 12),
    panel.spacing = unit(0.5, "lines")
  )

ggsave(file.path(io360_fig_dir, "cg", "barplot", "barplot_cg_sting_ifna.pdf"), p_sig, width = 4, height = 4)
```
