---
title: 'DSP Protein expression analysis (part 2 of 2)'
author: "Kenichi Shimada"
---

# TALAVE DSP Protein expression analysis

## ---- Setup ----

### Notebook Options
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Bootstrap

```{r bootstrap,message=F,warning=F}
source(file.path(root_dir, "functions", "bootstrap.R"))
initialize_project()

source(file.path(root_dir, "functions", "plotting.R"))
source(file.path(source_dir, "analysis_dsp_protein.R"))
source(file.path(source_dir, "stats_utils.R"))
source(file.path(source_dir, "io.R"))

pt.cols <- load_pt_colors()
```

### Load normalized data

```{r}
load(file.path(protein_obj_dir, "dsp_protein_expression_input_data.rda")) # df.dsp.all, all.prots
```

## ---- Protein-Level Modeling ----

### [EDF2E (Ki67), Fig 3E (STING), EDF5B (immune)] Individual Protein Profiles
```{r}
dir.create(file.path(protein_fig_dir,"gene_plot"), showWarnings = FALSE)

prot.stats <- list()

for(pr in all.prots){
  prot <- plot_ct_prot(df.dsp.all, y_var=pr, y_label=paste0("Log2(expression)"), title=paste0(gsub("\\.","-",pr), " (Protein)"), pt.cols=pt.cols, ymin=NULL)
  prot.stats[[pr]] <- prot$stat
  ggsave(file.path(protein_fig_dir,"gene_plot",paste0(gsub("[^A-Za-z0-9]","-",pr),".pdf")), prot$plot, width=9, height=7, units="in", dpi=300)
}
```


### Contrast Summary Table
```{r}
prot_long <- build_dsp_protein_contrast_table(prot.stats)
```

## ---- Model Diagnostics ----

### Normality Check: Pooled, Standardized Residuals (Multi-Draw Shapiro-Wilk)

Residuals standardized per protein before pooling; Shapiro-Wilk run at two subsample sizes
since it becomes oversensitive at large n. W is the primary statistic.
```{r}
resid_list <- lapply(all.prots, function(pr) {
  r <- get_protein_lmer_residuals(df.dsp.all, pr)
  if (is.null(r) || length(r) < 2) return(NULL)
  r / sd(r)
})
resid_list <- resid_list[!sapply(resid_list, is.null)]
pooled_resid_prot <- unlist(resid_list)

cat("n proteins:", length(resid_list), " | n pooled standardized residuals:", length(pooled_resid_prot), "\n")

qqnorm(pooled_resid_prot, main = "Q-Q plot of pooled residuals: DSP Protein (all proteins, standardized per protein)")
qqline(pooled_resid_prot)

set.seed(1)
draws_1000 <- run_shapiro_draws(pooled_resid_prot, 1000)
draws_2000 <- run_shapiro_draws(pooled_resid_prot, 2000)

summary_tbl <- tibble(
  draw_size = c(1000, 2000),
  W_median = c(median(draws_1000$W), median(draws_2000$W)),
  W_range = c(
    paste0(round(range(draws_1000$W), 4), collapse = "-"),
    paste0(round(range(draws_2000$W), 4), collapse = "-")
  ),
  p_median = c(median(draws_1000$p), median(draws_2000$p)),
  p_range = c(
    paste0(signif(range(draws_1000$p), 3), collapse = "-"),
    paste0(signif(range(draws_2000$p), 3), collapse = "-")
  )
)
summary_tbl
```

## ---- Contrast Heatmap ----

### [EDF5A] Protein Contrast Heatmap

Transposed to lay the heatmap out horizontally: proteins (many) run along columns, contrasts
(few) run along rows, so the dendrogram (protein clustering) moves to the column side and the
BRCA/Segment/Time annotation moves to the row side.
```{r}
heatmap_data <- prepare_dsp_protein_heatmap_data(prot_long)

heat_mat <- t(heatmap_data$heat_mat)
tier_mat <- t(heatmap_data$tier_mat)
col_hc <- heatmap_data$row_hc
cap <- heatmap_data$cap
col_meta <- heatmap_data$col_meta
ann_colors <- heatmap_data$ann_colors

col_fun <- colorRamp2(c(-cap, 0, cap), c("#2166AC", "white", "#B2182B"))

ha_left <- HeatmapAnnotation(
  df = col_meta[, c("BRCA","Segment","Time")],
  col = ann_colors,
  which = "row",
  annotation_name_side = "top",
  annotation_name_gp = grid::gpar(fontsize = 10),
  simple_anno_size = grid::unit(3.0, "mm")
)

ht <- Heatmap(
  heat_mat,
  name = "Signed logP (vs BS)",
  col = col_fun,
  left_annotation = ha_left,
  cluster_rows = FALSE,
  cluster_columns = col_hc,
  show_column_dend = TRUE,
  column_dend_height = grid::unit(18, "mm"),
  row_names_side = "left",
  cell_fun = function(j, i, x, y, w, h, fill) {
    tier <- tier_mat[i, j]
    ## Grey drawn first, black on top for the FDR tier.
    if (tier %in% c("nominal", "fdr")) {
      grid::grid.rect(x, y, w, h, gp = grid::gpar(col = "grey30", fill = NA, lwd = 1.5))
    }
    if (tier == "fdr") {
      grid::grid.rect(x, y, w, h, gp = grid::gpar(col = "black", fill = NA, lwd = 2))
    }
  }
)

lgd_sig_prot <- Legend(
  labels = c("FDR < 0.05", "nominal p < 0.05"),
  title = "Significance",
  type = "lines",
  legend_gp = grid::gpar(col = c("black", "grey30"), lwd = c(2, 1.5))
)

pdf(file.path(protein_fig_dir,"heatmap","prot_heatmap_v2_2.pdf"),width=14,height=6)
ComplexHeatmap::draw(ht, annotation_legend_list = list(lgd_sig_prot))
dev.off()
```
