---
title: "pipeline_talave_v1"
author: "Kenichi Shimada"
date: '2022-08-17'
---

# TALAVE CyCIF analysis

Uses the batch-corrected CyCIF objects to quantify cell-type frequencies, tumor frequencies,
marker-positive populations, double-positive states, T-cell stratifications, and tumor/stroma
classification.

**Required run order: `3` (this notebook) -> `4` -> `5` -> `6`.**
No dependency on the other CyCIF notebooks -- only needs the base `cs5`/`cs7` checkpoint.


## ---- Initial Setup ----

### Load Configuration
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Bootstrap
```{r bootstrap}
source(file.path(root_dir, "functions", "bootstrap.R"))
initialize_project()

if (!requireNamespace("CycifAnalyzeR", quietly = TRUE)) {
  remotes::install_github("<org>/CycifAnalyzeR")
}
library(CycifAnalyzeR)
source(file.path(root_dir,"functions","plotting.R"))
source(file.path(source_dir, "analysis_cycif.R"))
source(file.path(source_dir, "stats_utils.R"))
source(file.path(source_dir, "io.R"))
source(file.path(source_dir, "plotting_cycif.R"))
source(file.path(source_dir, "qc_cycif.R"))

brca.fills = c(MUT="#C9C0DD", WT="#D9DADA")
brca.cols = c(MUT="#62439c",WT="#58575f")
brca.mids = c(MUT="#62439c",WT="grey50")

```

### Load Batch-Corrected Objects

Only `cs5`/`cs7` are needed in this notebook (`frnn7.cmb`/`frnn7.exp`/`frnn7.rcn`/`is.sele.7`
aren't used here).
```{r}
cs5 <- readRDS(file.path(cycif_obj_dir, "cs5.rds"))
cs7 <- readRDS(file.path(cycif_obj_dir, "cs7.rds"))

pd5 <- pData(cs5)
pd7 <- pData(cs7)
cts5 <- cell_types(cs5)
cts7 <- cell_types(cs7)
nms7 <- names(cs7)
uniq.cts <- get_cycif_uniq_cts()
ct.cols7 <- get_cycif_ct_colors()
pt.cols <- load_pt_colors()
```


## ---- Define Shared Inputs ----
```{r}
cts.all <- c("Immune", "Tumor_panCK", "Fibro", "T", "Mac",
                "Tumor_neg", "CD8T", "CD4T", "T_other", "Mac_CD68",
                "Mac_CD163", "Mac_CD68_CD163", "Tumor_all")
cts.in.5 <- c("Tumor_panCK", "Tumor_neg", "CD8T","Mac_CD68", "Mac_CD163", "Mac_CD68_CD163",
              "Immune_other", "all_other")
cts.in.7 <- c("Tumor_panCK", "Tumor_neg", "Fibro", "CD8T", "CD4T", "T_other",
              "Mac_CD68", "Mac_CD163", "Mac_CD68_CD163", "Immune_other", "all_other")
cts.in.7only <- c("Fibro", "CD4T", "T_other")

cts.non.leaves <- c("Tumor_all","T","Mac","Immune")

center.cts <- uniq.cts[1:9]

nms5 <- names(cs5)
nms7only <- nms7[!nms7 %in% nms5]
```

```{r}
abs.5 <- c("AF488", "AF555", "AF647", "cCaspase3", "CD3", "PD1",
    "pERK", "panCK", "PDL1", "CD8a", "CD45", "GrB", "CD68", "CD163",
    "pAKT")
abs.7 <- c("AF488", "AF555", "AF647", "cCaspase3", "CD3", "PD1",
    "pERK", "panCK", "PDL1", "CD8a", "CD45", "GrB", "CD68", "CD163",
    "pAKT", "CD4", "MCL1", "BCLXL", "pTBK1", "aSMA", "gH2AX")
abs.7only <- abs.7[!abs.7 %in% abs.5]

all.equal(as.character(pd5$id),names(cs5))
all.equal(as.character(pd7$id),names(cs7))
```

## ---- Cell-Type Frequency Analysis ----

### Compute Cell-Type Frequencies
```{r}
nfreq5 <- cellTypeFrequency(cs5,simple=TRUE)
```



```{r}
nfreq7 <- cellTypeFrequency(cs7,simple=TRUE)
```

### [EDF2D] Merged Heatmap and Keys
```{r}
nfreq57 <- nfreq5
identical(rownames(nfreq5),rownames(nfreq7))
nfreq57[,nms7] <- nfreq7[,nms7]

this.ids <- pd5 %>% pull(id) %>% as.character()

pdf(file.path(cycif_fig_dir, "cell_type", "frequency", "cell_freq_merged_v3.pdf"),width=12,height=7)
par(mar=c(7,4,4,10))
xx <- barplot(nfreq57[11:1,this.ids],beside=F,col=ct.cols7[11:1],main="cell type frequency",names.arg=as.character(pd5$id),las=2,cex.names=1)
legend(par()$usr[2],par()$usr[4],names(ct.cols7[1:11]),fill=ct.cols7[1:11],xpd=T)

xi <- cumsum(table(pd5$Patient.ID))
mid.xi <- (xx[xi] + xx[xi+1])/2
abline(v=mid.xi,lty=2)
dev.off()

col1 <- RColorBrewer::brewer.pal(11,"Set3")[rep(seq(11),length=nlevels(pd5$Patient.ID))]
names(col1) <- unique(pd5$Patient.ID)
col2 <- c("grey20","grey80")
names(col2) <- c("MUT","WT")
col3 <- RColorBrewer::brewer.pal(length(unique(pd5$TimePoint)),"Set2")
names(col3) <- c("BS","BX2","BX3")
col4 <- RColorBrewer::brewer.pal(length(unique(pd5$BOR)),"Set1")[c(1,3,2)]
names(col4) <- levels(pd5$BOR)
col5 <- RColorBrewer::brewer.pal(length(unique(pd5$Subtype)),"Dark2")[2:1]
names(col5) <- levels(pd5$Subtype)  
col6 <- colorRampPalette(blues9)(round(sqrt(max(pd5$PFS_days))))

rsc <- list(
  Patient.ID = col1[pd5$Patient.ID],
  BRCA = col2[pd5$BRCA],
  TimePoint = col3[pd5$TimePoint],
  BOR = col4[pd5$BOR],
  Subtype = col5[pd5$Subtype],
  PFS = col6[round(sqrt(pd5$PFS_days))])[rev(c("BRCA","Subtype","BOR","PFS","Patient.ID","TimePoint"))]

pdf(file.path(cycif_fig_dir, "cell_type", "frequency", "color_palette_all_ct5.pdf"),width=12,height=7)
par(mfrow=c(6,1))
par(oma=c(5,5,5,5))
for(nm in names(rsc)){
  par(mar=c(0,0,0,0))
  image(seq(nrow(pd5)),1,as.matrix(seq(nrow(pd5))),col=rsc[[nm]],axes=F,xlab="",ylab="",main="")
}
dev.off()

n_colors <- 256
col6 <- colorRampPalette(blues9)(n_colors)
sqrt_breaks <- seq(0,sqrt(max(pd5$PFS_days, na.rm = TRUE)), length.out = n_colors + 1)
pfs_vals <- sqrt_breaks^2
yvals <- c(0, 1)
z <- matrix(1:n_colors, nrow = 1)
pdf(file.path(cycif_fig_dir, "cell_type", "frequency", "color_palette_pfs_ct5.pdf"),width=12,height=7)
par(mar = c(2, 4, 4, 2))
image(x = pfs_vals, y = yvals, z = t(z), col = col6, axes = FALSE, xlab = "", ylab = "")
axis(3, at = pretty(pfs_vals), labels = pretty(pfs_vals))
box()
dev.off()
```

```{r}
if(0){
  # Optional QC: export slide-level cell-type maps for manual review of sample composition.
  export_cycif_slide_celltype_plots(
    cs5 = cs5,
    cs7 = cs7,
    pd5 = pData(cs5),
    ct_cols = ct.cols7,
    out_dir = file.path(cycif_fig_dir, "misc", "qc", "slides")
  )
}
```

## ---- Tumor Frequency Analysis ----

### Prepare Tumor Frequency Data
```{r}
n7 <- apply(table(cell_types(cs7))[,uniq.cts],1,function(x)x/sum(x))

xs <- barplot(n7[11:1,],beside=F,col=rev(ct.cols7),plot=FALSE)
pid <- table(pd7 %>% filter(id %in% names(cs7)) %>% pull(Patient.ID))
pid <- pid[pid!=0]

x1 <- xs[cumsum(pid)]
x2 <- xs[cumsum(pid)+1]
xs7 <- (x1+x2)/2

uniq.ids7 <- as.character(pd7$id)


n5 <- apply(table(cell_types(cs5))[,uniq.cts],1,function(x)x/sum(x))

xs <- barplot(n5[11:1,],beside=F,col=rev(ct.cols7),plot=FALSE)
pid <- table(pd5 %>% filter(id %in% names(cs5)) %>% pull(Patient.ID))
pid <- pid[pid!=0]

x1 <- xs[cumsum(pid)]
x2 <- xs[cumsum(pid)+1]
xs5 <- (x1+x2)/2

uniq.ids5 <- as.character(pd5$id)


n57 <- n5
identical(rownames(n5),rownames(n7))
n57[,nms7] <- n7[,nms7]

```
### [Fig 2B] Tumor cell freq with `cs5` - with `plot_ct` function

```{r}
ctfreq57 <- array(NA,dim=c(nrow(n57),ncol(n57)))
ctfreq57[,] <- n57
rownames(ctfreq57) <- rownames(n57)
colnames(ctfreq57) <- colnames(n57)

df.57 <- prepare_tumor_frequency_df(ctfreq57, pd5)
df.5 <- prepare_tumor_frequency_df(apply(table(cell_types(cs5))[,uniq.cts],1,function(x)x/sum(x)), pd5)
df.7 <- prepare_tumor_frequency_df(apply(table(cell_types(cs7))[,uniq.cts],1,function(x)x/sum(x)), pd7)

p1 <- plot_ct(df.5 ,y_var="Tumor",y_label="Tumor frequency",title="Tumor frequency",pt.cols=pt.cols)
p2 <- plot_ct(df.7 ,y_var="Tumor",y_label="Tumor frequency",title="Tumor frequency",pt.cols=pt.cols)
p3 <- plot_ct(df.57 ,y_var="Tumor",y_label="Tumor frequency",title="Tumor frequency",pt.cols=pt.cols)

if(!dir.exists(file.path(cycif_fig_dir, "cell_type", "frequency"))){
  dir.create(file.path(cycif_fig_dir, "cell_type", "frequency"),recursive=TRUE)
}

pdf(file.path(cycif_fig_dir, "cell_type", "frequency", "tumor_freq_merged_v2.pdf"),width=5,height=7)
print(p3$plot)
dev.off()

```
## ---- Cell-Type Frequency Plots ----
### Generate Cell-Type Frequency Plots
Use `cs5` for cell types in `cts.in.5` and `cs7` for cell types in `cts.in.7only`.

```{r}
n.ctf5 <- build_augmented_celltype_frequency(cs5)
nm.ratios <- rownames(n.ctf5)
df.ctf5 <- prepare_sample_frequency_df(n.ctf5, pd5)
```

```{r}
n.ctf7 <- build_augmented_celltype_frequency(cs7)
nm.ratios <- rownames(n.ctf7)
df.ctf7 <- prepare_sample_frequency_df(n.ctf7, pd7)
```

```{r}
cmp.conds7 <- rownames(n.ctf7)
is7 <- sub("\\..+","",cmp.conds7) %in% cts.in.7only
names(is7) <- cmp.conds7

conds.bt3 <- list()

for(con in cmp.conds7){
  if(is7[con]){
    df.ctf <- df.ctf7
  }else{
    df.ctf <- df.ctf5
  }
  this.con <- strsplit(con,"\\.")[[1]]
  p <- plot_ct(df.ctf,y_var=con,
               y_label=paste0("Cell type frequency (",this.con[1],"/",this.con[2],")"),
               title=sub("\\.","/",con),pt.cols=pt.cols)
  conds.bt3[[con]] <- p$stat

  if(1){
    pdf(file.path(cycif_fig_dir, "cell_type", "frequency", paste0("ct_freq_",con,".pdf")),width=5,height=7)
    print(p$plot)
    dev.off()
  }
}
```

### Model Diagnostics: Normality Check (pooled across all cell-type/marker frequency variables, standardized)

Residuals standardized per variable before pooling, to avoid a scale-mixing artifact that
would otherwise look leptokurtic regardless of true normality.
```{r}
diag_ct_vars <- cmp.conds7

diag_ct_residuals_list <- lapply(diag_ct_vars, function(con) {
  df.ctf <- if (is7[con]) df.ctf7 else df.ctf5
  get_ct_freq_lmer_residuals(df.ctf, con)
})

diag_ct_residuals_std <- lapply(diag_ct_residuals_list, function(r) {
  if (is.null(r) || length(r) < 2) return(NULL)
  s <- sd(r)
  if (!is.finite(s) || s == 0) return(NULL)
  r / s
}) %>% unlist(use.names = FALSE)

cat("n cell-type/marker frequency variables:", length(diag_ct_vars),
    " | n pooled standardized residuals:", length(diag_ct_residuals_std), "\n")

qqnorm(diag_ct_residuals_std, main = "Q-Q plot of pooled residuals: CyCIF frequency LMER family (all variables, standardized per variable)")
qqline(diag_ct_residuals_std)

set.seed(1)
n_draws <- 20
draw_size <- 1000

shapiro_summary_ct <- run_shapiro_draws(diag_ct_residuals_std, draw_size, n_draws)

cat(sprintf(
  "Shapiro-Wilk across %d draws of n=%d:\n  W: median=%.4f, range=[%.4f, %.4f]\n  p: median=%.2e, range=[%.2e, %.2e]\n",
  n_draws, draw_size,
  median(shapiro_summary_ct$W), min(shapiro_summary_ct$W), max(shapiro_summary_ct$W),
  median(shapiro_summary_ct$p), min(shapiro_summary_ct$p), max(shapiro_summary_ct$p)
))

shapiro_summary_ct
```

### [Fig 5A, EDF7A] Heatmap of Frequency Changes Between Time Points

```{r}

heatmap_inputs <- prepare_signed_logp_heatmap_inputs(conds.bt3)
mat_val_filt <- heatmap_inputs$mat_val_filt
mat_val_filt1 <- heatmap_inputs$mat_val_filt1
tier_ps <- heatmap_inputs$tier_ps
tier_ps1 <- heatmap_inputs$tier_ps1
col_fun <- heatmap_inputs$col_fun
ann_colors <- heatmap_inputs$ann_colors
col_annot <- heatmap_inputs$col_annot
col_annot1 <- heatmap_inputs$col_annot1
```

```{r}
ht_ps <- build_signed_logp_heatmap(
  mat = mat_val_filt,
  tier_mat = tier_ps,
  annotation_df = col_annot,
  ann_colors = ann_colors,
  col_fun = col_fun
)
```

```{r}
lgd_sig_cycif <- Legend(
  labels = c("FDR < 0.05", "nominal p < 0.05"),
  title = "Significance",
  type = "lines",
  legend_gp = gpar(col = c("black", "grey50"), lwd = c(2, 1.5))
)

pdf(file.path(cycif_fig_dir, "cell_type", "frequency_heatmap", "sig_signed_logps_heatmap_merged_v3.pdf"),
    width = 5, height = 7)
ComplexHeatmap::draw(ht_ps, heatmap_legend_side = "right", annotation_legend_list = list(lgd_sig_cycif))
dev.off()

ht_ps1 <- build_signed_logp_heatmap(
  mat = mat_val_filt1,
  tier_mat = tier_ps1,
  annotation_df = col_annot1,
  ann_colors = ann_colors,
  col_fun = col_fun
)

pdf(file.path(cycif_fig_dir, "cell_type", "frequency_heatmap", "sig_signed_logps_heatmap_merged_full_v3.pdf"),
    width = 7, height = 7)
ComplexHeatmap::draw(ht_ps1, heatmap_legend_side = "right", annotation_legend_list = list(lgd_sig_cycif))
dev.off()
```

## ---- Marker-Positive Cell Frequency ----
### Prepare Marker-Positive Cell Frequencies
```{r}
exp.th5 <- exprs(cs5,type="logTh")
exp.log5 <- exprs(cs5,type="log")
exp.th7 <- exprs(cs7,type="logTh")
exp.log7 <- exprs(cs7,type="log")

cst <- cs7@cell_types$default@cell_state_def
cst.abs <- colnames(cst)
uniq.cts1 <- unique(sub("Tumor_.+","Tumor",uniq.cts))
uniq.cts1 <- uniq.cts1[!uniq.cts1 %in% c("Immune_other","all_other")]

mats.fr <- list()

for(ab in cst.abs){
  this.cts <- cst %>% filter(!is.na(!!sym(ab)) & !!sym(ab) == "CAN") %>% rownames()

  if(ab %in% abs.5){
    cts <- cts5
    exp.th <- exp.th5
    exp.log <- exp.log5
    pd <- pd5
  }else{
    cts <- cts7
    exp.th <- exp.th7
    exp.log <- exp.log7
    pd <- pd7
  }

  marker_tables <- prepare_marker_positive_frequency_tables(
    cts = cts,
    exp_th = exp.th,
    exp_log = exp.log,
    pd = pd,
    ab = ab,
    this_cts = this.cts
  )
  mats.fr[[ab]] <- marker_tables$mat_fr
}
```

### [Fig 3G, EDF6D] Marker-Positive Frequencies by Cell Type

gH2AX/pTBK1 marker-positive frequencies -> Fig.3G; PD1 marker-positive frequencies
(`pos_freq_PD1_<CD4T/CD8T/T_other>_v4.pdf`) -> EDF6D.
```{r}
d1 <- file.path(cycif_fig_dir, "marker_frequency", "individual", "individual")
if(!dir.exists(d1)){
  dir.create(d1, recursive = TRUE)
}

for(ab in cst.abs){
  if(ab %in% abs.5){
    pd <- pd5
  }else{
    pd <- pd7
  }
  mat.fr <- as.data.frame(mats.fr[[ab]]) %>%
    rownames_to_column("sample") %>%
    mutate(sample = factor(sample, levels = levels(pd$id))) %>%
    pivot_longer(-sample, names_to = "cell_types", values_to = "pos_frac") %>%
    left_join(pd %>% select(id, bt3), by = c("sample" = "id")) %>%
    mutate(
      cell_types = factor(cell_types,levels=uniq.cts1)
    )

  if(length(unique(mat.fr$cell_types)) > 1){
    df <- mats.fr[[ab]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column("id") %>%
      left_join(pd %>% select(id,BRCA,TimePoint,PFS_days,Patient.ID,BOR,bt3), by = "id")

    df <- df %>%
      group_by(Patient.ID) %>%
      mutate(
        max_bn = max(as.numeric(TimePoint)),
        text = ifelse(as.numeric(TimePoint) == max_bn,as.character(Patient.ID),"")
      ) %>%
      ungroup()

    this.cts <- colnames(mats.fr[[ab]])

    for(ct in this.cts){
      df1 <- df %>% filter(!is.na(!!sym(ct)))
      p <- plot_ct(df1, y_var = ct, y_label = paste0("Positive fraction in ",ct),
        title = paste0(ab, " positive fraction in ",ct),
        pt.cols = pt.cols)
      pdf(file.path(d1, paste0("pos_freq_", ab, "_",ct,"_v4.pdf")), width = 5, height = 7)
      print(p$plot)
      dev.off()
    }
  }else{
    ct <- as.character(mat.fr$cell_types[1])

    mat.fr <- mat.fr %>%
      select(1:3) %>%
      left_join(pd %>% select(id,BRCA,TimePoint,PFS_days,Patient.ID,BOR,bt3), by = c("sample" = "id"))

    p <- plot_ct(mat.fr, y_var = "pos_frac", y_label = paste0("Positive fraction in ",ct),
        title = paste0(ab, " positive fraction in ",ct),
        pt.cols = pt.cols)

    pdf(file.path(d1, paste0("pos_freq_", ab, "_",ct,"_v4.pdf")), width = 5, height = 7)
    print(p$plot)
    dev.off()
  }
}
```
### [Fig 3G (right)] Tumor gH2AX/pTBK1 Double-Positive Fraction

`pos_freq_gH2AX_pTBK1_Tumor_v4.pdf` -> Fig.3G right (paired with the single-marker
`pos_freq_gH2AX_Tumor_v4.pdf`/`pos_freq_pTBK1_Tumor_v4.pdf` -> Fig.3G left/middle, generated
in the main marker-positive-frequency loop above). `emmeans_gH2AX_pTBK1_v4.pdf` not
confirmed used.
```{r}
ab1 <- "gH2AX"
ab2 <- "pTBK1"

cts <- cts7
exp.th <- exp.th7
pd <- pd7

uniq.cts1 <- unique(sub("Tumor_.+","Tumor",uniq.cts))
uniq.cts1 <- uniq.cts1[!uniq.cts1 %in% c("Immune_other","all_other")]

mat.fr <- prepare_double_positive_frequency_df(
  cts = cts,
  exp_th = exp.th,
  pd = pd,
  ab1 = ab1,
  ab2 = ab2,
  selected_cts = "Tumor"
) %>%
  mutate(cell_types = factor(cell_types, levels = uniq.cts1))

ct <- "Tumor"
p <- plot_ct(mat.fr, y_var = "pos_frac", y_label = paste0("Positive fraction in ",ct),
    title = paste0(ab1,"/",ab2, " double-positive fraction in ",ct),
  pt.cols = pt.cols)

pdf(file.path(d1, paste0("pos_freq_", ab1,"_",ab2, "_",ct,"_v4.pdf")), width = 5, height = 7)
print(p$plot)
dev.off()

```

## ---- T Cell Stratification ----
### Prepare `CD8T`
```{r}
nms5 <- names(cs5)
non_cd8t.ts <- cd8.ts <- list()

for(nm in nms5){
  tmp <- cs5[[nm]]
  cts <- cell_types(tmp)
  is.cd8t <- cts$cell_types == "CD8T"
  is.non_cd8t <- cts$cell_types == "T_other"
  

  ex <- exprs(tmp,type="logTh")

  ## CD8T  
  if(sum(is.cd8t,na.rm=T)){
    pd1 <- ex[is.cd8t,"PD1"]
    grb <- ex[is.cd8t,"GrB"]
    tab <- matrix(0,nrow=2,ncol=2)
    colnames(tab) <- rownames(tab) <- c("TRUE","FALSE")
    tab1 <- table(pd1=pd1>0.5,GrB=grb>0.5)#[c("TRUE","FALSE"),c("TRUE","FALSE")]
    tab[rownames(tab1),colnames(tab1)] <- tab1
    tab <- as.vector(tab)
  }else{
    tab <- c(0,0,0,0)  # no CD8T cells
  }    
  names(tab) <- c("PD1_GrB","GrB","PD1","dn")

  cd8.ts[[nm]] <- tab[c(2,1,3,4)]
  
  ## Non-CD8T
  if(sum(is.non_cd8t,na.rm=T)){
    pd1 <- ex[is.non_cd8t,"PD1"]
    grb <- ex[is.non_cd8t,"GrB"]
    tab <- matrix(0,nrow=2,ncol=2)
    colnames(tab) <- rownames(tab) <- c("TRUE","FALSE")
    tab1 <- table(pd1=pd1>0.5,GrB=grb>0.5)#[c("TRUE","FALSE"),c("TRUE","FALSE")]
    tab[rownames(tab1),colnames(tab1)] <- tab1
    tab <- as.vector(tab)
  }else{
    tab <- c(0,0,0,0)  # no non_cd8t cells
  }
  names(tab) <- c("PD1_GrB","GrB","PD1","dn")

  non_cd8t.ts[[nm]] <- tab[c(2,1,3,4)]
}

cd8.stat <- do.call(rbind,cd8.ts)
n.cd8 <- apply(cd8.stat,1,function(x)x/sum(x))
n.cd8[,"1011-2"] <- 0

non_cd8t.stat <- do.call(rbind,non_cd8t.ts)
n.non_cd8t <- apply(non_cd8t.stat,1,function(x)x/sum(x))
n.non_cd8t[,"1011-2"] <- 0

t_all.stat <- cd8.stat + non_cd8t.stat
n.t_all <- apply(t_all.stat,1,function(x)x/sum(x))
```
### Prepare `CD4T`/`T_other`
```{r}
nms7 <- names(cs7)
t_other.ts <- cd4.ts <- list()

for(nm in nms7){
  tmp <- cs7[[nm]]
  cts <- cell_types(tmp)
  is.cd4t <- cts$cell_types == "CD4T"
  is.t_other <- cts$cell_types == "T_other"

  ex <- exprs(tmp,type="logTh")

  ## CD4T  
  if(sum(is.cd4t,na.rm=T)){
    pd1 <- ex[is.cd4t,"PD1"]
    grb <- ex[is.cd4t,"GrB"]
    tab <- matrix(0,nrow=2,ncol=2)
    colnames(tab) <- rownames(tab) <- c("TRUE","FALSE")
    tab1 <- table(pd1=pd1>0.5,GrB=grb>0.5)#[c("TRUE","FALSE"),c("TRUE","FALSE")]
    tab[rownames(tab1),colnames(tab1)] <- tab1
    tab <- as.vector(tab)
  }else{
    tab <- c(0,0,0,0)  # no CD4T cells
  }    
  names(tab) <- c("PD1_GrB","GrB","PD1","dn")

  cd4.ts[[nm]] <- tab[c(2,1,3,4)]
  
  ## T_other
  if(sum(is.t_other,na.rm=T)){
    pd1 <- ex[is.t_other,"PD1"]
    grb <- ex[is.t_other,"GrB"]
    tab <- matrix(0,nrow=2,ncol=2)
    colnames(tab) <- rownames(tab) <- c("TRUE","FALSE")
    tab1 <- table(pd1=pd1>0.5,GrB=grb>0.5)#[c("TRUE","FALSE"),c("TRUE","FALSE")]
    tab[rownames(tab1),colnames(tab1)] <- tab1
    tab <- as.vector(tab)
  }else{
    tab <- c(0,0,0,0)  # no T_other cells
  }
  names(tab) <- c("PD1_GrB","GrB","PD1","dn")

  t_other.ts[[nm]] <- tab[c(2,1,3,4)]
}

cd4.stat <- do.call(rbind,cd4.ts)
n.cd4 <- apply(cd4.stat,1,function(x)x/sum(x))
t_other.stat <- do.call(rbind,t_other.ts)
n.t_other <- apply(t_other.stat,1,function(x)x/sum(x))
```

### [Fig 4H, EDF6C, EDF6G] T-Cell States vs PFS

`TC_state/T_other_PD1_pfs_w_pval.pdf` -> EDF6C. `TC_state/CD4T_PD1_pfs_w_pval.pdf`,
`TC_state/CD4T_PD1GrB_pfs_w_pval.pdf`, `TC_state/CD8T_PD1_pfs_w_pval.pdf`,
`TC_state/CD8T_PD1GrB_pfs_w_pval.pdf`, `TC_state/T_other_PD1_pfs_w_pval.pdf`,
`TC_state/T_other_PD1GrB_pfs_w_pval.pdf` -> EDF6G (note `T_other_PD1_pfs_w_pval.pdf` is
shared between EDF6C and EDF6G, per author).

```{r}
n1 <- (t(n.cd8[c(2,1,3,4),]) %>% 
     as.data.frame() %>% 
     mutate(PD1_all = PD1 + PD1_GrB) %>%
     mutate(GrB_all = GrB + PD1_GrB) %>%       
     tibble::rownames_to_column("id"))

n2 <- (t(n.t_other[c(2,1,3,4),]) %>% 
     as.data.frame() %>% 
     mutate(PD1_all = PD1 + PD1_GrB) %>%
     mutate(GrB_all = GrB + PD1_GrB) %>%              
     tibble::rownames_to_column("id"))

n4 <- (t(n.cd4[c(2,1,3,4),]) %>%
     as.data.frame() %>% 
     mutate(PD1_all = PD1 + PD1_GrB) %>%
     mutate(GrB_all = GrB + PD1_GrB) %>%       
     tibble::rownames_to_column("id"))

pd5 <- pd5 %>%
  mutate(BRCA = as.character(BRCA)) %>%
  mutate(Patient.ID = factor(Patient.ID))

pd7 <- pd7 %>%
  mutate(BRCA = as.character(BRCA)) %>%
  mutate(Patient.ID = factor(Patient.ID))

pd.cd8 <- pd5 %>% left_join(n1,by="id")
pd.t_other <- pd5 %>% left_join(n2,by="id")
pd.cd4 <- pd7 %>% left_join(n4,by="id")

eps <- 0.5
pd.ratio <-n1[c(1,6)] %>%
  left_join(n4[c(1,6)],by="id",suffix=c("_cd8","_cd4")) %>%
  filter(!is.na(PD1_all_cd8) & !is.na(PD1_all_cd4)) %>%
  mutate(ratio = (PD1_all_cd8 + eps)/(PD1_all_cd4 + eps)) %>%
  mutate(log2ratio = log2(ratio)) %>%
  left_join(pd7 %>% select(id,BRCA,TimePoint,PFS_days,Patient.ID,BOR,Subtype),by="id")
hist(pd.ratio$log2ratio)
```

#### CD8T
```{r}
markers <- c("PD1_all", "PD1_GrB", "GrB_all","PD1","GrB","dn")
titles <- c(
  PD1_all = "PD-1+ CD8T cells",
  PD1_GrB = "PD-1+/GrB+ CD8T cells",
  GrB_all = "GrB+ CD8T cells",
  PD1 = "PD-1+/GrB- CD8T cells",
  GrB = "PD-1-/GrB+ CD8T cells",
  dn = "PD-1-/GrB- CD8T cells"
)
xlabels <- c(
  PD1_all = "% PD-1+ cells (out of CD8T cells)",
  PD1_GrB = "% PD-1+/GrB+ cells (out of CD8T cells)",
  GrB_all = "% GrB+ cells (out of CD8T cells)",
  PD1 = "% PD-1+/GrB- cells (out of CD8T cells)",
  GrB = "% PD-1-/GrB+ cells (out of CD8T cells)",
  dn = "% PD-1-/GrB- cells (out of CD8T cells)"
)
filenames <- c(
  PD1_all = "CD8T_PD1_all_pfs_w_pval.pdf",
  PD1_GrB = "CD8T_PD1GrB_pfs_w_pval.pdf",
  GrB_all = "CD8T_GrB_all_pfs_w_pval.pdf",
  PD1 = "CD8T_PD1_pfs_w_pval.pdf",
  GrB = "CD8T_GrB_pfs_w_pval.pdf",
  dn = "CD8T_dn_pfs_w_pval.pdf"
)

for (marker in markers) {
  p.vals <- matrix(NA, nrow = 2, ncol = 3, dimnames = list(c("MUT", "WT"), c("BS", "BX2", "BX3")))
  r.adj <- p.vals.marker <- p.vals
  
  for (brca in c("MUT", "WT")) {
    for (tp in c("BS", "BX2", "BX3")) {
      x <- pd.cd8 %>% filter(BRCA == brca, TimePoint == tp) %>% pull(!!sym(marker))
      y <- pd.cd8 %>% filter(BRCA == brca, TimePoint == tp) %>% pull(PFS_days)
      if (length(x) > 1 && length(unique(x)) > 1) {
        lm1 <- lm(log(y) ~ x)
        p.vals.marker[brca, tp] <- summary(lm1)$coefficients[2, 4]
        r.adj[brca, tp] <- summary(lm1)$r.squared
      }
    }
  }

  df.cd8 <- melt(p.vals.marker, varnames = c("BRCA", "TimePoint"), value.name = "p-val") %>%
    left_join(
      melt(r.adj, varnames = c("BRCA", "TimePoint"), value.name = "r-sq"),
      by = c("BRCA", "TimePoint")
    ) %>%
    mutate(txt = paste0("p=", round(`p-val`, 3), "\nR2=", round(`r-sq`, 3))) %>%
    mutate(
      y = case_when(
        BRCA == "MUT" ~ 400,
        BRCA == "WT" ~ 150,
        TRUE ~ 0.025
      ),
      x = case_when(
        marker == "PD1_all" ~ 0.125,
        marker == "PD1_GrB" ~ 0.15,
        marker == "GrB_all" ~ 0.25,        
        marker == "PD1" ~ c(0.2,0.4,0.2,0.125,0.3,0.8),
        marker == "GrB" ~ 0.06,
        marker == "dn" ~ c(0.7,0.4,0.7,0.125,0.7,0.15)
      )
    )

  p <- ggplot(pd.cd8, aes_string(x = marker, y = "PFS_days", color = "BRCA", label = "Patient.ID")) +
    scale_y_log10() +
    geom_point(aes(shape = Subtype)) +
    scale_color_manual(values = brca.mids) + 
    facet_wrap(~TimePoint) +
    labs(
      title = titles[[marker]],
      x = xlabels[[marker]],
      y = "PFS (days)"
    ) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_text_repel() +
    geom_text(data = df.cd8, aes(x = x, y = y, label = txt), size = 5, color = 1) +
    theme_bw()

  pdf(file.path(cycif_fig_dir, "marker_frequency", "pfs_by_state", filenames[[marker]]), width = 12, height = 4)
  print(p)
  dev.off()
}
```

```{r}
markers <- c("PD1_all", "PD1_GrB", "GrB_all")
titles <- c(
  PD1_all = "PD-1+ in CD8T cells",
  PD1_GrB = "PD-1+/GrB+ in CD8T cells",
  GrB_all = "GrB+ in CD8T cells"
)
xlabels <- c(
  PD1_all = "% PD-1+ cells (out of CD8T cells)",
  PD1_GrB = "% PD-1+/GrB+ cells (out of CD8T cells)",
  GrB_all = "% GrB+ cells (out of CD8T cells)"
)
filenames <- c(
  PD1_all = "CD8T_PD1_all_pfs_w_pval.pdf",
  PD1_GrB = "CD8T_PD1GrB_pfs_w_pval.pdf",
  GrB_all = "CD8T_GrB_all_pfs_w_pval.pdf"
)

for (marker in markers) {
  p.vals <- matrix(NA, nrow = 2, ncol = 3, dimnames = list(c("MUT", "WT"), c("BS", "BX2", "BX3")))
  r.adj.cd8 <- p.vals.cd8 <- p.vals
  
  for (brca in c("MUT", "WT")) {
    for (tp in c("BS", "BX2", "BX3")) {
      x <- pd.cd8 %>% filter(BRCA == brca, TimePoint == tp) %>% pull(!!sym(marker))
      y <- pd.cd8 %>% filter(BRCA == brca, TimePoint == tp) %>% pull(PFS_days)
      if (length(x) > 1 && length(unique(x)) > 1) {
        lm1 <- lm(log(y) ~ x)
        p.vals.cd8[brca, tp] <- summary(lm1)$coefficients[2, 4]
        r.adj.cd8[brca, tp] <- summary(lm1)$r.squared
      }
    }
  }

  df.cd8 <- melt(p.vals.cd8, varnames = c("BRCA", "TimePoint"), value.name = "p-val") %>%
    left_join(
      melt(r.adj.cd8, varnames = c("BRCA", "TimePoint"), value.name = "r-sq"),
      by = c("BRCA", "TimePoint")
    ) %>%
    mutate(txt = paste0("p=", round(`p-val`, 3), "\nR2=", round(`r-sq`, 3))) %>%
    mutate(x = c(0.2,0.2,0.2,0.5,0.2,0.2),
           y = c(400,75,500,50,350,75))

  p <- ggplot(pd.cd8, aes_string(x = marker, y = "PFS_days", color = "BRCA", label = "Patient.ID")) +
    scale_y_log10() +
    geom_point(aes(shape = Subtype)) +
    scale_color_manual(values = brca.mids) + 
    facet_wrap(~TimePoint) +
    labs(
      title = titles[[marker]],
      x = xlabels[[marker]],
      y = "PFS (days)"
    ) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_text_repel() +
    geom_text(data = df.cd8, aes(x = x, y = y, label = txt), size = 5, color = 1) +
    theme_bw()

  pdf(file.path(cycif_fig_dir, "marker_frequency", "pfs_by_state", filenames[[marker]]), width = 12, height = 4)
  print(p)
  dev.off()
}
```


#### CD4T
```{r}
markers <- c("PD1_all", "PD1_GrB", "GrB_all","PD1","GrB","dn")
titles <- c(
  PD1_all = "PD-1+ CD4T cells",
  PD1_GrB = "PD-1+/GrB+ CD4T cells",
  GrB_all = "GrB+ CD4T cells",
  PD1 = "PD-1+/GrB- CD4T cells",
  GrB = "PD-1-/GrB+ CD4T cells",
  dn = "PD-1-/GrB- CD4T cells"
)
xlabels <- c(
  PD1_all = "% PD-1+ cells (out of CD4T cells)",
  GrB_all = "% GrB+ cells (out of CD4T cells)",
  PD1_GrB = "% PD-1+/GrB+ cells (out of CD4T cells)",
  PD1 = "% PD-1+/GrB- cells (out of CD4T cells)",
  GrB = "% PD-1-/GrB+ cells (out of CD4T cells)",
  dn = "% PD-1-/GrB- cells (out of CD4T cells)"
)
filenames <- c(
  PD1_all = "CD4T_PD1_all_pfs_w_pval.pdf",
  PD1_GrB = "CD4T_PD1GrB_pfs_w_pval.pdf",
  GrB_all = "CD4T_GrB_all_pfs_w_pval.pdf",
  PD1 = "CD4T_PD1_pfs_w_pval.pdf",
  GrB = "CD4T_GrB_pfs_w_pval.pdf",
  dn = "CD4T_dn_pfs_w_pval.pdf"
)

for (marker in markers) {
  p.vals <- matrix(NA, nrow = 2, ncol = 3, dimnames = list(c("MUT", "WT"), c("BS", "BX2", "BX3")))
  r.adj <- p.vals.marker <- p.vals
  
  for (brca in c("MUT", "WT")) {
    for (tp in c("BS", "BX2", "BX3")) {
      x <- pd.cd4 %>% filter(BRCA == brca, TimePoint == tp) %>% pull(!!sym(marker))
      y <- pd.cd4 %>% filter(BRCA == brca, TimePoint == tp) %>% pull(PFS_days)
      if (length(x) > 1 && length(unique(x)) > 1) {
        lm1 <- lm(log(y) ~ x)
        p.vals.marker[brca, tp] <- summary(lm1)$coefficients[2, 4]
        r.adj[brca, tp] <- summary(lm1)$r.squared
      }
    }
  }

  df.cd4 <- melt(p.vals.marker, varnames = c("BRCA", "TimePoint"), value.name = "p-val") %>%
    left_join(
      melt(r.adj, varnames = c("BRCA", "TimePoint"), value.name = "r-sq"),
      by = c("BRCA", "TimePoint")
    ) %>%
    mutate(txt = paste0("p=", round(`p-val`, 3), "\nR2=", round(`r-sq`, 3))) %>%
    mutate(
      y = case_when(
        BRCA == "MUT" ~ 400,
        BRCA == "WT" ~ 150,
        TRUE ~ 0.025
      ),
      x = case_when(
        marker == "PD1_all" ~ c(0.4,0.4,0.4,0.4,0.3,0.7),
        marker == "PD1_GrB" ~ 0.04,
        marker == "GrB_all" ~ 0.05,        
        marker == "PD1" ~ c(0.4,0.4,0.4,0.4,0.4,0.8),
        marker == "GrB" ~ 0.04,
        marker == "dn" ~ c(0.5,0.5,0.5,0.4,0.2,0.5)
      )
    )

  p <- ggplot(pd.cd4, aes_string(x = marker, y = "PFS_days", color = "BRCA", label = "Patient.ID")) +
    scale_y_log10() +
    geom_point(aes(shape = Subtype)) +
    scale_color_manual(values = brca.mids) + 
    facet_wrap(~TimePoint) +
    labs(
      title = titles[[marker]],
      x = xlabels[[marker]],
      y = "PFS (days)"
    ) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_text_repel() +
    geom_text(data = df.cd4, aes(x = x, y = y, label = txt), size = 5, color = 1) +
    theme_bw()

  pdf(file.path(cycif_fig_dir, "marker_frequency", "pfs_by_state", filenames[[marker]]), width = 12, height = 4)
  print(p)
  dev.off()
}
```

```{r}
markers <- c("PD1_all", "PD1_GrB", "GrB_all")
titles <- c(
  PD1_all = "PD-1+ in CD4T cells",
  PD1_GrB = "PD-1+/GrB+ in CD4T cells",
  GrB_all = "GrB+ in CD4T cells"
)
xlabels <- c(
  PD1_all = "% PD-1+ cells (out of CD4T cells)",
  PD1_GrB = "% PD-1+/GrB+ cells (out of CD4T cells)",
  GrB_all = "% GrB+ cells (out of CD4T cells)"
)
filenames <- c(
  PD1_all = "CD4T_PD1_all_pfs_w_pval.pdf",
  PD1_GrB = "CD4T_PD1GrB_pfs_w_pval.pdf",
  GrB_all = "CD4T_GrB_all_pfs_w_pval.pdf"
)

for (marker in markers) {
  p.vals <- matrix(NA, nrow = 2, ncol = 3, dimnames = list(c("MUT", "WT"), c("BS", "BX2", "BX3")))
  r.adj.cd4 <- p.vals.cd4 <- p.vals
  
  for (brca in c("MUT", "WT")) {
    for (tp in c("BS", "BX2", "BX3")) {
      x <- pd.cd4 %>% filter(BRCA == brca, TimePoint == tp) %>% pull(!!sym(marker))
      y <- pd.cd4 %>% filter(BRCA == brca, TimePoint == tp) %>% pull(PFS_days)
      if (length(x) > 1 && length(unique(x)) > 1) {
        lm1 <- lm(log(y) ~ x)
        p.vals.cd4[brca, tp] <- summary(lm1)$coefficients[2, 4]
        r.adj.cd4[brca, tp] <- summary(lm1)$r.squared
      }
    }
  }

  df.cd4 <- melt(p.vals.cd4, varnames = c("BRCA", "TimePoint"), value.name = "p-val") %>%
    left_join(
      melt(r.adj.cd4, varnames = c("BRCA", "TimePoint"), value.name = "r-sq"),
      by = c("BRCA", "TimePoint")
    ) %>%
    mutate(txt = paste0("p=", round(`p-val`, 3), "\nR2=", round(`r-sq`, 3))) %>%
    mutate(
      y = case_when(
        BRCA == "MUT" ~ 400,
        BRCA == "WT" ~ 150,
        TRUE ~ 0.025
      ),
      x = case_when(
        marker == "PD1_all" ~ c(0.4,0.4,0.4,0.8,0.4,0.4),
        marker == "PD1_GrB" ~ 0.15,
        marker == "GrB_all" ~ 0.02)
      )

  p <- ggplot(pd.cd4, aes_string(x = marker, y = "PFS_days", color = "BRCA", label = "Patient.ID")) +
    scale_y_log10() +
    geom_point(aes(shape = Subtype)) +
    scale_color_manual(values = brca.mids) + 
    facet_wrap(~TimePoint) +
    labs(
      title = titles[[marker]],
      x = xlabels[[marker]],
      y = "PFS (days)"
    ) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_text_repel() +
    geom_text(data = df.cd4, aes(x = x, y = y, label = txt), size = 5, color = 1) +
    theme_bw()

  pdf(file.path(cycif_fig_dir, "marker_frequency", "pfs_by_state", filenames[[marker]]), width = 12, height = 4)
  print(p)
  dev.off()
}
```
#### T_other
```{r}
markers <- c("PD1_all", "PD1_GrB", "GrB_all","PD1","GrB","dn")
titles <- c(
  PD1_all = "PD-1+ T_other cells",
  PD1_GrB = "PD-1+/GrB+ T_other cells",
  GrB_all = "GrB+ T_other cells",
  PD1 = "PD-1+/GrB- T_other cells",
  GrB = "PD-1-/GrB+ T_other cells",
  dn = "PD-1-/GrB- T_other cells"
)
xlabels <- c(
  PD1_all = "% PD-1+ cells (out of T_other cells)",
  PD1_GrB = "% PD-1+/GrB+ cells (out of T_other cells)",
  GrB_all = "% GrB+ cells (out of T_other cells)",
  PD1 = "% PD-1+/GrB- cells (out of T_other cells)",
  GrB = "% PD-1-/GrB+ cells (out of T_other cells)",
  dn = "% PD-1-/GrB- cells (out of T_other cells)"
)
filenames <- c(
  PD1_all = "T_other_PD1_all_pfs_w_pval.pdf",
  PD1_GrB = "T_other_PD1GrB_pfs_w_pval.pdf",
  GrB_all = "T_other_GrB_all_pfs_w_pval.pdf",
  PD1 = "T_other_PD1_pfs_w_pval.pdf",
  GrB = "T_other_GrB_pfs_w_pval.pdf",
  dn = "T_other_dn_pfs_w_pval.pdf"
)

for (marker in markers) {
  p.vals <- matrix(NA, nrow = 2, ncol = 3, dimnames = list(c("MUT", "WT"), c("BS", "BX2", "BX3")))
  r.adj <- p.vals.marker <- p.vals
  
  for (brca in c("MUT", "WT")) {
    for (tp in c("BS", "BX2", "BX3")) {
      x <- pd.t_other %>% filter(BRCA == brca, TimePoint == tp) %>% pull(!!sym(marker))
      y <- pd.t_other %>% filter(BRCA == brca, TimePoint == tp) %>% pull(PFS_days)
      if (length(x) > 1 && length(unique(x)) > 1) {
        lm1 <- lm(log(y) ~ x)
        p.vals.marker[brca, tp] <- summary(lm1)$coefficients[2, 4]
        r.adj[brca, tp] <- summary(lm1)$r.squared
      }
    }
  }

  df.t_other <- melt(p.vals.marker, varnames = c("BRCA", "TimePoint"), value.name = "p-val") %>%
    left_join(
      melt(r.adj, varnames = c("BRCA", "TimePoint"), value.name = "r-sq"),
      by = c("BRCA", "TimePoint")
    ) %>%
    mutate(txt = paste0("p=", round(`p-val`, 3), "\nR2=", round(`r-sq`, 3))) %>%
    mutate(
      y = case_when(
        BRCA == "MUT" ~ 400,
        BRCA == "WT" ~ 150,
        TRUE ~ 0.025
      ),
      x = case_when(
        marker == "PD1_all" ~ 0.9,
        marker == "PD1_GrB" ~ 0.025,
        marker == "GrB_all" ~ 0.1,        
        marker == "PD1" ~ c(0.2,0.8,0.2,0.8,0.8,0.8),
        marker == "GrB" ~ 0.075,
        marker == "dn" ~ c(0.8,0.25,0.8,0.25,0.1,0.1)
      )
    )

  p <- ggplot(pd.t_other, aes_string(x = marker, y = "PFS_days", color = "BRCA", label = "Patient.ID")) +
    scale_y_log10() +
    geom_point(aes(shape = Subtype)) +
    scale_color_manual(values = brca.mids) + 
    facet_wrap(~TimePoint) +
    labs(
      title = titles[[marker]],
      x = xlabels[[marker]],
      y = "PFS (days)"
    ) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_text_repel() +
    geom_text(data = df.t_other, aes(x = x, y = y, label = txt), size = 5, color = 1) +
    theme_bw()

  pdf(file.path(cycif_fig_dir, "marker_frequency", "pfs_by_state", filenames[[marker]]), width = 12, height = 4)
  print(p)
  dev.off()
}
```

#### Save Objects
```{r}
write.csv(pd.cd8,file=file.path(cycif_fig_dir, "marker_frequency", "pfs_by_state", "pd_cd8.csv"),row.names=F)
write.csv(pd.cd4,file=file.path(cycif_fig_dir, "marker_frequency", "pfs_by_state", "pd_cd4.csv"),row.names=F)
write.csv(pd.t_other,file=file.path(cycif_fig_dir, "marker_frequency", "pfs_by_state", "pd_t_other.csv"),row.names=F)
```

### [EDF6A, EDF6B] T-Cell Frequency vs PFS

`TC_state_1/CD4T_all_pfs_w_pval.pdf`, `TC_state_1/CD8T_all_pfs_w_pval.pdf` -> EDF6A.
`TC_state_1/CD8T_PD1_all_pfs_w_pval.pdf`, `TC_state_1/CD4T_PD1_all_pfs_w_pval.pdf`,
`TC_state_1/T_other_PD1_all_pfs_w_pval.pdf` -> EDF6B.
#### CD8T
```{r}
cd8.stat <- do.call(rbind,cd8.ts)
cd8.stat <- as.data.frame(cd8.stat) %>% 
  rownames_to_column("id") %>%
  mutate(PD1_all=PD1 + PD1_GrB,
         GrB_all=GrB + PD1_GrB)
colnames(cd8.stat)[-1] <- paste0("CD8T_",colnames(cd8.stat)[-1])

non_cd8t.stat <- do.call(rbind,non_cd8t.ts)
non_cd8t.stat <- as.data.frame(non_cd8t.stat) %>% 
  rownames_to_column("id") %>%
  mutate(PD1_all=PD1 + PD1_GrB,
         GrB_all=GrB + PD1_GrB)
colnames(non_cd8t.stat)[-1] <- paste0("Non_CD8T_",colnames(cd8.stat)[-1])

cts5.1 <- table(cell_types(cs5)) %>%
  as.data.frame() %>%
  spread(cell_types,Freq) %>%
  rename(id="sample") %>%
  left_join(cd8.stat,by="id")

n.cts5 <- t(apply(cts5.1[-1],1,function(x)x/sum(x[2:12]))) %>% 
  as.data.frame() %>%
  mutate(id=cts5.1$id) %>%
  relocate(id) %>%
  left_join(pd5,by="id") %>%
  rename(CD8T_all="CD8T")

t.subs <- c("all","PD1_all","PD1_GrB","GrB_all","PD1","GrB","dn")  

r.sq.cd8 <- p.vals.cd8 <- list()
xys <- list(
  all = list(x=c(0.2,0.2,0.2,0.2,0.2,0.2),y=c(600,75,600,75,600,75)),
  PD1_all = list(x=c(0.1,0.1,0.1,0.1,0.1,0.1),y=c(600,75,600,75,600,75)),
  PD1_GrB = list(x=c(0.01,0.01,0.01,0.01,0.01,0.01),y=c(600,75,600,75,600,75)),
  GrB_all = list(x=c(0.01,0.01,0.01,0.01,0.01,0.01),y=c(600,75,600,75,600,75)),
  PD1 = list(x=c(0.1,0.1,0.1,0.1,0.1,0.1),y=c(600,75,600,75,600,75)),
  GrB = list(x=c(0.0025,0.0025,0.0025,0.0025,0.0025,0.0025),y=c(600,75,600,75,600,75)),
  dn = list(x=c(0.1,0.1,0.1,0.1,0.1,0.1),y=c(600,75,600,75,600,75))
)

xys.3 <- list(
  all = list(x=c(0.128,0.128),y=c(300,90))
)


pretty_titles <- c(
  all = "all CD8T cells",
  PD1_all = "PD-1+ CD8T cells",
  PD1_GrB = "PD-1+/GrB+ CD8T cells",
  GrB_all = "GrB+ CD8T cells",
  PD1 = "PD-1+/GrB- CD8T cells",
  GrB = "PD-1-/GrB+ CD8T cells",
  dn = "PD-1-/GrB- CD8T cells"
)

for(ct in t.subs){
  ct1 <- paste0("CD8T_",ct)
  r.sq.cd8[[ct1]] <- p.vals.cd8[[ct1]] <- p.vals
  for(brca in c("MUT","WT")){
    for(tp in c("BS","BX2","BX3")){
      x <- n.cts5 %>% filter(BRCA==brca,TimePoint==tp) %>% pull(!!sym(ct1))
      y <- n.cts5 %>% filter(BRCA==brca,TimePoint==tp) %>% pull(PFS_days)
      lm1 <- lm(log(y) ~ x)
      
      p.vals.cd8[[ct1]][brca,tp] <- summary(lm1)$coefficients[2,4]
      r.sq.cd8[[ct1]][brca,tp] <- summary(lm1)$r.squared
    }
  }

  df.cd8 <- reshape2::melt(p.vals.cd8[[ct1]], c("BRCA", "TimePoint"), value.name = "p-val") %>%
    left_join(reshape2::melt(r.sq.cd8[[ct1]], c("BRCA", "TimePoint"), value.name = "r-sq"),by=c("BRCA","TimePoint")) %>%
    mutate(txt=paste0("p=",round(`p-val`,3),"\nR2=",round(`r-sq`,3))) %>%
    mutate(x=xys[[ct]]$x,y=xys[[ct]]$y)
  
  df.cd8.3 <- reshape2::melt(p.vals.cd8[[ct1]], c("BRCA", "TimePoint"), value.name = "p-val") %>%
    left_join(reshape2::melt(r.sq.cd8[[ct1]], c("BRCA", "TimePoint"), value.name = "r-sq"),by=c("BRCA","TimePoint")) %>%
    mutate(txt=paste0("p=",round(`p-val`,3),"\nR2=",round(`r-sq`,3))) %>%
    filter(TimePoint %in% c("BX3")) %>%
    mutate(x=xys.3[[ct]]$x,y=xys.3[[ct]]$y)
  
  p <- ggplot(n.cts5,aes(x=!!sym(ct1),y=PFS_days,color=BRCA,label=Patient.ID)) +
    scale_y_log10() +
    geom_point(aes(shape=Subtype)) +
    scale_color_manual(values=brca.mids) +
    facet_wrap(~TimePoint) +
    labs(
      title = pretty_titles[[ct]],
      x = "Frequency (out of all cells)",
      y = "PFS (days)"
    ) +
    geom_smooth(method="lm",se=FALSE) +
    geom_text_repel() +
    geom_text(data=df.cd8,aes(x=x,y=y,label=txt),size=5,color=1) +
    theme_bw()

  pdf(file.path(cycif_fig_dir, "marker_frequency", "pfs_by_frequency", paste0(ct1,"_pfs_w_pval.pdf")),width=12,height=4)
  print(p)
  dev.off()
  if(ct == "all"){
    p <- ggplot(n.cts5 %>% filter(TimePoint == "BX3"),aes(x=!!sym(ct1),y=PFS_days,color=BRCA,label=Patient.ID)) +
      scale_y_log10() +
      geom_point(aes(shape=Subtype)) +
      scale_color_manual(values=brca.mids) +
      labs(
        title = pretty_titles[[ct]],
        x = "Frequency (out of all cells)",
        y = "PFS (days)"
      ) +
      geom_smooth(method="lm",se=FALSE) +
      geom_text_repel() +
      geom_text(data=df.cd8.3,aes(x=x,y=y,label=txt),size=5,color=1) +
      theme_bw() + 
      theme(legend.position="none")
  
    pdf(file.path(cycif_fig_dir, "marker_frequency", "pfs_by_frequency", paste0(ct1,"_pfs_w_pval_bx3.pdf")),width=4,height=4)
    print(p)
    dev.off()
  }
}
```

#### CD4T
```{r}
cd4.stat <- do.call(rbind,cd4.ts)
cd4.stat <- as.data.frame(cd4.stat) %>% 
  rownames_to_column("id") %>%
  mutate(PD1_all=PD1 + PD1_GrB,
         GrB_all=GrB + PD1_GrB)
colnames(cd4.stat)[-1] <- paste0("CD4T_",colnames(cd4.stat)[-1])

cts7.1 <- table(cell_types(cs7)) %>%
  as.data.frame() %>%
  spread(cell_types,Freq) %>%
  rename(id="sample") %>%
  left_join(cd4.stat,by="id")

n.cts7 <- t(apply(cts7.1[-1],1,function(x)x/sum(x[2:12]))) %>% 
  as.data.frame() %>%
  mutate(id=cts7.1$id) %>%
  relocate(id) %>%
  left_join(pd7,by="id") %>%
  rename(CD4T_all="CD4T")

t.subs <- c("all","PD1_all","GrB_all","PD1_GrB","PD1","GrB","dn")  

r.sq.cd4 <- p.vals.cd4 <- list()
xys <- list(
  all = list(x=c(0.2,0.2,0.2,0.2,0.2,0.2),y=c(600,75,600,75,600,75)),
  PD1_all = list(x=c(0.1,0.1,0.1,0.1,0.1,0.1),y=c(600,75,600,75,600,75)),
  GrB_all = list(x=c(0.01,0.01,0.01,0.01,0.01,0.01),y=c(600,75,600,75,600,75)),
  PD1_GrB = list(x=c(0.01,0.01,0.01,0.01,0.01,0.01),y=c(600,75,600,75,600,75)),
  PD1 = list(x=c(0.1,0.1,0.1,0.1,0.1,0.1),y=c(600,75,600,75,600,75)),
  GrB = list(x=c(0.0025,0.0025,0.0025,0.0025,0.0025,0.0025),y=c(600,75,600,75,600,75)),
  dn = list(x=c(0.1,0.1,0.1,0.1,0.1,0.1),y=c(600,75,600,75,600,75))
)
  
pretty_titles <- c(
  all = "all CD4T cells",
  PD1_all = "PD-1+ CD4T cells",
  GrB_all = "GrB+ CD4T cells",
  PD1_GrB = "PD-1+/GrB+ CD4T cells",
  PD1 = "PD-1+/GrB- CD4T cells",
  GrB = "PD-1-/GrB+ CD4T cells",
  dn = "PD-1-/GrB- CD4T cells"
)

for(ct in t.subs){
  ct1 <- paste0("CD4T_",ct)
  r.sq.cd4[[ct1]] <- p.vals.cd4[[ct1]] <- p.vals
  for(brca in c("MUT","WT")){
    for(tp in c("BS","BX2","BX3")){
      x <- n.cts7 %>% filter(BRCA==brca,TimePoint==tp) %>% pull(!!sym(ct1))
      y <- n.cts7 %>% filter(BRCA==brca,TimePoint==tp) %>% pull(PFS_days)
      lm1 <- try(lm(log(y) ~ x))
        
      if(nrow(summary(lm1)$coefficients)==2){
        p.vals.cd4[[ct1]][brca,tp] <- summary(lm1)$coefficients[2,4]
        r.sq.cd4[[ct1]][brca,tp] <- summary(lm1)$r.squared
      }else{
        p.vals.cd4[[ct1]][brca,tp] <- NA
        r.sq.cd4[[ct1]][brca,tp] <- NA
      }
    }
  }

  df.cd4 <- reshape2::melt(p.vals.cd4[[ct1]], c("BRCA", "TimePoint"), value.name = "p-val") %>%
    left_join(reshape2::melt(r.sq.cd4[[ct1]], c("BRCA", "TimePoint"), value.name = "r-sq"),by=c("BRCA","TimePoint")) %>%
    mutate(txt=paste0("p=",round(`p-val`,3),"\nR2=",round(`r-sq`,3))) %>%
    mutate(x=xys[[ct]]$x,y=xys[[ct]]$y)
    
  df.cd4.3 <- reshape2::melt(p.vals.cd4[[ct1]], c("BRCA", "TimePoint"), value.name = "p-val") %>%
    left_join(reshape2::melt(r.sq.cd4[[ct1]], c("BRCA", "TimePoint"), value.name = "r-sq"),by=c("BRCA","TimePoint")) %>%
    mutate(txt=paste0("p=",round(`p-val`,3),"\nR2=",round(`r-sq`,3))) %>%
    filter(TimePoint %in% c("BX3")) %>%
    mutate(x=xys.3[[ct]]$x,y=xys.3[[ct]]$y)

  p <- ggplot(n.cts7,aes(x=!!sym(ct1),y=PFS_days,color=BRCA,label=Patient.ID)) +
    scale_y_log10() +
    geom_point(aes(shape=Subtype)) +
    scale_color_manual(values=brca.mids) +
    facet_wrap(~TimePoint) +
    labs(
      title = pretty_titles[[ct]],
      x = "Frequency (out of all cells)",
      y = "PFS (days)"
    ) +
    geom_smooth(method="lm",se=FALSE) +
    geom_text_repel() +
    geom_text(data=df.cd4,aes(x=x,y=y,label=txt),size=5,color=1) +
    theme_bw()

  pdf(file.path(cycif_fig_dir, "marker_frequency", "pfs_by_frequency", paste0(ct1,"_pfs_w_pval.pdf")),width=12,height=4)
  print(p)
  dev.off()
  if(ct == "all"){
    p <- ggplot(n.cts7 %>% filter(TimePoint == "BX3"),aes(x=!!sym(ct1),y=PFS_days,color=BRCA,label=Patient.ID)) +
      scale_y_log10() +
      geom_point(aes(shape=Subtype)) +
      scale_color_manual(values=brca.mids) +
      labs(
        title = pretty_titles[[ct]],
        x = "Frequency (out of all cells)",
        y = "PFS (days)"
      ) +
      geom_smooth(method="lm",se=FALSE) +
      geom_text_repel() +
      geom_text(data=df.cd4.3,aes(x=x,y=y,label=txt),size=5,color=1) +
      theme_bw() + 
      theme(legend.position="none")
  
    pdf(file.path(cycif_fig_dir, "marker_frequency", "pfs_by_frequency", paste0(ct1,"_pfs_w_pval_bx3.pdf")),width=4,height=4)
    print(p)
    dev.off()
  }
}
```

#### T_other and their subtypes
```{r}
t_other.stat <- do.call(rbind,t_other.ts)
t_other.stat <- as.data.frame(t_other.stat) %>% 
  rownames_to_column("id") %>%
  mutate(PD1_all=PD1 + PD1_GrB,
         GrB_all=GrB + PD1_GrB)
colnames(t_other.stat)[-1] <- paste0("T_other_",colnames(t_other.stat)[-1])

cts5.1 <- table(cell_types(cs5)) %>%
  as.data.frame() %>%
  spread(cell_types,Freq) %>%
  rename(id="sample") %>%
  select(-T_other) %>%
  left_join(t_other.stat,by="id")

n.cts5 <- t(apply(cts5.1[-1],1,function(x)x/sum(x))) %>% 
  as.data.frame() %>%
  mutate(id=cts5.1$id) %>%
  relocate(id) %>%
  left_join(pd5,by="id")

t.subs <- c("PD1_all","GrB_all","PD1_GrB","PD1","GrB","dn")  

r.sq.t_other <- p.vals.t_other <- list()
xys <- list(
  PD1_all = list(x=c(0.1,0.1,0.1,0.1,0.1,0.1),y=c(600,75,600,75,600,75)),
  GrB_all = list(x=c(0.01,0.01,0.01,0.01,0.01,0.01),y=c(600,75,600,75,600,75)),
  PD1_GrB = list(x=c(0.0025,0.0025,0.0025,0.0025,0.0025,0.0025),y=c(600,75,600,75,600,75)),
  PD1 = list(x=c(0.1,0.1,0.1,0.1,0.1,0.1),y=c(600,75,600,75,600,75)),
  GrB = list(x=c(0.0025,0.0025,0.0025,0.0025,0.0025,0.0025),y=c(600,75,600,75,600,75)),
  dn = list(x=c(0.075,0.075,0.075,0.075,0.075,0.075),y=c(600,75,600,75,600,75))
)

pretty_titles <- c(
  PD1_all = "PD-1+ T_other cells",
  GrB_all = "GrB+ T_other cells",
  PD1_GrB = "PD-1+/GrB+ T_other cells",
  PD1 = "PD-1+/GrB- T_other cells",
  GrB = "PD-1-/GrB+ T_other cells",
  dn = "PD-1-/GrB- T_other cells"
)
  
for(ct in t.subs){
  ct1 <- paste0("T_other_",ct)
  r.sq.t_other[[ct1]] <- p.vals.t_other[[ct1]] <- p.vals
  for(brca in c("MUT","WT")){
    for(tp in c("BS","BX2","BX3")){
      x <- n.cts5 %>% filter(BRCA==brca,TimePoint==tp) %>% pull(!!sym(ct1))
      y <- n.cts5 %>% filter(BRCA==brca,TimePoint==tp) %>% pull(PFS_days)
      lm1 <- lm(log(y) ~ x)
      
      p.vals.t_other[[ct1]][brca,tp] <- summary(lm1)$coefficients[2,4]
      r.sq.t_other[[ct1]][brca,tp] <- summary(lm1)$r.squared
    }
  }

  df.t_other <- reshape2::melt(p.vals.t_other[[ct1]], c("BRCA", "TimePoint"), value.name = "p-val") %>%
    left_join(reshape2::melt(r.sq.t_other[[ct1]], c("BRCA", "TimePoint"), value.name = "r-sq"),by=c("BRCA","TimePoint")) %>%
    mutate(txt=paste0("p=",round(`p-val`,3),"\nR2=",round(`r-sq`,3))) %>%
    mutate(x=xys[[ct]]$x,y=xys[[ct]]$y)
  
  p <- ggplot(n.cts5,aes(x=!!sym(ct1),y=PFS_days,color=BRCA,label=Patient.ID)) +
    scale_y_log10() +
    geom_point(aes(shape=Subtype)) +
    scale_color_manual(values=brca.mids) +
    facet_wrap(~TimePoint) +
    labs(title=ct1,x="Frequency (out of all cells)",y="PFS (days)") +
    geom_smooth(method="lm",se=FALSE) +
    geom_text_repel() +
    geom_text(data=df.t_other,aes(x=x,y=y,label=txt),size=5,color=1) +
    theme_bw()

  pdf(file.path(cycif_fig_dir, "marker_frequency", "pfs_by_frequency", paste0(ct1,"_pfs_w_pval.pdf")),width=12,height=4)
  print(p)
  dev.off()
}
```

#### T_all and their subtypes
```{r}
t_all.stat <- (cd8.stat %>% arrange(id) %>% select(-id) +
               non_cd8t.stat %>% arrange(id) %>% select(-id)) %>%
  cbind(id = sort(cd8.stat$id)) %>%
  relocate(id)
colnames(t_all.stat)[-1] <- paste0("T_all_",t.subs[c(5,3,4,6,1,2)])

cts5 <- table(cell_types(cs5)) %>%
  as.data.frame() %>%
  spread(cell_types,Freq) %>%
  rename(id="sample") %>%
  select(-c(CD8T,T_other)) %>%
  left_join(t_all.stat,by="id")

n.cts5 <- t(apply(cts5[-1],1,function(x)x/sum(x))) %>% 
  as.data.frame() %>%
  mutate(id=cts5$id) %>%
  relocate(id) %>%
  left_join(pd5,by="id")

t.subs <- c("PD1_all","GrB_all","PD1_GrB","PD1","GrB","dn")  

r.sq.t_all <- p.vals.t_all <- list()
xys <- list(
  PD1_all = list(x=c(0.1,0.1,0.1,0.1,0.1,0.1),y=c(600,75,600,75,600,75)),
  GrB_all = list(x=c(0.01,0.01,0.01,0.01,0.01,0.01),y=c(600,75,600,75,600,75)),
  PD1_GrB = list(x=c(0.0025,0.0025,0.0025,0.0025,0.0025,0.0025),y=c(600,75,600,75,600,75)),
  PD1 = list(x=c(0.1,0.1,0.1,0.1,0.1,0.1),y=c(600,75,600,75,600,75)),
  GrB = list(x=c(0.0025,0.0025,0.0025,0.0025,0.0025,0.0025),y=c(600,75,600,75,600,75)),
  dn = list(x=c(0.075,0.075,0.075,0.075,0.075,0.075),y=c(600,75,600,75,600,75))
)
  
for(ct in t.subs){
  ct1 <- paste0("T_all_",ct)
  r.sq.t_all[[ct1]] <- p.vals.t_all[[ct1]] <- p.vals
  for(brca in c("MUT","WT")){
    for(tp in c("BS","BX2","BX3")){
      x <- n.cts5 %>% filter(BRCA==brca,TimePoint==tp) %>% pull(!!sym(ct1))
      y <- n.cts5 %>% filter(BRCA==brca,TimePoint==tp) %>% pull(PFS_days)
      lm1 <- lm(log(y) ~ x)
      
      p.vals.t_all[[ct1]][brca,tp] <- summary(lm1)$coefficients[2,4]
      r.sq.t_all[[ct1]][brca,tp] <- summary(lm1)$r.squared
    }
  }

  df.t_all <- reshape2::melt(p.vals.t_all[[ct1]], c("BRCA", "TimePoint"), value.name = "p-val") %>%
    left_join(reshape2::melt(r.sq.t_all[[ct1]], c("BRCA", "TimePoint"), value.name = "r-sq"),by=c("BRCA","TimePoint")) %>%
    mutate(txt=paste0("p=",round(`p-val`,3),"\nR2=",round(`r-sq`,3))) %>%
    mutate(x=xys[[ct]]$x,y=xys[[ct]]$y)
  
  p <- ggplot(n.cts5,aes(x=!!sym(ct1),y=PFS_days,color=BRCA,label=Patient.ID)) +
    scale_y_log10() +
    geom_point(aes(shape=Subtype)) +
    scale_color_manual(values=brca.mids) +
    facet_wrap(~TimePoint) +
    labs(title=ct1,x="Frequency (out of all cells)",y="PFS (days)") +
    geom_smooth(method="lm",se=FALSE) +
    geom_text_repel() +
    geom_text(data=df.t_all,aes(x=x,y=y,label=txt),size=5,color=1) +
    theme_bw()

  pdf(file.path(cycif_fig_dir, "marker_frequency", "pfs_by_frequency", paste0(ct1,"_pfs_w_pval.pdf")),width=12,height=4)
  print(p)
  dev.off()
}
```


All UMAP-based analysis (All Cells, Tumor Cells, Macrophages, and the Save UMAP
Objects/Coordinates step) is in `4_cycif_umap_analysis.Rmd`, which loads
`cs7`/`pd7`/`cts7`/`ct.cols7`/`pt.cols` directly from `cs7.rds` (the same checkpoint file
this notebook uses) rather than depending on an export from here.
