---
title: "TALAVE CyCIF analysis: UMAP Analysis (All Cells, Tumor, Macrophages)"
author: "Kenichi Shimada"
---

# TALAVE CyCIF analysis

Runs UMAP-based cell-state analysis (All Cells, Tumor Cells, Macrophages), producing the
macrophage clustering (`cluster_mac`) that `5` merges with its own TME clustering for `6`.

**Required run order: `3` -> `4` (this notebook) -> `5` -> `6`.**
No dependency on `3` or `5` -- only needs the base `cs7`/`frnn7.cmb` checkpoint.

`cs7.rds` and `cycif_spatial_frnn_input_data.rda` (frnn7.cmb, frnn7.exp, frnn7.rcn,
is.sele.7) are two separate checkpoint files, also loaded directly by `3` and `6`. This
notebook doesn't use `cs5` at all, so `cs5.rds` is never loaded here. `pd7`/`cts7`/`ct.cols7`
are one-line derivations from `cs7`, defined locally below rather than saved in the
checkpoint.

## ---- Initial Setup ----

### Load Configuration
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Bootstrap
```{r bootstrap}
source(file.path(root_dir, "functions", "bootstrap.R"))
initialize_project()

if (!requireNamespace("CycifAnalyzeR", quietly = TRUE)) {
  remotes::install_github("<org>/CycifAnalyzeR")
}
library(CycifAnalyzeR)
source(file.path(root_dir, "functions", "plotting.R"))
source(file.path(source_dir, "analysis_cycif.R"))
source(file.path(source_dir, "io.R"))
source(file.path(source_dir, "plotting_cycif.R"))
source(file.path(source_dir, "qc_cycif.R"))

brca.fills = c(MUT="#C9C0DD", WT="#D9DADA")
brca.cols = c(MUT="#62439c",WT="#58575f")
brca.mids = c(MUT="#62439c",WT="grey50")
```

## ---- Load Inputs ----

```{r}
cs7 <- readRDS(file.path(cycif_obj_dir, "cs7.rds"))
load(file.path(cycif_obj_dir, "cycif_spatial_frnn_input_data.rda")) # frnn7.cmb, frnn7.exp, frnn7.rcn, is.sele.7

pd7 <- pData(cs7)
cts7 <- cell_types(cs7)
ct.cols7 <- get_cycif_ct_colors()
pt.cols <- load_pt_colors()
```

## ---- UMAP Analysis ----
### Load CN Data
```{r}
x <- load(file=file.path(cycif_obj_dir,"cycif7_neighborhood_profiles.rda")) # frnn7.rcn,frnn7.exp
frnn7.rcn.n <- rowSums(frnn7.rcn)
is.ge10.7 <- frnn7.rcn.n >= 10
```

```{r}
exp.1 <- exprs(cs7, type="log") %>% cbind(cts7) %>% filter(cell_types != "outOfROI") 
exp.2 <- exprs(cs7, type="logTh") %>%  mutate(across(everything(), ~ . > 0.5)) %>% cbind(cts7) %>% filter(cts7$cell_types != "outOfROI")

m.exp.1 <- exp.1 %>% filter(is.sele.7)
m.exp.2 <- exp.2 %>% filter(is.sele.7)

cst <- cs7@cell_types$default@cell_state_def
```

### Define UMAP Plotting Helpers
`plotting_cycif.R` defines the UMAP plotting helpers used below.

### 1. All Cells
```{r}
used.abs <- names(exp.1)[4:21]

umap <- build_umap_embedding(m.exp.1, features = used.abs, n_neighbors = 20, seed = 123)

dt.all <- bind_umap_to_tables(
  df_list = list(m.exp.1, m.exp.2),
  umap = umap,
  pd = pd7
)
dt.all.1 <- dt.all[[1]]

```
#### [EDF4A, EDF4B] All-Cell UMAP Outputs

```{r}
if(!dir.exists(file.path(cycif_fig_dir, "umap", "all", "metadata"))){
  dir.create(file.path(cycif_fig_dir, "umap", "all", "metadata"),recursive=T)
  dir.create(file.path(cycif_fig_dir, "umap", "all", "ab"),recursive=T)
}
p1_full <- umap_base(dt.all.1, aes(x = umap_x, y = umap_y, col = cell_types)) +
  scale_color_manual(values = ct.cols7) +
  ggtitle("Cell types")
p1 <- p1_full + theme(legend.position = "none")
save_legend_pdf(p1_full, file.path(cycif_fig_dir, "umap", "all", "metadata", "umap_cts_legend.pdf"), width = 4, height = 6)  
ggsave(file.path(cycif_fig_dir, "umap", "all", "metadata", "umap_cts.png"), p1, width = 9, height = 10, units ="in", dpi = 300)

p2_full <- umap_base(dt.all.1, aes(x = umap_x, y = umap_y, col = sample)) +
  ggtitle("Samples")
p2 <- p2_full + theme(legend.position = "none")
save_legend_pdf(p2_full, file.path(cycif_fig_dir, "umap", "all", "metadata", "umap_smpls_legend.pdf"), width = 4, height = 8)  
ggsave(file.path(cycif_fig_dir, "umap", "all", "metadata", "umap_sample.png"), p2, width = 9, height = 10, units ="in", dpi = 300)

p3_full <- umap_base(dt.all.1, aes(x = umap_x, y = umap_y, col = TimePoint)) +
  scale_color_brewer(palette = "Set1") +
  ggtitle("TimePoint, stratified by BRCA")
p3 <- p3_full + theme(legend.position = "none")
p3 <- p3 + facet_wrap(~BRCA) + theme(strip.text = element_text(size = 36))
save_legend_pdf(p3_full, file.path(cycif_fig_dir, "umap", "all", "metadata", "umap_bt3_wide_legend.pdf"), width = 4, height = 6)  
ggsave(file.path(cycif_fig_dir, "umap", "all", "metadata", "umap_bt3_wide.png"), p3, width = 14, height = 7, units = "in", dpi = 300)


```
##### Protein Expression: Numerical
```{r}
for(ab1 in used.abs){
  dtt <- dt.all.1 %>% 
    select(umap_x,umap_y,!!sym(ab1)) %>%
    rename(ab=ab1)
  abx <- dtt$ab
  th <- quantile(abx,c(.05,.95))
  abx[abx < th[1]] <- th[1]
  abx[abx > th[2]] <- th[2]
  
  dtt$ab <- abx

  p <- ggplot(dtt,aes(x=umap_x,y=umap_y,col=ab)) + 
    geom_point(size=.1) +
    coord_fixed(ratio=1) +
    viridis::scale_color_viridis(guide = "none") +
    theme_minimal() +
    ggtitle(ab1) + 
    umap_styling_1(
      title_size = 80,
      legend_title_margin = FALSE,
      legend_key_height = 1.5
    )

  if (ab1 == used.abs[1]) {
    p_full <- ggplot(dtt, aes(x = umap_x, y = umap_y, col = ab)) + 
      geom_point(size = 0.1) +
      coord_fixed(ratio = 1) +
      viridis::scale_color_viridis(name = ab1) +
      theme_minimal() +
      ggtitle(ab1) + 
      umap_styling_1(
        title_size = 80,
        legend_title_margin = FALSE,
        legend_key_height = 1.5
      )

    save_legend_pdf(p_full,
      file.path(cycif_fig_dir, "umap", "all", "ab", paste0("legend__", ab1, ".pdf")),
      width = 4, height = 6
    )
  }    
  
  png(file.path(cycif_fig_dir, "umap", "all", "ab", paste0("umap__",ab1,".png")),
      width=1000,height=1000)
  print(p)
  dev.off()
}
```

### 2. Tumor Cells
```{r}
this.cts <- c("Tumor_panCK","Tumor_neg")
rs <- rowSums(table(cts7)[,this.cts])
rs.all <- rowSums(table(cts7))

n.cells <- cbind(rs=rs,all=rs.all) # identical

n.used <- quantile(n.cells[,"rs"],.2)

set.seed(12345)
lst.sele.tum <- sapply(rs,function(x){
  n1 <- min(x,n.used)
  sele <- sample(x,n1)
  is.sele <- seq(x) %in% sele
  return(is.sele)
})
is.sele.tum <- unlist(lst.sele.tum)
```


```{r}
used.abs1 <- c("panCK",names(cst)[apply(cst[this.cts,]=="CAN",2,function(x)any(!is.na(x) & x))])
tum.exp.1 <- exp.1 %>%
  filter(cell_types %in% this.cts) %>%
  filter(is.sele.tum)
tum.exp.2 <- exp.2 %>%
  filter(cell_types %in% this.cts) %>%
  filter(is.sele.tum)

umap2 <- build_umap_embedding(tum.exp.1, features = used.abs1, n_neighbors = 20, seed = 123)

dt.tum <- bind_umap_to_tables(
  df_list = list(tum.exp.1, tum.exp.2),
  umap = umap2,
  pd = pd7,
  include_bor = TRUE
)
dt.tum.1 <- dt.tum[[1]]
dt.tum.2 <- dt.tum[[2]]
```
#### [EDF2H, Fig 3F, EDF3E, EDF3F] Tumor UMAP Outputs
##### Volcano Plot Between Tumor_panCK and Tumor_neg

Per-marker `lmer(marker ~ cell_types + (1 | sample))` comparing Tumor_neg vs Tumor_panCK.
P-values are deliberately left uncorrected (nominal p only): small marker family,
single-cell-level panel with only 2 samples.
```{r}
dt.tum.1.1 <- dt.tum.1 %>%
  filter(sample %in% c("3005-3","1014-3")) %>%
  filter(cell_types %in% this.cts) %>%
  select(sample,cell_types,!!!syms(used.abs1[-1]))

markers <- setdiff(names(dt.tum.1.1), c("sample", "cell_types"))

dt_sub <- dt.tum.1.1 %>%
  filter(cell_types %in% c("Tumor_panCK", "Tumor_neg"))

results <- lapply(markers, function(m) {
  formula <- as.formula(paste0(m, " ~ cell_types + (1 | sample)"))
  fit <- lmer(formula, data = dt_sub)
  coef_df <- tidy(fit, effects = "fixed") %>%
    filter(term == "cell_typesTumor_neg") %>%
    mutate(marker = m)
  return(coef_df)
})

res_df <- bind_rows(results) %>%
  select(marker, estimate, std.error, statistic, p.value)

p <- ggplot(res_df, aes(x = estimate, y = -log10(p.value), label = marker)) +
  geom_point(aes(color = p.value < 0.05), size = 3) +
  geom_text_repel(size=5) +
  geom_vline(xintercept = 0, linetype = "solid", color = "grey20") +
  geom_hline(yintercept = 0, linetype = "solid", color = "grey20") +  
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",color = "grey50") +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")) +
  labs(x = "Mean Difference (Tumor_neg - Tumor_pos)",
       y = "-log10(p-value)",
       title = "Volcano Plot (p < 0.05)",
       color = "p < 0.05") +
  theme_minimal() +
  theme(legend.position = "none")

pdf(file.path(cycif_fig_dir, "umap", "tumor", "volcano_plot_tumor_panCK_vs_neg_ori.pdf"),width=5,height=7)
print(p)
dev.off()
```
##### Density-Based Co-Expression
```{r}
coex_res <- prepare_density_coexpression_df(
  dt.tum.1 %>% select(!!!syms(c(used.abs1)))
)
expr_mat <- coex_res$expr_mat
coex_tum <- coex_res$coex_mat
rho <- coex_res$rho
r <- coex_res$r

p <- ggplot(expr_mat, aes(x = gH2AX, y = pTBK1, color = density)) +
  geom_point(size = 1.5) +
  scale_color_identity() +
  annotate(
    "text",
    x = Inf, y = -Inf,
    label = paste0(
      "Pearson r = ", round(r, 2),"\n",
      "Spearman \u03C1 = ", round(rho, 2)
    ),
    hjust = 1.1, vjust = -0.7,
    size = 10
  ) +
  theme_minimal() +
  labs(
    title = "γH2AX vs pTBK1 expression (Tumor cells)",
    x = expression(gamma*"H2AX expression (log)"),
    y = "pTBK1 expression (log)"
  ) +
  theme(
    plot.title = element_text(size = 30, hjust = 0.5, color = "black"),
    axis.text = element_text(size = 20, color = "black"),
    axis.title = element_text(size = 24, color = "black"),
    axis.key.size = unit(1.2, "cm")
  )

ggsave(file.path(cycif_fig_dir, "umap", "tumor", "gh2ax_ptbk1_tum_v2.png"),
       p, width = 9, height = 10,  units ="in", dpi = 300)
```

```{r}
mean_bt3 <- prepare_sample_marker_means_df(
  df = dt.tum.1,
  markers = c("gH2AX", "pTBK1")
)
fit_bt3 <- lm(pTBK1_mean ~ gH2AX_mean, data = mean_bt3)
summary(fit_bt3)

cor_bt3 <- cor.test(
  mean_bt3$gH2AX_mean,
  mean_bt3$pTBK1_mean,
  method = "spearman"
)

rho_bt3  <- cor_bt3$estimate
p_bt3    <- cor_bt3$p.value

cor_bt3_pearson <- cor.test(
  mean_bt3$gH2AX_mean,
  mean_bt3$pTBK1_mean,
  method = "pearson"
)

r_bt3  <- cor_bt3_pearson$estimate
p_bt3    <- cor_bt3_pearson$p.value

p <- ggplot(mean_bt3, aes(x = gH2AX_mean, y = pTBK1_mean)) +

  # bt3 mean points
  geom_point(aes(color = bt3), size = 4) +

  # regression line across bt3 means
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black",
    linewidth = 1.5
  ) +

  theme_minimal() +
  labs(
    title = "γH2AX vs pTBK1 expression (Tumor cells)",
    x = expression(gamma*"H2AX expression (log)"),
    y = "pTBK1 expression (log)"
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf,
      label = paste0("Pearson r = ", round(r_bt3, 2),
                   "\nSpearman \u03C1  = ", round(rho_bt3,2)),
    size = 9,
    hjust = 1,vjust=-0.8
  ) +
  theme(
    plot.title = element_text(size = 30, hjust = 0.5, color = "black"),
    axis.text = element_text(size = 20, color = "black"),
    axis.title = element_text(size = 24, color = "black")
  )

ggsave(file.path(cycif_fig_dir, "umap", "tumor", "gh2ax_ptbk1_bt3_means_pts.pdf"), 
       p, width = 10, height = 10,  units ="in", dpi = 300, device = cairo_pdf)
```

##### Mean Expression Summary
```{r}
p.gh2ax <- plot_ct(mean_bt3,y_var="gH2AX_mean",y_label="gH2AX (mean per sample)",title="gH2AX in tumor cells (CyCIF)",pt.cols=pt.cols)

pdf(file.path(cycif_fig_dir, "umap", "tumor", "gH2AX_mean.pdf"),width=5,height=7)
print(p.gh2ax$plot)
dev.off()

p.ptbk1 <- plot_ct(mean_bt3,y_var="pTBK1_mean",y_label="pTBK1 (mean per sample)",title="pTBK1 in tumor cells (CyCIF)",pt.cols=pt.cols)
pdf(file.path(cycif_fig_dir, "umap", "tumor", "pTBK1_mean.pdf"),width=5,height=7)
print(p.ptbk1$plot)
dev.off()
```

### 3. Macrophages
```{r}
this.cts2 <- c("Mac_CD68","Mac_CD68_CD163","Mac_CD163")
rs <- rowSums(table(cts7)[,this.cts2])
n.cells <- cbind(rs=rs,all=rs.all) # identical

n.used <- round(quantile(n.cells[,"rs"],.2))

set.seed(12345)
lst.sele.mac <- sapply(rs,function(x){
  n1 <- min(x,n.used)
  sele <- sample(x,n1)
  is.sele <- seq(x) %in% sele
  return(is.sele)
})
is.sele.mac <- unlist(lst.sele.mac)
```

```{r}
cell_types <- cts7[, "cell_types"]
idx.split <- split(seq_along(cell_types), cell_types)
this.cts2 <- c("Mac_CD68", "Mac_CD68_CD163", "Mac_CD163")
idx.mac.all <- unlist(idx.split[this.cts2])
is.sele.full <- rep(FALSE, length(cell_types))
is.sele.full[idx.mac.all] <- is.sele.mac
is.sele.full <- is.sele.full[!is.na(cell_types)]
```

#### Manual Subclassification by PDL1 and cCaspase3

Tumor cells and macrophages were manually classified into PDL1-/PDL1+CC3-/PDL1+CC3+ by
interactively clicking boundary points on a PDL1-vs-cCaspase3 scatter plot (`plot_scatter_locator_base()`/
`split_by_clicks()` in `functions/plotting_cycif.R`, applied to `frnn7.cmb.mac`/`frnn7.cmb.tum`); the
resulting decision boundary (a PDL1 threshold plus a cCaspase3-vs-PDL1 curve) is saved at
`figures/cycif/exp_2/umap_mac/pdl1_cc3_boundary_coords.rds` and reused (not re-clicked) wherever this
classification is applied downstream (see `6_cycif_spatial_continued.Rmd`).

#### Prepare Macrophage UMAP Data
```{r}
used.abs2 <- c("CD68","CD163",names(cst)[apply(cst[this.cts2,]=="CAN",2,function(x)any(!is.na(x) & x))])
mac.exp.1 <- exp.1 %>%
  filter(cell_types %in% this.cts2) %>%
  filter(is.sele.mac)
mac.exp.2 <- exp.2 %>%
  filter(cell_types %in% this.cts2) %>%
  filter(is.sele.mac)

umap3 <- build_umap_embedding(mac.exp.1, features = used.abs2, n_neighbors = 20, seed = 123)

dt.mac <- bind_umap_to_tables(
  df_list = list(mac.exp.1, mac.exp.2),
  umap = umap3,
  pd = pd7,
  include_bor = TRUE
)
dt.mac.1 <- dt.mac[[1]]
dt.mac.2 <- dt.mac[[2]]

```

#### [Fig 5C, 5E, 5F, 5G, EDF8F, EDF8H] Macrophage UMAP Outputs
```{r}
if(!dir.exists(file.path(cycif_fig_dir, "umap", "mac", "metadata"))){
  dir.create(file.path(cycif_fig_dir, "umap", "mac", "metadata"),recursive=T)
  dir.create(file.path(cycif_fig_dir, "umap", "mac", "ab"),recursive=T)
  dir.create(file.path(cycif_fig_dir, "umap", "mac", "cluster_freq"),recursive=T)
}

p1_full <- umap_base(dt.mac.1, aes(x = umap_x, y = umap_y, col = cell_types), point_size = 0.8) +
  scale_color_brewer(palette="Set1") +
  ggtitle("Cell types")
p1 <- p1_full + theme(legend.position = "none")
save_legend_pdf(p1_full, file.path(cycif_fig_dir, "umap", "mac", "metadata", "umap_cts_legend.pdf"), width = 4, height = 6)  
ggsave(file.path(cycif_fig_dir, "umap", "mac", "metadata", "umap_cts.png"), p1, width = 7, height = 7, units ="in", dpi = 300)

p2_full <- umap_base(dt.mac.1, aes(x = umap_x, y = umap_y, col = sample), point_size = 0.8) +
  ggtitle("Samples")
p2 <- p2_full + theme(legend.position = "none")
save_legend_pdf(p2_full, file.path(cycif_fig_dir, "umap", "mac", "metadata", "umap_smpls_legend.pdf"), width = 4, height = 8)  
ggsave(file.path(cycif_fig_dir, "umap", "mac", "metadata", "umap_sample.png"), p2, width = 7, height = 7, units ="in", dpi = 300)

p3_full <- umap_base(dt.mac.1, aes(x = umap_x, y = umap_y, col = TimePoint), point_size = 0.8) +
  scale_color_brewer(palette = "Set1") +
  ggtitle("TimePoint, stratified by BRCA")
p3 <- p3_full + theme(legend.position = "none")
p3 <- p3 + facet_wrap(~BRCA) + theme(strip.text = element_text(size = 36))
save_legend_pdf(p3_full, file.path(cycif_fig_dir, "umap", "mac", "metadata", "umap_bt3_wide_legend.pdf"), width = 4, height = 6)  
ggsave(file.path(cycif_fig_dir, "umap", "mac", "metadata", "umap_bt3_wide.png"), p3, width = 14, height = 7, units = "in", dpi = 300)

```
##### Protein Expression: Numerical
```{r}
for(ab1 in c(used.abs2,"gH2AX")){
  dtt <- dt.mac.1 %>% 
    select(umap_x,umap_y,!!sym(ab1)) %>%
    rename(ab=ab1)
  abx <- dtt$ab
  
  dtt$ab <- abx
  coords <- dtt %>% select(umap_x, umap_y) %>% as.matrix()
  k <- 20
  knn <- get.knnx(coords, coords, k = k)

  local_abx <- sapply(1:nrow(knn$nn.index), function(i) {
    mean(abx[knn$nn.index[i, ]])
  })

  th <- quantile(local_abx,c(.05,.95))
  local_abx[local_abx < th[1]] <- th[1]
  local_abx[local_abx > th[2]] <- th[2]
  
  dtt$local_abx <- local_abx
  
  p <- ggplot(dtt,aes(x=umap_x,y=umap_y,col=local_abx)) + 
    geom_point(size=0.8) +
    coord_fixed(ratio=1) +
    viridis::scale_color_viridis(guide = "none") +
    theme_minimal() +
    ggtitle(ab1) + 
    umap_styling_1()

  if (ab1 == used.abs2[1]) {
    p_full <- ggplot(dtt, aes(x = umap_x, y = umap_y, col = ab)) +
      geom_point(size = 0.1) +
      coord_fixed(ratio = 1) +
      viridis::scale_color_viridis(name = ab1) +
      theme_minimal() +
      ggtitle(ab1) +
      umap_styling_1()

    save_legend_pdf(p_full,
      file.path(cycif_fig_dir, "umap", "mac", "ab", paste0("legend__", ab1, ".pdf")),
      width = 4, height = 6
    )
  }

  ggsave(file.path(cycif_fig_dir, "umap", "mac", "ab", paste0("umap__",ab1,"_v2.png")),p, width = 7, height = 7, units = "in", dpi = 300)  
}
```
```{r}
make_viridis_legend(
  limits = c(0,5),
  file = file.path(
    cycif_fig_dir,
    "umap",
    "mac",
    "ab",
    "umap_cts_legend_mac_viridis.pdf"
  )
)
```

##### Leiden Clustering

Clusters macrophages by marker expression: k-NN graph (k=20) on the scaled marker matrix,
then Leiden community detection on that graph.
```{r}
mat3 <- exp.1 %>%
  filter(cell_types %in% this.cts2) %>%
  filter(is.sele.mac) %>%
  select(all_of(used.abs2)) %>%
  as.matrix()

k <- 20
mat3_scaled <- scale(mat3)
knn <- get.knn(mat3_scaled,k = k)

edges <- do.call(rbind, lapply(1:nrow(knn$nn.index), function(i) {
  cbind(i, knn$nn.index[i, ])
}))

g <- graph_from_edgelist(edges, directed = FALSE)
g <- igraph::simplify(g)

set.seed(12345)
leiden_res <- leiden_find_partition(g, resolution_parameter = 5e-4,seed=12345)
cluster_labels <- membership(leiden_res)
length(table(cluster_labels))
```

`frnn7.cmb$cluster_mac` is saved below for `5` to merge with its own TME clustering.
```{r}
frnn7.cmb$cluster_mac <- NA
frnn7.cmb$cluster_mac[is.sele.full] <- paste0("Mac_",cluster_labels)
```

```{r}
dt.mac.1 <- dt.mac.1 %>%
  mutate(cluster = factor(cluster_labels, levels = sort(unique(cluster_labels)))) %>%
  mutate(cluster = factor(paste0("Mac_", cluster),levels= paste0("Mac_", sort(unique(cluster_labels))))) 
dt.mac.2 <- dt.mac.2 %>%
  mutate(cluster = factor(cluster_labels, levels = sort(unique(cluster_labels)))) %>%
  mutate(cluster = factor(paste0("Mac_", cluster),levels= paste0("Mac_", sort(unique(cluster_labels)))))

label_df <- dt.mac.2 %>%
  group_by(cluster) %>%
  summarize(
    x = median(umap_x, na.rm = TRUE),
    y = median(umap_y, na.rm = TRUE)
  ) %>% 
  ungroup() %>%
  mutate(cluster = as.numeric(sub("Mac_","",cluster)))

p7 <- umap_base(dt.mac.1, aes(x = umap_x, y = umap_y, col = cluster)) +
  geom_text(data = label_df, aes(x = x, y = y, label = cluster),
            color = "black", size = 10) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  ggtitle("Mac clusters") + 
  theme(legend.position = "none")
  
ggsave(file.path(cycif_fig_dir, "umap", "mac", "metadata", "umap_cts_cluster_10.png"),
       p7, width = 7, height = 7, units = "in", dpi = 300)  
```

##### Sample-Level Frequencies
```{r}

sample_cluster_freq <- dt.mac.1 %>%
  count(sample, cluster) %>%
  complete(sample, cluster, fill = list(n = 0)) %>%
  group_by(sample) %>%
  mutate(rel_freq = n / sum(n)) %>%
  ungroup() %>%
  left_join(pd7 %>% select(id, Patient.ID, BRCA, TimePoint, BOR, bt3), by = c("sample" = "id"))

uniq.clusts <- levels(dt.mac.1$cluster)
d <- file.path(cycif_fig_dir, "umap", "mac", "cluster_freq")
if(!dir.exists(d)){
  dir.create(d,recursive=T)
}

for(cl in uniq.clusts){
  dfx <- sample_cluster_freq %>% filter(cluster == cl) %>% filter(!is.na(bt3))
  p <- plot_ct(dfx, y_var = "rel_freq", y_label = paste0("Frequency of ",cl," per sample"),
          title = paste0(cl, " frequency"),
          pt.cols = pt.cols)$plot
  pdf(file.path(d,paste0("sample_freq_",cl,".pdf")),width=5,height=7)
  print(p)
  dev.off()
}
```


### 4. Save UMAP Objects and Coordinates

T-cell UMAP (`dt.t.1`/`dt.cd4.1`/`dt.cd8.1`/`dt.t_other.1`) discarded entirely -- no longer
used in the manuscript. Only Tumor and Macrophage cell-type-specific UMAPs are needed.
```{r}
is.non.na <- !is.na(cts7$cell_types)

dim(dt.tum.1)
dim(dt.mac.1)

sum(is.sele.tum)
sum(is.sele.mac)

length(is.sele.tum)
length(is.sele.mac)

sum(table(exp.1$cell_types)[1:2]) # tumor
sum(table(exp.1$cell_types)[7:9]) # Mac cells

save(dt.tum.1,dt.mac.1,
     is.sele.tum,is.sele.mac,
     is.non.na,
     file=file.path(cycif_obj_dir,"cycif7_umap_tumor_and_macrophage.rda"))
```

## ---- Export frnn7.cmb for Notebook 5 ----

`5_cycif_spatial_frnn_tme.Rmd` loads this file and merges `cluster_mac` onto its own
`cluster_tme`-populated `frnn7.cmb`, producing the combined checkpoint `6` needs.
```{r}
saveRDS(frnn7.cmb, file.path(cycif_obj_dir, "cycif7_macrophage_clusters.rds"))
```
