---
title: "TALAVE CyCIF analysis: TME clustering + figure generation (part 2 of 2)"
author: "Kenichi Shimada"
---

# TALAVE CyCIF analysis

Loads the spatial FRNN checkpoint, generates the [EDF2C] cell-type hierarchy and [Fig 2C]
tumor cell density figures, and runs TME-level Leiden clustering with its UMAP and
cluster-frequency outputs.

**Required run order: `3` -> `4` -> `5` (this notebook) -> `6`.**
This notebook's own TME clustering has no dependency on `3` or `4` -- only the final export
step does, where `cluster_mac` (from `4`'s output) is merged onto this notebook's
`cluster_tme`-populated `frnn7.cmb` to produce the combined checkpoint `6` needs.

## ---- Initial Setup ----

### Load Configuration

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Bootstrap

```{r bootstrap}
source(file.path(root_dir, "functions", "bootstrap.R"))
initialize_project()

if (!requireNamespace("CycifAnalyzeR", quietly = TRUE)) {
  remotes::install_github("<org>/CycifAnalyzeR")
}
library(CycifAnalyzeR)
source(file.path(root_dir,"functions","plotting.R"))
source(file.path(source_dir, "plotting_cycif.R"))
source(file.path(source_dir, "analysis_cycif.R"))
source(file.path(source_dir, "io.R"))

for (d in c("cell_type/hierarchy")) {
  dir.create(file.path(cycif_fig_dir, d), recursive = TRUE, showWarnings = FALSE)
}
```

### Load Batch-Corrected Objects

```{r}
cs5 <- readRDS(file.path(cycif_obj_dir, "cs5.rds"))
cs7 <- readRDS(file.path(cycif_obj_dir, "cs7.rds"))
load(file.path(cycif_obj_dir, "cycif_spatial_frnn_input_data.rda")) # frnn7.cmb, frnn7.exp, frnn7.rcn, is.sele.7

pd5 <- pData(cs5)
pd7 <- pData(cs7)
cst7.abs <- colnames(cs7@cell_types$default@cell_state_def)
uniq.cts <- get_cycif_uniq_cts()
pt.cols <- load_pt_colors()

load(file.path(cycif_obj_dir, "cycif5_neighborhood_profiles.rda")) # frnn5.rcn, frnn5.exp
frnn5.tum.n <- rowSums(frnn5.rcn[,1:2])
```

## ---- Cell-Type Hierarchy ----

### [EDF2C] `cs5`/`cs7` Cell-Type Hierarchy

```{r}
cycif_cell_types <- load_cycif_cell_type_definitions()
ctype.5 <- cycif_cell_types$ctype.5
ctype.7 <- cycif_cell_types$ctype.7

pdf(file.path(cycif_fig_dir, "cell_type", "hierarchy", "ab_graph_cyc5.pdf"),width=7,height=7)
par(mar=c(0,0,3,0))
CellTypeGraph(ctype.5,plot=T,main='Cycles 1-5')
dev.off()

pdf(file.path(cycif_fig_dir, "cell_type", "hierarchy", "ab_graph_cyc7.pdf"),width=7,height=7)
par(mar=c(0,0,3,0))
CellTypeGraph(ctype.7,plot=T,main='Cycles 1-7')
dev.off()
```

## ---- Tumor Cell Density Overview ----

### [Fig 2C] Tumor Cell Density by BRCA and Time Point

```{r}
smpls.x5 <- cell_types(cs5) %>% filter(!is.na(cell_types)) %>% pull(sample)
dfx5 <- data.frame(n=frnn5.tum.n,id=smpls.x5) %>%
  filter(n>0) %>%
  group_by(id) %>%
  summarize(n=median(n)) %>%
  left_join(pd5 %>% select(id,Patient.ID,BRCA,TimePoint,BOR),by="id") %>%
  mutate(bt3 = paste0(BRCA,"_",TimePoint)) %>%
  mutate(bt3 = factor(bt3,levels=c("MUT_BS","MUT_BX2","MUT_BX3","WT_BS","WT_BX2","WT_BX3"))) %>%
  group_by(Patient.ID) %>%
  mutate(
    max_bn = max(as.numeric(TimePoint)),
    text = ifelse(as.numeric(TimePoint) == max_bn,as.character(Patient.ID),"")
  ) %>%
  ungroup() %>%
  arrange(bt3)

p5 <- plot_ct(dfx5, y_var = "n", y_label = "Tumor cell density", 
        title = "Tumor cell density around tumor cells",
        pt.cols = pt.cols)

pdf(file.path(cycif_fig_dir, "spatial", "tumor_density", "n_tumor_cells_density_cs5.pdf"),width=5,height=7)
print(p5$plot)
dev.off()

```


## ---- TME UMAP and Cluster Characterization ----

### Build TME UMAP
```{r}
mat <- as.matrix(frnn7.cmb[, 3:13])
n <- rowSums(mat)

if(1){
  set.seed(1)
  mat.jittered <- mat + matrix(rnorm(length(mat), mean = 0, sd = 1e-6), nrow = nrow(mat))
  mat_log <- log1p(mat.jittered)  # log(x + 1) to handle zeros safely
  mat_row_norm <- sweep(mat_log, 1, rowSums(mat_log), "/")
}else{
  mat_freq <- t(apply(mat,1,function(x)x/sum(x)))
  mat.jittered <- mat_freq + matrix(rnorm(length(mat), mean = 0, sd = 1e-6), nrow = nrow(mat))
  mat_row_norm <- mat_log <- log1p(mat.jittered)
}

is.sele.tme <- n>= 10 & is.sele.7

mat.exp <- frnn7.exp
mat.cmb.all <- cbind(mat_row_norm, mat.exp %>% select(-1))
mat.cmb <- mat.cmb.all[n >= 10 & is.sele.7, ]  # Filter rows with n >= 10

```

```{r}
if(1){
  mat.cmb.imputed20 <- knn_impute_dt(mat.cmb, k = 20, weighted = TRUE, fallback = "median")
}else{
  mat.cmb.imputed50 <- knn_impute_dt(mat.cmb, k = 50, weighted = TRUE,
                                     rebuild_per_col = TRUE, fallback = "median")
}
mat.1 <- as.data.frame(mat.cmb.imputed20)
```

#### Compute Embedding
```{r}
set.seed(12345)
k <- 50
umap <- uwot::umap(mat.1,scale=TRUE,n_neighbors = k, verbose=TRUE)

colnames(umap) <- c("umap_x","umap_y")

frnn7.cmb <- frnn7.cmb %>% mutate(.rid = dplyr::row_number())

dt.tme.1 <- frnn7.cmb %>% 
  filter(n>=10 & is.sele.7) %>% 
  cbind(umap) %>%
  left_join(pd7 %>% 
              rename(lspid="sample") %>%
              mutate(batch = case_when(
                substr(lspid, 1, 2) == "13" ~ 1,
                substr(lspid, 1, 2) == "14" ~ 2,
                substr(lspid, 1, 2) == "22" ~ 3,
                TRUE ~ NA_real_
              ),
              batch=factor(as.numeric(batch))) %>%
              select(id,batch,bt3),by=c("sample" = "id"))

idx1 <- rowSums(mat.cmb[,1:2])>0.8
dt.tme.1[names(mat.cmb)] <- mat.cmb.imputed20
dt.tme.1 <- dt.tme.1 %>% mutate(idx=idx1)
```

### [Fig 6B, 6L, EDF9A, EDF9B, EDF9C, EDF9D] TME UMAP Outputs

#### Metadata overlays (tumor fraction, sample, BRCA/timepoint)
```{r}
p1_leg <- umap_base(dt.tme.1, aes(x = umap_x, y = umap_y, col = idx), point_size = 0.8) +
  ggtitle("Tumor freq > 0.8") +   
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "black"),
                     name = "Tumor freq > 0.8")
p1_main <- p1_leg  + theme(legend.position = "none")
ggsave(file.path(cycif_fig_dir, "umap", "tme", "metadata", paste0("umap__tme_tumor_ge80.png")),
   p1_main, width = 7, height = 7, units = "in", dpi = 300)
save_legend_pdf(p1_leg, file.path(cycif_fig_dir, "umap", "tme", "metadata", paste0("umap__tme_tumor_ge80_legend.pdf")),
   width = 4, height = 8)
```

```{r}
p2_full <- umap_base(dt.tme.1, aes(x = umap_x, y = umap_y, col = sample), point_size = 0.8) +
  ggtitle("Samples")
p2 <- p2_full + theme(legend.position = "none")
save_legend_pdf(p2_full, file.path(cycif_fig_dir, "umap", "tme", "metadata", "umap_smpls_legend.pdf"), width = 4, height = 8)  
ggsave(file.path(cycif_fig_dir, "umap", "tme", "metadata", "umap_sample.png"), p2, width = 7, height = 7, units ="in", dpi = 300)

p3_full <- umap_base(dt.tme.1, aes(x = umap_x, y = umap_y, col = bt3), point_size = 0.8) +
  scale_color_brewer(palette = "Set3") +
  ggtitle("BRCA + TimePoint")
p3 <- p3_full + theme(legend.position = "none")
save_legend_pdf(p3_full, file.path(cycif_fig_dir, "umap", "tme", "metadata", "umap_bt3_legend.pdf"), width = 4, height = 6)  
ggsave(file.path(cycif_fig_dir, "umap", "tme", "metadata", "umap_bt3.png"), p3, width = 7, height = 7, units = "in", dpi = 300)

p3_full <- umap_base(dt.tme.1, aes(x = umap_x, y = umap_y, col = TimePoint), point_size = 0.8) +
  scale_color_brewer(palette = "Set1") +
  ggtitle("TimePoint, stratified by BRCA")
p3 <- p3_full + theme(legend.position = "none")
p3 <- p3 + facet_wrap(~BRCA) + theme(strip.text = element_text(size = 36))
save_legend_pdf(p3_full, file.path(cycif_fig_dir, "umap", "tme", "metadata", "umap_bt3_wide_legend.pdf"), width = 4, height = 6)
ggsave(file.path(cycif_fig_dir, "umap", "tme", "metadata", "umap_bt3_wide.png"), p3, width = 14, height = 7, units = "in", dpi = 300)
```

#### Marker expression, per antibody
```{r}
for(ab1 in cst7.abs){
  dtt <- dt.tme.1 %>%
    select(umap_x, umap_y, !!sym(paste0(ab1, ".log"))) %>%
    rename(ab = paste0(ab1, ".log")) %>%
    mutate(ab = pmin(pmax(ab, quantile(ab, 0.05)), quantile(ab, 0.95))) %>%
    mutate(
      ab_raw = ab,
      density = ab_raw
    )
    
  coords <- as.matrix(dtt[, c("umap_x", "umap_y")])
  
  knn <- get.knn(coords, k = 50)
  knn_indices <- cbind(1:nrow(coords), knn$nn.index)
  
  dtt <- dtt %>%
    mutate(density = apply(knn_indices, 1, function(idx){
      mean(dtt$density[idx], na.rm = TRUE)
    })) %>%
    mutate(density = {
        q <- quantile(density, probs = c(0.01, 0.99), na.rm = TRUE)
        pmin(pmax(density, q[1]), q[2])
    })

  p <- ggplot(dtt,aes(x=umap_x,y=umap_y,col=density)) + 
    geom_point(size=0.8) +
    coord_fixed(ratio=1) +
    viridis::scale_color_viridis(guide = "none") +
    theme_minimal() +
    ggtitle(ab1) +
    umap_styling_1()

  if (ab1 == cst7.abs[1]) {
    p_full <- ggplot(dtt, aes(x = umap_x, y = umap_y, col = ab)) +
      geom_point(size = 0.1) +
      coord_fixed(ratio = 1) +
      viridis::scale_color_viridis(name = ab1) +
      theme_minimal() +
      ggtitle(ab1) +
      umap_styling_1()

    save_legend_pdf(p_full,
      file.path(cycif_fig_dir, "umap", "tme", "ab", paste0("legend__", ab1, ".pdf")),
      width = 4, height = 6
    )
  }

  ggsave(file.path(cycif_fig_dir, "umap", "tme", "ab", paste0("umap__",ab1,".png")),p, width = 7, height = 7, units = "in", dpi = 300)  
}
```

#### Cell-type density, per cell type
```{r}
for(ct1 in uniq.cts){
  dtt <- dt.tme.1 %>%
    select(umap_x,umap_y,!!sym(ct1)) %>%
    rename(ct=ct1) %>%
    mutate(
      ct_raw = ct,
      density = ct_raw
    )

  coords <- as.matrix(dtt[, c("umap_x", "umap_y")])
  
  knn <- get.knn(coords, k = 50)
  knn_indices <- cbind(1:nrow(coords), knn$nn.index)
  
  dtt <- dtt %>%
    mutate(density = apply(knn_indices, 1, function(idx){
      mean(dtt$density[idx], na.rm = TRUE)
    })) %>%
    mutate(density = {
        q <- quantile(density, probs = c(0.01, 0.99), na.rm = TRUE)
        pmin(pmax(density, q[1]), q[2])
    })

  p <- ggplot(dtt,aes(x=umap_x,y=umap_y,col=density)) + 
    geom_point(size=0.8) +
    coord_fixed(ratio=1) +
    scale_color_distiller(palette = "YlGnBu", direction = 1, name = "Local Pos %", guide = "none") + 
    theme_minimal() +
    ggtitle(ct1) +
    umap_styling_1() +
    theme(legend.position="none")

  ggsave(file.path(cycif_fig_dir, "umap", "tme", "ct", paste0("umap__ct_",ct1,".png")),p, width = 7, height = 7, units = "in", dpi = 300)  
}

```

#### Marker co-expression heatmap
```{r}
expr_mat <- dt.tme.1 %>%
  select(!!!syms(paste0(cst7.abs, ".log"))) %>%
  select(-GrB.log)

coex_tme <- cor(expr_mat, method = "spearman", use = "pairwise.complete.obs")
coex_tme0 <- coex_tme

mat_sig_04 <- coex_tme0 > 0.4
mat_sig_05 <- coex_tme0 > 0.5

diag(mat_sig_04) <- FALSE
diag(mat_sig_05) <- FALSE

cap <- 1
col_fun <- circlize::colorRamp2(
  c(-cap, 0, cap),
  c("#2166ac", "white", "#b2182b")
)

ht <- Heatmap(
  coex_tme,
  name = "Spearman\nrho",
  col = col_fun,
  na_col = "white",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  rect_gp = gpar(col = NA),
  column_title = "Co-expression of cell state markers (TME)",
  cell_fun = function(j, i, x, y, w, h, fill) {
    
    if (mat_sig_05[i, j]) {
      grid.rect(
        x = x, y = y, width = w, height = h,
        gp = gpar(fill = NA, col = "black", lwd = 1.5)
      )
      
    } else if (mat_sig_04[i, j]) {
      grid.rect(
        x = x, y = y, width = w, height = h,
        gp = gpar(fill = NA, col = "grey40", lwd = 1.2)
      )
    }
  }
)
  
lgd_border <- Legend(
  title = "Correlation",
  labels = c("rho > 0.5", "0.4 < rho <= 0.5"),
  type = "lines",
  legend_gp = gpar(col = c("black", "grey40"), lwd = 2),
  direction = "vertical"
)

pdf(file.path(cycif_fig_dir, "umap", "tme", "expression_by_cluster", "coex_tme_heatmap.pdf"),
    width = 7, height = 7)

ComplexHeatmap::draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  merge_legends = TRUE,
  heatmap_legend_list = list(lgd_border)
)

dev.off()

```

### Characterize TME Clusters


```{r}
k <- 50
mat.cmb_scaled <- scale(mat.1)
knn <- get.knn(mat.cmb_scaled,k = k)

edges <- do.call(rbind, lapply(1:nrow(knn$nn.index), function(i) {
  cbind(i, knn$nn.index[i, ])
}))

g <- graph_from_edgelist(edges, directed = FALSE)
g <- igraph::simplify(g)

leiden_res <- leiden_find_partition(g, resolution_parameter = 2e-3,seed=12345)
cluster_labels <- membership(leiden_res)
lc <- length(table(cluster_labels))
lc
```

```{r}
good_hues <- c(seq(0, 40, length.out = 5),
               seq(80, 160, length.out = 5),
               seq(200, 360, length.out = 7))

custom_colors <- hcl(h = good_hues, c = 100, l = 65)
spectral_colors <- colorRampPalette(brewer.pal(11, "Spectral"))(lc)
set.seed(7)
custom_colors <- sample(spectral_colors)[c(1,2,3,5,6,4,7,10,9,8,11,14,13,12,16,15,17)]
ct.cols17 <- setNames(custom_colors, paste0("TME_", seq_along(custom_colors)))
```

### [Fig 6A, 6C] TME Cluster Composition and Expression

#### Labeled cluster UMAP
```{r}
dt.tme.1 <- dt.tme.1 %>%
  mutate(cluster = factor(paste0("TME_", cluster_labels),levels=paste0("TME_",seq(lc))))

frnn7.cmb$cluster <- NA
frnn7.cmb$cluster[n >= 10 & is.sele.7] <- as.character(dt.tme.1$cluster)
frnn7.cmb <- frnn7.cmb %>%
  mutate(cluster = factor(cluster,levels=paste0("TME_",1:17)))

label_coords <- dt.tme.1 %>%
  group_by(cluster) %>%
  summarise(
    x = median(umap_x, na.rm = TRUE),
    y = median(umap_y, na.rm = TRUE)
  ) %>%
  mutate(text = ifelse(!cluster %in% c("TME_4","TME_11","TME_12","TME_16"),
                       as.numeric(sub("TME_","",cluster)),
                       "")) %>%
  ungroup()  %>%
  arrange(text)

p_full <- umap_base(dt.tme.1, aes(x = umap_x, y = umap_y, col = cluster)) +
  scale_color_manual(values = ct.cols17) +
  umap_styling() +
  geom_text(data = label_coords, aes(x = x, y = y, label = text),
            color = "grey30", size = 15) +
  plot_annotation(
    title = "TME clusters",
    theme = theme(
      plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
      plot.background = element_blank(),
      panel.background = element_blank(),
      panel.border = element_blank()
    )
  ) 

p_main <- p_full + theme(legend.position = "none")

ggsave(file.path(cycif_fig_dir, "umap", "tme", "ct", "umap_cts_cluster-new.png"),
   p_main, width = 7, height = 7, units = "in", dpi = 300)
```

#### UMAP highlighting TME_11/TME_12/TME_16
```{r}
dt.tme.1s <- dt.tme.1 %>%
  mutate(cluster = if_else(cluster %in% c("TME_11", "TME_12", "TME_16"),
                           cluster,
                           "others")) %>%
  mutate(cluster = factor(cluster, levels = c("TME_16","TME_12","TME_11","others"))) %>%
  arrange(desc(cluster))
p_full1 <- umap_base(dt.tme.1, aes(x = umap_x, y = umap_y, col = cluster)) + 
  scale_color_manual(values = ct.cols17) +
  umap_styling() +
  plot_annotation(
    title = "TME clusters",
    theme = theme(
      plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
      plot.background = element_blank(),
      panel.background = element_blank(),
      panel.border = element_blank()
    )
  ) 
p_main1 <- p_full1 + theme(legend.position = "none")
ggsave(file.path(cycif_fig_dir, "umap", "tme", "ct", "umap_cts_cluster-new_nolabel_1.png"),
   p_main1, width = 7, height = 7, units = "in", dpi = 300)  
```

#### Cell-type composition heatmap by cluster
```{r}
dt.ct <- dt.tme.1 %>%
  select(3:13,cluster)

long_counts <- dt.ct %>%
  pivot_longer(-cluster, names_to = "celltype", values_to = "count") %>%
  mutate(celltype = factor(celltype, levels = uniq.cts))

clusters <- levels(long_counts$cluster)
celltypes <- levels(long_counts$celltype)

pvals_df <- long_counts %>%
  group_by(celltype) %>%
  group_modify(~{
    df <- .x
    tibble(
      cluster = clusters,
      pval = vapply(clusters, function(cl) {
        in_vals  <- df$count[df$cluster == cl]
        out_vals <- df$count[df$cluster != cl]
        if (length(in_vals) == 0 || length(out_vals) == 0) return(NA_real_)
        suppressWarnings(
          wilcox.test(in_vals, out_vals, alternative = "greater")$p.value
        )
      }, numeric(1))
    )
  }) %>%
  ungroup()

pvals_df <- pvals_df %>%
  group_by(celltype) %>%
  mutate(padj = p.adjust(pval, method = "BH")) %>%
  ungroup()

p_mat <- pvals_df %>%
  select(cluster, celltype, pval) %>%
  pivot_wider(names_from = cluster, values_from = pval) %>%
  column_to_rownames("celltype") %>%
  as.matrix()
p_mat <- p_mat[uniq.cts,paste0("TME_",seq(length(clusters)))]

sig_labels <- ifelse(p_mat < 1e-100, "*", "")

long_counts <- dt.ct %>%
  pivot_longer(-cluster, names_to = "celltype", values_to = "count")

mat <- long_counts %>%
  group_by(cluster, celltype) %>%
  summarise(count = mean(count, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from  = cluster,
    values_from = count,
    values_fill = list(count = 0)
  ) %>%
  arrange(celltype) %>%
  column_to_rownames("celltype") %>%
  as.matrix()
mat <- mat[uniq.cts,paste0("TME_",seq(length(clusters)))]

mat_rownorm <- t(apply(mat, 1, function(x) {
  if (max(x) == min(x)) {
    rep(0.5, length(x))
  } else {
    (x / max(x))
  }
}))

col_fun <- colorRamp2(
  seq(0, 1, length = 6),
  brewer.pal(6, "YlGnBu")
)

p <- Heatmap(
  mat_rownorm, 
  name = "Rel. freq (row-norm)",
  col = col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  cell_fun = function(j, i, x, y, w, h, fill){
    lab <- sig_labels[i, j]
    if (!is.na(lab) && nzchar(lab)) {
      grid.text(lab, x, y, gp = gpar(fontsize = 12, fontface = "bold"))
    }
  },
  heatmap_legend_param = list(
    title = "Frequency\n(row-norm)",
    legend_direction = "vertical"
  )
) %>%
  ComplexHeatmap::draw(heatmap_legend_side = "right")

pdf(file.path(cycif_fig_dir, "umap", "tme", "expression_by_cluster", "mean_ct_freq_by_cluster_v2.pdf"),width=8,height=5)
print(p)
dev.off()
```

#### Marker expression heatmap by cluster
```{r}
cst7.abs.1 <- rev(c("PD1","PDL1","cCaspase3","gH2AX","pTBK1","BCLXL","MCL1","pAKT","pERK"))

dt.ab <- dt.tme.1 %>% select(15:23,"cluster") 

long_counts <- dt.ab %>%
  pivot_longer(-cluster, names_to = "protein", values_to = "exp")

clusters <- levels(dt.ab$cluster)

pvals_df <- long_counts %>%
  group_by(protein) %>%
  group_modify(~{
    df <- .x
    tibble(
      cluster = clusters,
      pval = vapply(clusters, function(cl) {
        in_vals  <- df$exp[df$cluster == cl]
        out_vals <- df$exp[df$cluster != cl]
        if (length(in_vals) == 0 || length(out_vals) == 0) return(NA_real_)
        suppressWarnings(
          wilcox.test(in_vals, out_vals, alternative = "greater")$p.value
        )
      }, numeric(1))
    )
  }) %>%
  ungroup()

pvals_df <- pvals_df %>%
  group_by(protein) %>%
  mutate(padj = p.adjust(pval, method = "BH")) %>%
  ungroup()

p_mat <- pvals_df %>%
  select(cluster, protein, pval) %>%
  pivot_wider(names_from = cluster, values_from = pval) %>%
  column_to_rownames("protein") %>%
  as.matrix()

p_mat <- p_mat[cst7.abs.1,paste0("TME_",seq(length(clusters)))]

sig_labels <- ifelse(p_mat < 1e-75, "*", "")

mat <- long_counts %>%
  group_by(cluster, protein) %>%
  summarise(exp = mean(exp, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from  = cluster,
    values_from = exp,
    values_fill = list(exp = 0)
  ) %>%
  column_to_rownames("protein") %>%
  as.matrix()
mat <- mat[cst7.abs.1,]

mat_rownorm <- t(apply(mat, 1, function(x) {
  tmp <- scale(x,center=TRUE,scale=TRUE)
  tmp <- pmin(pmax(tmp, -3), 3)
  return(tmp)
}))
colnames(mat_rownorm) <- colnames(mat)

col_fun <- colorRamp2(
  seq(-3, 3, length = 9),
  rev(brewer.pal(11, "RdBu")[2:10])
)

p <- Heatmap(
  mat_rownorm,
  name = "Rel. freq (row-norm)",
  col = col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  cell_fun = function(j, i, x, y, w, h, fill){
    lab <- sig_labels[i, j]
    if (!is.na(lab) && nzchar(lab)) {
      grid.text(lab, x, y, gp = gpar(fontsize = 12, fontface = "bold"))
    }
  },
  heatmap_legend_param = list(
    title = "Expression\n(row-scaled)",
    legend_direction = "vertical"
  )
) %>%
  ComplexHeatmap::draw(heatmap_legend_side = "right")

pdf(file.path(cycif_fig_dir, "umap", "tme", "expression_by_cluster", "mean_ab_freq_by_cluster_v2.pdf"),width=8,height=5)
print(p)
dev.off()
```

### [Fig 6D, 6E, 6H, EDF9F] TME Cluster Frequency by BRCA and Time Point

Computes each sample's relative frequency of cells falling into each of the 17 TME clusters,
then plots each cluster's frequency across BRCA x TimePoint groups via `plot_ct()` (same LMER
framework used throughout this pipeline).
```{r}
dt.tme.means <- dt.tme.1 %>%
  mutate(sample = factor(sample)) %>%
  select(cluster,sample) %>%
  group_by(cluster,sample) %>%
  summarise(n=n(),.groups = "drop") %>%
  complete(cluster, sample, fill = list(n = 0)) %>%
  left_join(pd7 %>% select(id,TimePoint,BRCA,bt3,BOR,Patient.ID),
            by=c("sample"="id")) %>%
  group_by(sample) %>%
  mutate(rel_freq = n / sum(n)) %>%
  ungroup()

if(!dir.exists(file.path(cycif_fig_dir, "umap", "tme", "cluster_freq"))){
  dir.create(file.path(cycif_fig_dir, "umap", "tme", "cluster_freq"),recursive=TRUE)
}

uniq.tmes <- paste0("TME_",1:17)
for(ut in uniq.tmes){
  tmp <- dt.tme.means %>% filter(cluster == ut)
  p <- plot_ct(tmp, y_var = "rel_freq", y_label = "Relative frequency",
    title = paste0(ut, " frequency"),
    pt.cols = pt.cols)$plot

  ggsave(file.path(cycif_fig_dir, "umap", "tme", "cluster_freq", paste0("freq_",ut,".pdf")),p,width=5,height=7,units="in",dpi=300)
}
```

```{r}
frnn7.cmb$cluster_tme <- NA
frnn7.cmb$cluster_tme[is.sele.tme] <- as.character(dt.tme.1$cluster)
```

### Propagate cluster_tme to the Full Cell Population

`cluster_tme` above is only assigned for the `is.sele.tme` subsample (~1-3% of cells per
cell type) that the TME-level Leiden clustering was actually run on -- Leiden has no direct
"predict for new points" the way a centroid-based method does, since it operates on a
neighbor graph, not a parametric model. Extend it to the remaining ~97-99% of cells via k-NN
majority-vote label propagation: for each out-of-sample cell, find its nearest neighbors
(in the same RCN-composition + expression feature space Leiden was run on -- `mat.cmb.all`,
scaled using the `is.sele.tme` reference population's own center/scale, matching
`mat.cmb_scaled`) among the `is.sele.tme`-labeled cells, and take the majority `cluster_labels`
vote among those neighbors.
```{r}
ref_scaled <- mat.cmb_scaled  # is.sele.tme population, already scaled -- Leiden's own input
ref_cols <- colnames(ref_scaled)

cat("ncol(mat.cmb.all):", ncol(mat.cmb.all),
    "| ncol(ref_scaled):", length(ref_cols),
    "| duplicated colnames in mat.cmb.all:", sum(duplicated(colnames(mat.cmb.all))),
    "| all ref_cols in mat.cmb.all:", all(ref_cols %in% colnames(mat.cmb.all)), "\n")

query_idx <- which(!is.sele.tme & stats::complete.cases(mat.cmb.all))
query_mat <- as.matrix(mat.cmb.all)[query_idx, ref_cols, drop = FALSE]
storage.mode(query_mat) <- "numeric"
stopifnot(ncol(query_mat) == length(ref_cols))

query_scaled <- scale(
  query_mat,
  center = attr(ref_scaled, "scaled:center")[ref_cols],
  scale  = attr(ref_scaled, "scaled:scale")[ref_cols]
)

k_prop <- 20
nn_prop <- FNN::get.knnx(data = ref_scaled, query = query_scaled, k = k_prop)

propagated_cluster <- apply(nn_prop$nn.index, 1, function(idx) {
  votes <- cluster_labels[idx]
  names(sort(table(votes), decreasing = TRUE))[1]
})

frnn7.cmb$cluster_tme[query_idx] <- paste0("TME_", propagated_cluster)

cat("cluster_tme coverage after propagation:",
    sum(!is.na(frnn7.cmb$cluster_tme)), "/", nrow(frnn7.cmb),
    sprintf("(%.1f%%)\n", 100 * mean(!is.na(frnn7.cmb$cluster_tme))))
```

## ---- Export for Notebook 6 ----

`6_cycif_spatial_continued.Rmd` gets its shared setup objects (`cs7` from `cs7.rds`, `pd7`,
etc.) directly from the same checkpoint files this notebook loads, so only what's genuinely
new here -- `dt.tme.1`, `frnn7.cmb` (with `cluster_tme` now assigned), `ct.cols17`, and
`is.sele.tme` -- needs to be exported. `frnn7.cmb` is also merged here with `cluster_mac`
from `4`'s `cycif7_macrophage_clusters.rds` (same base checkpoint, so row order matches) to
produce the combined checkpoint `6` actually loads.
```{r}
saveRDS(dt.tme.1, file.path(cycif_obj_dir, "cycif7_tme_cluster_assignments.rds"))
saveRDS(frnn7.cmb, file.path(cycif_obj_dir, "cycif7_tme_clusters.rds"))

save(ct.cols17, is.sele.tme,
     file = file.path(cycif_obj_dir, "cycif7_tme_colors_and_selection.rda"))

frnn7.cmb$cluster_mac <- readRDS(file.path(cycif_obj_dir, "cycif7_macrophage_clusters.rds"))$cluster_mac
saveRDS(frnn7.cmb, file.path(cycif_obj_dir, "cycif7_tme_and_macrophage_clusters.rds"))
```
