---
title: "TALAVE CyCIF analysis: spatial FRNN + TME clustering (part 2 of 2)"
author: "Kenichi Shimada"
---

# TALAVE CyCIF analysis

Continues the spatial pipeline: adjacent-cell and tumor-border distances, tumor density,
macrophage and T-cell spatial programs, TS ratios, and marker-density regressions.

**Required run order: `3` -> `4` -> `5` -> `6` (this notebook).**
(This notebook resumes directly from `5`'s output.)

## ---- Initial Setup ----

### Load Configuration
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Bootstrap
```{r bootstrap}
source(file.path(root_dir, "functions", "bootstrap.R"))
initialize_project()

if (!requireNamespace("CycifAnalyzeR", quietly = TRUE)) {
  remotes::install_github("<org>/CycifAnalyzeR")
}
library(CycifAnalyzeR)
source(file.path(root_dir,"functions","plotting.R"))
source(file.path(source_dir, "plotting_cycif.R"))
source(file.path(source_dir, "analysis_cycif.R"))
source(file.path(source_dir, "io.R"))
```

### CyCIF-Specific Colors
```{r}
brca.fills = c(MUT="#C9C0DD", WT="#D9DADA")
brca.cols = c(MUT="#62439c", WT="#58575f")
brca.mids = c(MUT="#62439c", WT = "grey60")
```

## ---- Load Inputs from 5 ----

Resumes the spatial FRNN + TME-clustering pipeline from `5_cycif_spatial_frnn_tme.Rmd`, which
merges in the macrophage clustering from `4_cycif_umap_analysis.Rmd` before saving the
combined checkpoint this notebook loads.

```{r}
cs7 <- readRDS(file.path(cycif_obj_dir, "cs7.rds"))
load(file.path(cycif_obj_dir, "cycif_spatial_frnn_input_data.rda")) # frnn7.cmb, frnn7.exp, frnn7.rcn, is.sele.7

pd7 <- pData(cs7)
cts7d <- cell_types(cs7) %>% filter(!is.na(cell_types))
cst7.abs <- colnames(cs7@cell_types$default@cell_state_def)
ct.cols7 <- get_cycif_ct_colors()
ct.cols7w <- get_cycif_ct_colors_white_blend()
uniq.cts <- get_cycif_uniq_cts()
nms7 <- names(cs7)
pt.cols <- load_pt_colors()

lst.dists7 <- readRDS(file.path(cycif_obj_dir, "cycif7_tumor_border_distances_by_sample.rds"))
lst.frnn7 <- readRDS(file.path(cycif_obj_dir, "cycif7_neighborhood_composition_by_sample.rds"))

load(file.path(cycif_obj_dir, "cycif7_tme_colors_and_selection.rda")) # ct.cols17, is.sele.tme
dt.tme.1 <- readRDS(file.path(cycif_obj_dir, "cycif7_tme_cluster_assignments.rds"))

frnn7.cmb <- readRDS(file.path(cycif_obj_dir, "cycif7_tme_and_macrophage_clusters.rds")) # cluster_tme populated for all cells (propagated); cluster_mac only for the down-sampled is.sele.mac subset, NA elsewhere
```

### [Fig 6F, 6G, 6K, EDF10C, EDF10D, EDF10E, EDF10F] TME Cluster Associations and Marker States
```{r}
frnn7.cmb <- frnn7.cmb %>%
  mutate(cluster_tme = factor(cluster_tme,levels=paste0("TME_",1:17))) %>%
  mutate(cluster_mac = factor(cluster_mac,levels=paste0("Mac_",1:10)))
```

#### Macrophage neighbor-count per TME cluster

For each TME cluster, counts macrophages (`Mac_CD68`/`Mac_CD68_CD163`/`Mac_CD163`) among the
spatial neighbors of every cell assigned to that cluster, then labels each macrophage with
the (at most one) TME cluster it borders.
```{r}
mac_types <- c("Mac_CD68","Mac_CD68_CD163","Mac_CD163")

tme_targets <- paste0("TME_",1:17)
newcols <- paste0("mac_neighbor_count_", tme_targets)

for (tme in tme_targets) {
  message("Processing ", tme, "...")

  newcol <- paste0("mac_neighbor_count_", tme)
  frnn7.cmb[[newcol]] <- NA_real_

  for (sm in names(lst.frnn7)) {
    obj <- lst.frnn7[[sm]]
    neigh_list <- obj@nn@id
    cts_sample <- cts7d %>% filter(sample == sm)

    idx_cmb <- which(frnn7.cmb$sample == sm)
    cmb_sub <- frnn7.cmb[idx_cmb, ]

    idx_tme_local <- which(!is.na(cmb_sub$cluster_tme) &
                             cmb_sub$cluster_tme == tme)
    if (length(idx_tme_local) == 0) next

    neigh_all <- unlist(neigh_list[idx_tme_local])

    mac_local_ids <- which(cts_sample$cell_types %in% mac_types)

    tab_all <- table(neigh_all)

    mac_count_local <- rep(NA_real_, nrow(cmb_sub))

    mac_count_local[mac_local_ids] <- tab_all[as.character(mac_local_ids)]

    mac_count_local[mac_local_ids][is.na(mac_count_local[mac_local_ids])] <- 0

    frnn7.cmb[[newcol]][idx_cmb[idx_tme_local]] <- 
      mac_count_local[idx_tme_local]
  }
}
```

```{r}
macs.nb <- frnn7.cmb %>% 
  dplyr::select(cell_types,!!!syms(newcols))
macs.nb$shared_tme_n <- macs.nb %>%
  dplyr::select(all_of(newcols)) %>%
  mutate(across(everything(), ~ .x > 0)) %>%
  rowSums(na.rm = TRUE)
table(macs.nb$shared_tme_n)

frnn7.cmb <- frnn7.cmb %>%
  mutate(tme_neighbor_label = apply(
      macs.nb[newcols] > 0, 1, 
      function(x) {
        if (!any(x, na.rm = TRUE)) return(NA_character_)
        tme_targets[which(x)]
      }
    ),
    tme_neighbor_label = factor(tme_neighbor_label, levels = tme_targets)
  )
```

#### CD163+ macrophage frequency: TME_5 vs TME_12 vs other clusters

Paired Wilcoxon tests, per sample, comparing the mean fraction of neighborhood macrophages
(CD68+, CD163+, or CD68+/CD163+ double-positive) that are the CD163+ subtype, between TME_5,
TME_12, and all other clusters.
```{r}
n.cd163 <- as.data.frame(frnn7.rcn[is.sele.tme, ]) %>%
  mutate(
    total_mac  = Mac_CD68 + Mac_CD163 + Mac_CD68_CD163,
    frac_CD163 = ifelse(total_mac > 0, Mac_CD163 / total_mac, NA_real_)
  ) %>%
  bind_cols(
    frnn7.cmb %>%
      filter(is.sele.tme) %>%
      dplyr::select(sample, cluster_tme)
  ) %>%
  mutate(
    tme_group = case_when(
      cluster_tme == "TME_5"  ~ "TME_5",
      cluster_tme == "TME_12" ~ "TME_12",
      TRUE                    ~ "others"
    ),
    tme_group = factor(tme_group, levels = c("TME_5", "TME_12", "others"))
  ) %>%
  dplyr::select(tme_group, sample, frac_CD163) %>%
  group_by(tme_group, sample) %>%
  summarise(mean_frac_CD163 = mean(frac_CD163, na.rm = TRUE), .groups = "drop")

df.paired <- n.cd163 %>%
  pivot_wider(
    names_from = tme_group,
    values_from = mean_frac_CD163
  ) %>%
  filter(
    !is.na(TME_5),
    !is.na(TME_12),
    !is.na(others)
  )

p_5_12   <- wilcox.test(df.paired$TME_5, df.paired$TME_12, paired = TRUE)$p.value
p_12_oth <- wilcox.test(df.paired$TME_12, df.paired$others, paired = TRUE)$p.value
p_5_oth  <- wilcox.test(df.paired$TME_5, df.paired$others, paired = TRUE)$p.value

p_adj <- p.adjust(c(p_5_12, p_12_oth, p_5_oth), method = "BH")

df_plot <- df.paired %>%
  dplyr::select(sample, TME_5, TME_12, others) %>%
  pivot_longer(
    cols = c(TME_5, TME_12, others),
    names_to = "region",
    values_to = "value"
  ) %>%
  mutate(
    region = factor(region, levels = c("TME_5","TME_12","others")),
    x = case_when(
      region == "TME_5" ~ 1,
      region == "TME_12" ~ 2,
      region == "others" ~ 3
    )
  ) %>%
  arrange(sample, x)


y_base <- max(df_plot$value, na.rm = TRUE)

ann_df <- tibble(
  x = c(1.1, 2.1, 1.1),
  xend = c(1.9, 2.9, 2.9),
  xlab = c(1.5, 2.5, 2.0),
  y = c(y_base*1.06, y_base*1.06, y_base*1.15),
  label = signif(p_adj, 2)
)

p <- ggplot(df_plot, aes(x = x, y = value, group = sample)) +
  geom_path(color="grey70", linewidth=0.4) +
  geom_point(aes(color=region), size=2) +
  stat_summary(
    aes(group = region),
    fun = median,
    geom = "crossbar",
    width = 0.35,
    color = "black",
    fatten = 1,
    linewidth = 0.5
  ) +

  geom_segment(
    data = ann_df,
    inherit.aes = FALSE,
    aes(x=x, xend=xend, y=y, yend=y),
    linewidth=0.6
  ) +

  geom_text(
    data = ann_df,
    inherit.aes = FALSE,
    aes(x=xlab, y=y, label=label),
    vjust=-0.25,
    size=4
  ) +

  scale_x_continuous(
    breaks=c(1,2,3),
    labels=c("TME_5","TME_12","others"),
    expand = expansion(mult = c(0.1, 0.1))    
  ) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0.1, 0.1))   
  ) +

  scale_color_manual(
    values=c(
      ct.cols17[c(5,12)],
      "others"="grey40"
    )
  ) +

  labs(
    title="CD163+ macs",
    x=NULL,
    y="Mean fraction of macrophages\nthat are CD163+ (per sample)"
  ) +

  theme_bw(base_size=14) +
  theme(
    legend.position="none",
    axis.text.x = element_text(angle=90, vjust=0.5, hjust=1)
  )

ggsave(
  file.path(cycif_fig_dir, "umap", "tme", "expression_by_cluster", "mean_cd163_frac_by_tme.pdf"),
  p, width = 2.5, height = 4, units = "in", dpi = 300
)

```

#### Neighboring T cells per CD163+ macrophage: TME_5 vs TME_12 vs other clusters

Same TME_5/TME_12/others grouping and paired-Wilcoxon approach as above, now applied per
T-cell subtype (CD8T/CD4T/T_other) to the count of that T-cell type among each CD163+
macrophage's spatial neighbors.
```{r}
n.t.per.mac <- frnn7.cmb %>%
  filter(
    cell_types %in% "Mac_CD163",
    !is.na(tme_neighbor_label)
  ) %>%
  mutate(
    is_tme5_12 = case_when(
      tme_neighbor_label == "TME_5"  ~ "TME_5",
      tme_neighbor_label == "TME_12" ~ "TME_12",
      TRUE                           ~ "others"
    ),
    is_tme5_12 = factor(is_tme5_12, levels = c("TME_5","TME_12","others"))
  ) %>%
  dplyr::select(cell_types, is_tme5_12, CD4T, CD8T, T_other, sample) %>%
  pivot_longer(
    cols = c("CD4T","CD8T","T_other"),
    names_to = "T_cell_type",
    values_to = "n_T_cells"
  ) %>%
  mutate(T_cell_type = factor(T_cell_type, levels = c("CD8T","CD4T","T_other")))

df.sample <- n.t.per.mac %>%
  group_by(sample, T_cell_type, is_tme5_12) %>%
  summarize(
    mean_T = mean(n_T_cells, na.rm = TRUE),
    n_mac  = n(),
    .groups = "drop"
  ) %>%
  mutate(T_cell_type = factor(T_cell_type, levels = c("CD8T","CD4T","T_other")))

df.sample %>%
  count(T_cell_type, is_tme5_12)

df.paired <- df.sample %>% 
  pivot_wider(
    names_from = is_tme5_12,
    values_from = c(mean_T, n_mac),
    names_sep = "_"
  ) %>%
  filter(
    !is.na(mean_T_TME_5),
    !is.na(mean_T_TME_12),
    !is.na(mean_T_others)
  ) %>%
  mutate(T_cell_type = factor(T_cell_type, levels = c("CD8T","CD4T","T_other")))


df_plot <- df.paired %>%
  dplyr::select(sample, T_cell_type, mean_T_TME_5, mean_T_TME_12, mean_T_others) %>%
  pivot_longer(
    cols = c(mean_T_TME_5, mean_T_TME_12, mean_T_others),
    names_to = "region",
    values_to = "value"
  ) %>%
  mutate(
    region = factor(
      region,
      levels = c("mean_T_TME_5", "mean_T_TME_12", "mean_T_others"),
      labels = c("TME_5", "TME_12", "others")
    ),
    x = case_when(
      region == "TME_5"  ~ 1,
      region == "TME_12" ~ 2,
      region == "others" ~ 3
    )
  ) %>%
  arrange(T_cell_type, sample, x)

stats <- df.paired %>%
  group_by(T_cell_type) %>%
  summarize(
    p_5_12   = wilcox.test(mean_T_TME_5, mean_T_TME_12, paired = TRUE)$p.value,
    p_12_oth = wilcox.test(mean_T_TME_12, mean_T_others, paired = TRUE)$p.value,
    p_5_oth  = wilcox.test(mean_T_TME_5, mean_T_others, paired = TRUE)$p.value,
    y_base   = max(c(mean_T_TME_5, mean_T_TME_12, mean_T_others), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    p_adj_5_12   = p.adjust(p_5_12, method = "BH"),
    p_adj_12_oth = p.adjust(p_12_oth, method = "BH"),
    p_adj_5_oth  = p.adjust(p_5_oth, method = "BH")
  )

y_ref <- stats %>%
  filter(T_cell_type == "CD4T") %>%
  pull(y_base)

if (length(y_ref) == 0) {
  y_ref <- max(df_plot$value, na.rm = TRUE)
}

y_12 <- y_ref * 1.12
y_23 <- y_ref * 1.12
y_13 <- y_ref * 1.28

seg_pad <- 0.08

ann_df <- bind_rows(
  stats %>%
    transmute(
      T_cell_type,
      x = 1 + seg_pad, xend = 2 - seg_pad,
      xlab = 1.5,
      y = y_12,
      p_adj = p_adj_5_12
    ),
  stats %>%
    transmute(
      T_cell_type,
      x = 2 + seg_pad, xend = 3 - seg_pad,
      xlab = 2.5,
      y = y_23,
      p_adj = p_adj_12_oth
    ),
  stats %>%
    transmute(
      T_cell_type,
      x = 1 + seg_pad, xend = 3 - seg_pad,
      xlab = 2.0,
      y = y_13,
      p_adj = p_adj_5_oth
    )
) %>%
  mutate(
    label = signif(p_adj, 2),
    T_cell_type = factor(T_cell_type, levels = c("CD8T", "CD4T", "T_other"))
  )

y_top <- max(ann_df$y) * 1.06

p <- ggplot(df_plot, aes(x = x, y = value, group = sample)) +
  geom_path(color = "grey70", linewidth = 0.5) +
  geom_point(aes(color = region), size = 2) +
  stat_summary(
    aes(group = region),
    fun = median,
    geom = "crossbar",
    width = 0.35,
    color = "black",
    fatten = 1,
    linewidth = 0.5
  ) +
  facet_wrap(~T_cell_type) +
  geom_segment(
    data = ann_df,
    inherit.aes = FALSE,
    aes(x = x, xend = xend, y = y, yend = y),
    linewidth = 0.5,
    color = "black"
  ) +
  geom_text(
    data = ann_df,
    inherit.aes = FALSE,
    aes(x = xlab, y = y, label = label),
    vjust = -0.25,
    size = 3.8
  ) +
  scale_x_continuous(
    breaks = c(1, 2, 3),
    labels = c("TME_5", "TME_12", "others"),
    expand = expansion(mult = c(0.15, 0.15))
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.02)),
    limits = c(NA, y_top)
  ) +
  scale_color_manual(
    values=c(
      ct.cols17[c(5,12)],
      "others"="grey40"
    )
  ) +
  labs(
    x = NULL,
    y = "Mean neighboring T cells per CD163+ macs\n(per sample)"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(size = 12, angle = 90, vjust = 0.5, hjust = 1),
    axis.text.y = element_text(size = 11),
    axis.title.y = element_text(size = 12)
  )

ggsave(
  file.path(cycif_fig_dir, "umap", "tme", "expression_by_cluster", "t_cells_per_mac_CD163_by_tme12_boxplot_v2.pdf"),
  p, width = 4, height = 4, units = "in", dpi = 300
)

```

#### Marker expression heatmaps by TME cluster

`frnn7.cmb` alone serves as both `frnn_meta` and `frnn_exp` for `plot_marker_tme_heatmap()`;
a per-sample sequential `cell_id` is added as the join key it expects.
```{r}
frnn7.cmb.id <- frnn7.cmb %>% group_by(sample) %>% mutate(cell_id = dplyr::row_number()) %>% ungroup()

markers_use <- c("pTBK1", "PDL1", "gH2AX","cCaspase3")
tme_levels <- tme_targets
res_list <- lapply(markers_use, function(mk) {
  x <- plot_marker_tme_heatmap(
    marker = mk,
    tme_levels = tme_levels,
    frnn_meta = frnn7.cmb.id,
    frnn_exp = frnn7.cmb.id,
    cluster_columns = FALSE,
    height = 6,
    width = 4.5,
    file = file.path(cycif_fig_dir, "umap", "tme", "expression_by_cluster", paste0("heatmap_", mk, "_by_TME.pdf"))
  )
})

names(res_list) <- markers_use
```


### [Fig 6M, EDF10G] TME Marker Scatterplots

#### TME-cluster–colored scatterplots
```{r}
ab_pairs <- list(
  c("PDL1","cCaspase3"),
  c("gH2AX", "pTBK1")
)

scatters <- list()

for (pair in ab_pairs) {
  ab1 <- pair[1]
  ab2 <- pair[2]

  xlab_text <- if (ab1 == "gH2AX") {
    expression(gamma * "H2AX (CN)")
  } else {
    paste0(ab1, " (CN)")
  }

  dt.tme.1s <- dt.tme.1 %>%
    mutate(cluster = if_else(cluster %in% c("TME_5","TME_11", "TME_12", "TME_16"),
                             cluster,
                             "others")) %>%
    mutate(cluster = factor(cluster, levels = c("TME_16","TME_5","TME_12","TME_11","others"))) %>%
    arrange(desc(cluster)) %>%
    filter(gH2AX > -1 | pTBK1 > -1)

  set.seed(123)

  important <- c("TME_5","TME_11","TME_12","TME_16")

  dt.tme.1s <- dt.tme.1s %>%
    mutate(is_imp = cluster %in% important) %>%
    arrange(desc(is_imp)) %>%
    mutate(rand = runif(n())) %>%
    arrange(is_imp, rand) %>%
    dplyr::select(-is_imp, -rand)

  p_main <- ggplot(dt.tme.1s , aes(x = .data[[ab1]], y = .data[[ab2]], col = cluster)) +
    geom_point() +
    scale_color_manual(values = c(ct.cols17,others="grey70")) +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 28,color="black"),
      axis.title = element_text(size = 32,color="black"),
      legend.title = element_text(size = 20, margin = margin(b = 10)),
      legend.text = element_text(size = 16),
      legend.key.size = unit(1.2, "cm"),
      legend.spacing.y = unit(1, "cm"),
      legend.key.height = unit(1.2, "cm"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(size=1,color="black")
    ) +
    xlab(xlab_text) + 
    ylab(paste0(ab2," (CN)")) +
    plot_annotation(
      title = paste0(ab1," vs ",sub("ase","",ab2)),
      theme = theme(
        plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
        plot.background = element_blank()
      )
    ) 
  ggsave(file.path(cycif_fig_dir, "spatial", "coexpression", "tme", "cluster", paste0("scatter_",ab1,"_",ab2,"_clust_v2.png")),
     p_main + theme(legend.position="none"), width = 7, height = 7, units = "in", dpi = 300)
  
}

save_legend_pdf(p_main,
              file.path(cycif_fig_dir, "spatial", "coexpression", "tme", "cluster",
          paste0("scatter_cluster_legend.pdf")),
  width = 4, height = 6
)
```

#### Neighborhood-density–colored scatterplots (by cell type / antibody)
```{r}
ab_pairs <- list(
  c("PDL1","cCaspase3"),
  c("gH2AX", "pTBK1")
)

scatters <- list()

for (ct in uniq.cts[4:6]) {
  scatters[[ct]] <- list()
  
  for (pair in ab_pairs) {
    ab1 <- pair[1]
    ab2 <- pair[2]
    
    xlab_text <- if (ab1 == "gH2AX") {
      expression(gamma * "H2AX (CN)")
    } else {
      paste0(ab1, " (CN)")
    }
        
    dt.tme.1s <- dt.tme.1 %>%
      mutate(ct_raw = !!sym(ct)) %>%
      mutate(
        density = {
          ct_vals <- ct_raw
          q <- quantile(ct_vals, probs = c(0.01, 0.99), na.rm = TRUE)
          pmin(pmax(ct_vals, q[1]), q[2])
        }
      ) %>%
      filter(gH2AX > -1 | pTBK1 > -1)
    
    coords <- as.matrix(dt.tme.1s[c(ab1, ab2)])
    knn <- get.knn(coords, k = 20)
    knn_indices <- cbind(1:nrow(coords), knn$nn.index)
    
    dt.tme.1s$density <- apply(knn_indices, 1, function(idx) {
      mean(dt.tme.1s$density[idx], na.rm = TRUE)
    })
    
    p_main <- ggplot(dt.tme.1s , aes(x = .data[[ab1]], y = .data[[ab2]], col = density)) +
      geom_point() +
      scale_color_distiller(palette = "YlGnBu", direction = 1, type = "seq", name = "mean\ndensity") +
      theme_minimal() +
      theme(
        axis.text = element_text(size = 28,color="black"),
        axis.title = element_text(size = 32,color="black"),
        legend.title = element_text(size = 20, margin = margin(b = 10)),
        legend.text = element_text(size = 16),
        legend.key.size = unit(1.2, "cm"),
        legend.spacing.y = unit(1, "cm"),
        legend.key.height = unit(1.2, "cm"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(size=1,color="black")
      ) +
      xlab(xlab_text) +
      ylab(paste0(ab2," (CN)")) +
      plot_annotation(
        title = ct,
        theme = theme(
          plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
          plot.background = element_blank()
        )
      ) + common_scale(0.42)
    
    out_path <- file.path(cycif_fig_dir, "spatial", "coexpression", "tme", "ct",
                          paste0("scatter_", ab1, "_", ab2, "_", ct, ".png"))
    ggsave(out_path, p_main + theme(legend.position="none"), width = 7, height = 7, units = "in", dpi = 300)
    
    scatters[[ct]][[paste0(ab1, "_", ab2)]] <- p_main
  }
}

save_legend_pdf(p_main,
                file.path(cycif_fig_dir, "spatial", "coexpression", "tme", "ct",
            paste0("scatter_celltype_legend.pdf")),
  width = 4, height = 6
)

for (ab in cst7.abs[c(2,3,4,9,10)]){
  scatters[[ab]] <- list()
  
  for (pair in ab_pairs) {
    ab1 <- pair[1]
    ab2 <- pair[2]
  
    xlab_text <- if (ab1 == "gH2AX") {
      expression(gamma * "H2AX (CN)")
    } else {
      paste0(ab1, " (CN)")
    }

    dt.tme.1s <- dt.tme.1 %>%
      mutate(ab_raw = .data[[ab]]) %>%
      mutate(
        density = {
          ab_vals <- ab_raw
          q <- quantile(ab_vals, probs = c(0.01, 0.99), na.rm = TRUE)
          pmin(pmax(ab_vals, q[1]), q[2])
        }
      ) %>%
      filter(gH2AX > -1 | pTBK1 > -1)      
    
    coords <- as.matrix(dt.tme.1s[c(ab1, ab2)])
    knn <- get.knn(coords, k = 20)
    knn_indices <- cbind(1:nrow(coords), knn$nn.index)
    
    dt.tme.1s$density <- apply(knn_indices, 1, function(idx) {
      mean(dt.tme.1s$density[idx], na.rm = TRUE)
    })
    
    p_main <- ggplot(dt.tme.1s , aes(x = .data[[ab1]], y = .data[[ab2]], col = density)) +
      geom_point() +
      viridis::scale_color_viridis(guide = "none") +
      theme_minimal() +
      theme(
        axis.text = element_text(size = 28,color="black"),
        axis.title = element_text(size = 32,color="black"),
        legend.title = element_text(size = 20, margin = margin(b = 10)),
        legend.text = element_text(size = 16),
        legend.key.size = unit(1.2, "cm"),
        legend.spacing.y = unit(1, "cm"),
        legend.key.height = unit(1.2, "cm"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(size=1,color="black")
      ) +
      xlab(xlab_text) +
      ylab(paste0(ab2," (CN)")) +
      plot_annotation(
        title = ab,
        theme = theme(
          plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
          plot.background = element_blank()
        )
      )
    
    out_path <- file.path(cycif_fig_dir, "spatial", "coexpression", "tme", "ab",
                          paste0("scatter_", ab1, "_", ab2, "_", ab, ".png"))
    ggsave(out_path, p_main, width = 7, height = 7, units = "in", dpi = 300)
    
    scatters[[ab]][[paste0(ab1, "_", ab2)]] <- p_main
  }
}

p_main <- ggplot(dt.tme.1s , aes(x = !!sym(ab1), y = !!sym(ab2), col = density)) +
  geom_point() +
  viridis::scale_color_viridis() +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 20),
    axis.title = element_text(size = 24),
    legend.title = element_text(size = 20, margin = margin(b = 10)),
    legend.text = element_text(size = 16),
    legend.key.size = unit(1.2, "cm"),
    legend.spacing.y = unit(1, "cm"),
    legend.key.height = unit(1.2, "cm"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(size=1)
  ) +
  xlab(paste0(ab1," (log)")) + 
  ylab(paste0(ab2," (log)")) +
  plot_annotation(
    title = ab,
    theme = theme(
      plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
      plot.background = element_blank()
    )
  )
save_legend_pdf(p_main,
                file.path(cycif_fig_dir, "spatial", "coexpression", "tme", "ab",
            paste0("scatter_ab_legend.pdf")),
  width = 4, height = 6
)
```
### [Fig 6K, EDF10B] TME Marker Violin Plots
```{r}
for(ab1 in cst7.abs[c(2,3,4,9,10)]){
  dt.tme.1s <- dt.tme.1 %>%
    mutate(ab_raw = .data[[ab1]]) %>%
    mutate(
      density = {
        ab_vals <- ab_raw
        q <- quantile(ab_vals, probs = c(0.01, 0.99), na.rm = TRUE)
        pmin(pmax(ab_vals, q[1]), q[2])
      }
    ) %>%
    mutate(cluster = if_else(cluster %in% c("TME_5","TME_11", "TME_12", "TME_16"),
                             cluster,  # keep as is
                             "others")) %>%    
    filter(gH2AX > -1 | pTBK1 > -1)    
  
  others_med <- median(
    dt.tme.1s %>% filter(cluster == "others") %>% pull(.data[[ab1]]),
    na.rm = TRUE
  )
  
  
  p <- ggplot(dt.tme.1s %>%
                mutate(cluster=factor(cluster,levels=c("TME_5","TME_11","TME_12","TME_16","others"))) ,
              aes(x = cluster, y= .data[[ab1]], fill = cluster))+ 
    geom_hline(yintercept = others_med, linetype = "dashed", color = "black", linewidth = 0.7) +
    geom_violin() + 
    scale_fill_manual(values = c(ct.cols17,others="grey70")) +  # optionally restore this
    theme_minimal() +
    theme(
      axis.text = element_text(color="black",size = 28),
      axis.text.x = element_text(angle=90,hjust = 1, vjust = 0.5),
      axis.title = element_text(size = 28),
      legend.title = element_text(size = 20, margin = margin(b = 10)),
      legend.text = element_text(size = 16),
      legend.key.size = unit(1.2, "cm"),
      legend.spacing.y = unit(1, "cm"),
      legend.key.height = unit(1.2, "cm"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(size=1)
    ) +
    xlab("")  + 
    ylab(paste0(ab1," (log)")) +
    plot_annotation(
      title = paste0(ab1," (CN)"),
      theme = theme(
        plot.title = element_text(hjust =0.5, size = 50, margin = margin(b = 20)),
        plot.background = element_blank()
      )
    ) + theme(legend.position="none")
  
  stats_df <- compute_cluster_stats(dt.tme.1s, ab1)
  
  ypos <- max(dt.tme.1s[[ab1]], na.rm = TRUE) *1.05
  
  p <- p +
    geom_text(
      data = stats_df,
      aes(x = cluster, y = ypos, label = sig, color = sig_col),
      size = 16,
      show.legend = FALSE
    ) +
    scale_color_identity()
  
  ggsave(file.path(cycif_fig_dir, "spatial", "exps_in_tme_clusters", paste0("violin_",ab1,"_v2.pdf")),
     p, width = 5, height = 8, units = "in", dpi = 300)
}
```

## ---- Tumor Spatial Associations ----

```{r}
xx <- load(file.path(cycif_obj_dir,"cycif7_umap_tumor_and_macrophage.rda")) # dt.tum.1, dt.mac.1
```


```{r}
dim(frnn7.cmb)
dim(dt.tum.1)
table(frnn7.cmb$cell_types)
```

```{r}
lst.density <- list()
lst.density$cd8t <- list()
lst.density$cd4t <- list()
lst.density$t_other <- list()
```

### Tumor Marker Relationships
```{r}
frnn7.cmb.tum <- frnn7.cmb %>%
  filter(cell_types %in% c("Tumor_panCK","Tumor_neg")) %>%
  filter(is.sele.tum) %>%
  cbind(dt.tum.1 %>% dplyr::select(umap_x,umap_y)) %>%
  mutate(across(all_of(uniq.cts), as.numeric))
frnn7.cmb.mac <- frnn7.cmb %>%
  filter(cell_types %in% c("Mac_CD68","Mac_CD163","Mac_CD68_CD163")) %>%
  filter(is.sele.mac) %>%
  cbind(dt.mac.1 %>% dplyr::select(umap_x,umap_y)) %>%
  mutate(cluster = dt.mac.1 %>% pull(cluster)) %>%
  mutate(across(all_of(uniq.cts), as.numeric))
rngs <- list()
for(ab1 in   paste0(c("PDL1","cCaspase3","gH2AX", "pTBK1"),".log")){
  vals_ab1 <- c(
    frnn7.cmb.mac[[ab1]],
    frnn7.cmb.tum[[ab1]]
  )
  tmp <- vals_ab1[is.finite(vals_ab1)]
  if(ab1 == "pTBK1.log"){
    tmp <- tmp[tmp > -2]
  }else if(ab1 == "gH2AX.log"){
    tmp <- tmp[tmp > -1]
  }
  rngs[[ab1]] <- range(tmp)
}


```

Classification reconstructed from `bound_coords` (`vline` = PDL1 threshold, `curve` =
cCaspase3-vs-PDL1 boundary).
```{r}
bound_coords <- readRDS(file.path(cycif_obj_dir, "pdl1_cc3_boundary_coords.rds"))

classify_pdl1_cc3 <- function(df, xcol = "PDL1.log", ycol = "cCaspase3.log", bc = bound_coords) {
  x0 <- bc$vline$x
  curve_pts <- bc$curve[order(bc$curve$x), ]

  y_on_curve <- function(xq) {
    xi <- curve_pts$x; yi <- curve_pts$y
    xq2 <- pmin(pmax(xq, min(xi)), max(xi))
    i <- findInterval(xq2, xi, left.open = FALSE, rightmost.closed = TRUE)
    i <- pmin(pmax(i, 1), length(xi) - 1)
    x1 <- xi[i]; x2 <- xi[i + 1]; y1 <- yi[i]; y2 <- yi[i + 1]
    y1 + (xq2 - x1) * (y2 - y1) / pmax(x2 - x1, 1e-12)
  }

  x <- df[[xcol]]; y <- df[[ycol]]
  cls <- ifelse(
    x <= x0, "PDL1-",
    ifelse(y > y_on_curve(x), "PDL1+CC3+", "PDL1+CC3-")
  )
  factor(cls, levels = c("PDL1+CC3+", "PDL1+CC3-", "PDL1-"))
}

frnn7.cmb.tum$classification <- classify_pdl1_cc3(frnn7.cmb.tum)
frnn7.cmb.mac$classification <- classify_pdl1_cc3(frnn7.cmb.mac)
```

### [Fig 6O, EDF10J] Neighboring T-cell Density by PD-L1/cCasp3 and pTBK1/gH2AX Classification

Combines Tumor and Mac rows into one model per T-cell type, pooled across all samples/timepoints,
using both the PDL1/CC3 and gH2AX/pTBK1 classifications defined above.

```{r}
classify_gh2ax_ptbk1 <- function(df, xcol = "gH2AX.log", ycol = "pTBK1.log", v = 0.06300655, h = -0.09366919) {
  x <- df[[xcol]]; y <- df[[ycol]]
  cls <- (x > v) + (y > h) * 2 + 1
  factor(cls, levels = c(1,2,3,4),
         labels = c("gH2AX-/pTBK1-","gH2AX+/pTBK1-","gH2AX-/pTBK1+","gH2AX+/pTBK1+"))
}

frnn7.cmb.tum$classification_ptbk <- classify_gh2ax_ptbk1(frnn7.cmb.tum)
frnn7.cmb.mac$classification_ptbk <- classify_gh2ax_ptbk1(frnn7.cmb.mac)

fit_neighboring_tcell_effect <- function(class_col, out_dir, out_prefix) {
  # Explicit display order per classification scheme -- 4 Tumor then 4 Mac (or 3/3), and
  # within each source, PDL1+CC3-/PDL1+CC3+/PDL1- or gH2AX+/pTBK1+, gH2AX-/pTBK1+,
  # gH2AX+/pTBK1-, gH2AX-/pTBK1- -- rather than relying on interaction()'s default level
  # ordering (alphabetical/factor-definition order, not this specific sequence).
  classification_order <- list(
    classification = c("PDL1+CC3-", "PDL1+CC3+", "PDL1-"),
    classification_ptbk = c("gH2AX+/pTBK1+", "gH2AX-/pTBK1+", "gH2AX+/pTBK1-", "gH2AX-/pTBK1-")
  )[[class_col]]
  group_levels <- paste(
    rep(c("Tumor", "Mac"), each = length(classification_order)),
    rep(classification_order, times = 2),
    sep = "_"
  )

  # Significance comparisons, all nominal p (adjust = "none", no FDR), shown as raw p-value
  # text rather than stars:
  # - Within each source (Tumor, Mac) separately: every pairwise combination among classes
  #   (3 pairs for PDL1/CC3, 6 for gH2AX/pTBK1).
  # - Cross-source: Tumor vs Mac for each class (3 or 4), stacked above the within-source
  #   rows ("on top").
  xpos <- setNames(seq_along(group_levels), group_levels)
  within_source_pairs <- combn(classification_order, 2, simplify = FALSE)
  n_cross_rows <- length(classification_order)

  # Explicit bracket-height tiers (0 = lowest), keyed by local class position "i-j" within a
  # source (1-based index into classification_order, sorted so i<j) -- adjacent pairs share
  # the lowest tier, wider-spanning pairs stack above them so lines don't visually cross.
  # PDL1/CC3 (3 classes): (1,2)/(2,3) low, (1,3) above.
  # gH2AX/pTBK1 (4 classes): (2,3) lowest (own tier, since it's nested inside (2,4)); (1,2)/
  # (2,4) next; (1,3)/(3,4) above that; (1,4) (full span, not explicitly specified) highest.
  n_classes <- length(classification_order)
  tier_map <- if (n_classes == 3) {
    c("1-2" = 0, "2-3" = 0, "1-3" = 1)
  } else if (n_classes == 4) {
    c("2-3" = 0, "1-2" = 1, "2-4" = 1, "1-3" = 2, "3-4" = 2, "1-4" = 3)
  } else {
    stop("no explicit bracket-tier layout defined for ", n_classes, " classes")
  }
  n_within_rows <- max(tier_map) + 1

  within_tier_of <- function(cl1, cl2) {
    i <- match(cl1, classification_order)
    j <- match(cl2, classification_order)
    tier_map[[paste(sort(c(i, j)), collapse = "-")]]
  }

  # Two passes: build all 3 cell types' emm_df/contrasts first so a shared y-axis range
  # (across CD8T/CD4T/T_other, for this protein pair) can be computed before any plot is
  # drawn -- otherwise each of the 3 plots auto-scales independently and isn't comparable.
  emm_dfs <- list()
  sig_dfs <- list()
  for (ct in c("CD8T", "CD4T", "T_other")) {
    gr.n.ct <- bind_rows(
      frnn7.cmb.tum %>%
        transmute(sample, Patient.ID, source = "Tumor", classification = .data[[class_col]], ct_raw = .data[[ct]]),
      frnn7.cmb.mac %>%
        transmute(sample, Patient.ID, source = "Mac", classification = .data[[class_col]], ct_raw = .data[[ct]])
    ) %>%
      group_by(sample, Patient.ID, source, classification) %>%
      summarise(mean_tcells = mean(ct_raw, na.rm = TRUE), .groups = "drop") %>%
      mutate(
        group = paste0(source, "_", classification),
        Patient.ID = factor(Patient.ID)
      )

    fit <- lmer(mean_tcells ~ group + (1 | Patient.ID), data = gr.n.ct)
    emm <- emmeans(fit, ~ group)
    # tidyr::separate() (regex split via stringi/ICU) reproducibly triggered "vector memory
    # exhausted" on this ~6-8-row table on at least one machine (tidyr 1.3.1/stringi 1.8.4) --
    # a platform-specific stringi/ICU issue, not a real memory requirement. Splitting on the
    # first "_" with base R sub() avoids stringi entirely and is equivalent to the original
    # separate(group, into = c("source","classification"), sep = "_", extra = "merge").
    emm_dfs[[ct]] <- as.data.frame(emm) %>%
      mutate(
        source = sub("_.*", "", group),
        classification = sub("^[^_]+_", "", group),
        group_label = factor(group, levels = group_levels)
      ) %>%
      dplyr::select(-group) %>%
      mutate(source = factor(source, levels = c("Tumor", "Mac")))

    grid_groups <- as.character(emm@grid$group)
    contrast_vecs <- list()

    # Within-source: every pairwise combination among classes, for each source separately.
    for (src in c("Tumor", "Mac")) {
      for (pr in within_source_pairs) {
        name1 <- paste0(src, "_", pr[1])
        name2 <- paste0(src, "_", pr[2])
        vec <- setNames(rep(0, length(grid_groups)), grid_groups)
        vec[name1] <- 1
        vec[name2] <- -1
        contrast_vecs[[paste(name1, name2, sep = " - ")]] <- vec
      }
    }
    # Cross-source: Tumor vs Mac, for each class.
    for (cl in classification_order) {
      name1 <- paste0("Tumor_", cl)
      name2 <- paste0("Mac_", cl)
      vec <- setNames(rep(0, length(grid_groups)), grid_groups)
      vec[name1] <- 1
      vec[name2] <- -1
      contrast_vecs[[paste(name1, name2, sep = " - ")]] <- vec
    }

    contrast_df <- as.data.frame(contrast(emm, method = contrast_vecs, adjust = "none"))
    contrast_df$group1 <- sub(" - .+", "", contrast_df$contrast)
    contrast_df$group2 <- sub(".+ - ", "", contrast_df$contrast)
    contrast_df$source1 <- sub("_.*", "", contrast_df$group1)
    contrast_df$source2 <- sub("_.*", "", contrast_df$group2)
    contrast_df$is_cross_source <- contrast_df$source1 != contrast_df$source2

    contrast_df <- contrast_df %>%
      rowwise() %>%
      mutate(
        row_rank = if (is_cross_source) {
          n_within_rows + match(sub("^Tumor_", "", group1), classification_order)
        } else {
          1 + within_tier_of(
            sub("^(Tumor|Mac)_", "", group1),
            sub("^(Tumor|Mac)_", "", group2)
          )
        }
      ) %>%
      ungroup() %>%
      mutate(
        x1_full = xpos[group1],
        x2_full = xpos[group2],
        # Shorten each segment by ~3% of its span (inset symmetrically), so brackets don't
        # touch the data points/each other's endpoints directly.
        x1 = x1_full + 0.015 * (x2_full - x1_full),
        x2 = x2_full - 0.015 * (x2_full - x1_full),
        # sprintf() formats each p-value independently (2 significant figures via "%.1e",
        # e.g. "5.1e-04") -- unlike format(scientific=FALSE) on the whole vector, which pads
        # every element to the widest element's decimal width, appending spurious trailing
        # zeros to values that needed fewer digits.
        label = sprintf("%.1e", p.value)
      )

    sig_dfs[[ct]] <- contrast_df
  }

  n_total_rows <- n_within_rows + n_cross_rows
  ylim_data <- range(
    unlist(lapply(emm_dfs, function(d) c(d$lower.CL, d$upper.CL))),
    0 # keep the yintercept = 0 reference line visible on every plot
  )
  y_range <- diff(ylim_data)
  bracket_step <- 0.08 * y_range
  bracket_y0 <- ylim_data[2] + 0.06 * y_range
  ylim_shared <- c(ylim_data[1], bracket_y0 + n_total_rows * bracket_step)

  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  for (ct in names(emm_dfs)) {
    sig_df <- sig_dfs[[ct]] %>%
      mutate(y = bracket_y0 + (row_rank - 1) * bracket_step)

    p <- ggplot(emm_dfs[[ct]], aes(x = group_label, y = emmean)) +
      geom_point(size = 2) +
      geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
      geom_hline(yintercept = 0, linetype = 2, color = "grey60") +
      geom_segment(data = sig_df, aes(x = x1, xend = x2, y = y, yend = y), inherit.aes = FALSE) +
      geom_text(data = sig_df, aes(x = (x1 + x2) / 2, y = y, label = label), inherit.aes = FALSE, vjust = -0.3, size = 4) +
      coord_cartesian(ylim = ylim_shared) +
      xlab("") +
      ylab("Effect size (emmean, T cells per CN)") +
      labs(title = ct) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))

    ggsave(file.path(out_dir, paste0(out_prefix, "_", ct, ".pdf")), p, width = 7, height = 6, units = "in", dpi = 300)
  }
}

fit_neighboring_tcell_effect("classification", file.path(cycif_fig_dir, "spatial", "coexpression", "tcell_density_by_classification", "emmeans"), "emmeans_tcells_pdl1_cc3")
fit_neighboring_tcell_effect("classification_ptbk", file.path(cycif_fig_dir, "spatial", "coexpression", "tcell_density_by_classification", "emmeans"), "emmeans_tcells_ptbk1_gh2ax")
```

### [Fig 6N left, EDF10I left] Tumor Marker Relationships

#### PDL1 vs cCaspase3, in Tumor
```{r}
scatters <- list()
for(ct in uniq.cts[4:6]){
  scatters[[ct]] <- list()
  frnn7.cmb.tum <- frnn7.cmb.tum %>%
    mutate(ct_raw = !!sym(ct)) %>%
    mutate(
      density = {
        ct_vals <- ct_raw
        q <- quantile(ct_vals, probs = c(0.01, 0.99), na.rm = TRUE)
        pmin(pmax(ct_vals, q[1]), q[2])
      }
    )

  ab1 <- "PDL1.log"
  ab2 <- "cCaspase3.log"
  ab1s <- sub("\\.log","",ab1)
  ab2s <- sub("\\.log","",ab2)
  
  coords <- as.matrix(frnn7.cmb.tum[c(ab1,ab2)])

  knn <- get.knn(coords, k = 20)
  
  knn_indices <- cbind(1:nrow(coords), knn$nn.index)
  
  frnn7.cmb.tum$density <- apply(knn_indices, 1, function(idx) {
    mean(frnn7.cmb.tum$density[idx], na.rm = TRUE)
  })

  p_main <- ggplot(frnn7.cmb.tum , aes(x = .data[[ab1]], y = .data[[ab2]], col = density)) + #%>% arrange(density)
    geom_point() + 
    scale_color_distiller(palette = "YlGnBu", direction = 1, type = "seq",name = "mean\ndensity") +
    theme_minimal() + 
    theme(
      axis.text = element_text(size = 28,color="black"),
      axis.title = element_text(size = 32,color="black"),
      legend.title = element_text(size = 20, margin = margin(b = 10)),
      legend.text = element_text(size = 16),
      legend.key.size = unit(1.2, "cm"),
      legend.spacing.y = unit(1, "cm"),
      legend.key.height = unit(1.2, "cm"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(size=1,color="black")
    ) +
    geom_vline(xintercept = bound_coords$vline$x, linetype="dashed",color="black",linewidth=1) +
    geom_path(data = bound_coords$curve, aes(x = x, y = y), linetype="dashed",color="black",linewidth=1) +
    xlab(paste0(ab1s," (log) in Tumor")) +
    ylab(paste0(ab2s," (log) in Tumor")) +
    plot_annotation(
      title = ct,
      theme = theme(
        plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
        plot.background = element_blank()
      )
    ) 
  
  p_main <- p_main  + theme(legend.position="none") + common_scale(max_limit = 5)
  scatters[[ct]]$PDL_CC3 <- p_main 
  ggsave(file.path(cycif_fig_dir, "spatial", "coexpression", "tumor", paste0("scatter_",ab1s,"_",ab2s,"_",ct,".png")),
     p_main, width = 7, height = 7, units = "in", dpi = 300)
}
```

`boundaries2` is a manually-determined gH2AX/pTBK1 decision-boundary threshold, reused below
for both the tumor and macrophage versions of this plot.
```{r}
boundaries2 <- list(v=c(0.06300655),h=c(-0.09366919))
```

#### gH2AX vs pTBK1, in Tumor
```{r}
scatters <- list()
for(ct in uniq.cts[4:6]){
  scatters[[ct]] <- list()
  frnn7.cmb.tum <- frnn7.cmb.tum %>%
    mutate(ct_raw = !!sym(ct)) %>%
    mutate(
      density = {
        ct_vals <- ct_raw
        q <- quantile(ct_vals, probs = c(0.01, 0.99), na.rm = TRUE)
        pmin(pmax(ct_vals, q[1]), q[2])
      }
    )
  
  ab1 <- "gH2AX.log"
  ab2 <- "pTBK1.log"
  ab1s <- sub("\\.log","",ab1)
  ab2s <- sub("\\.log","",ab2)
  
  coords <- as.matrix(frnn7.cmb.tum[c(ab1,ab2)])

  knn <- get.knn(coords, k = 20)
  
  knn_indices <- cbind(1:nrow(coords), knn$nn.index)
  
  frnn7.cmb.tum$density <- apply(knn_indices, 1, function(idx) {
    mean(frnn7.cmb.tum$density[idx], na.rm = TRUE)
  })

  p_full <- ggplot(frnn7.cmb.tum , aes(x = .data[[ab1]], y = .data[[ab2]], col = density)) + #%>% arrange(density)
    geom_point() + 
    scale_color_distiller(palette = "YlGnBu", direction = 1, type = "seq",name = "mean\ndensity") +
    theme_minimal() + 
    theme(
      axis.text = element_text(size = 28,color="black"),
      axis.title = element_text(size = 32,color="black"),
      legend.title = element_text(size = 20, margin = margin(b = 10)),
      legend.text = element_text(size = 16),
      legend.key.size = unit(1.2, "cm"),
      legend.spacing.y = unit(1, "cm"),
      legend.key.height = unit(1.2, "cm"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(size=1,color="black")
    ) +
    coord_cartesian(
      xlim = rngs[[ab1]],
      ylim = rngs[[ab2]]
    ) +
    geom_vline(xintercept = boundaries2$v, linetype="dashed",color="black",linewidth=1) +
    geom_hline(yintercept = boundaries2$h, linetype="dashed",color="black",linewidth=1) +
    xlab(paste0(ab1s," (log) in Tumor")) + 
    ylab(paste0(ab2s," (log) in Tumor")) +
    plot_annotation(
      title = ct,
      theme = theme(
        plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
        plot.background = element_blank()
      )
    ) 
  
  p_main <- p_full  + theme(legend.position="none") + common_scale(max_limit = 3)
  scatters[[ct]]$gH2AX1_pTBK1 <- p_main 
  ggsave(file.path(cycif_fig_dir, "spatial", "coexpression", "tumor", paste0("scatter_",ab1s,"_",ab2s,"_",ct,".png")),
     p_main, width = 7, height = 7, units = "in", dpi = 300)  
}
save_legend_pdf(p_full + common_scale(max_limit =3), file.path(cycif_fig_dir, "spatial", "coexpression", "tumor", "umap_gh2ax_ptbk1_legend.pdf"), width = 4, height = 6)

```

#### TME-level scatterplots by cell type / antibody
```{r}
ab_pairs <- list(
  c("PDL1","PD1"),
  c("PDL1","cCaspase3"),
  c("gH2AX", "pTBK1")
)

scatters <- list()

for (ct in uniq.cts[4:6]) {
  scatters[[ct]] <- list()

  for (pair in ab_pairs) {
    ab1 <- pair[1]
    ab2 <- pair[2]

    xlab_text <- if (ab1 == "gH2AX") {
      expression(gamma * "H2AX (CN)")
    } else {
      paste0(ab1, " (CN)")
    }

    dt.tme.1s <- dt.tme.1 %>%
      mutate(ct_raw = !!sym(ct)) %>%
      mutate(
        density = {
          ct_vals <- ct_raw
          q <- quantile(ct_vals, probs = c(0.01, 0.99), na.rm = TRUE)
          pmin(pmax(ct_vals, q[1]), q[2])
        }
      ) %>%
      filter(gH2AX > -1 | pTBK1 > -1)

    coords <- as.matrix(dt.tme.1s[c(ab1, ab2)])
    knn <- get.knn(coords, k = 20)
    knn_indices <- cbind(1:nrow(coords), knn$nn.index)

    dt.tme.1s$density <- apply(knn_indices, 1, function(idx) {
      mean(dt.tme.1s$density[idx], na.rm = TRUE)
    })

    p_main <- ggplot(dt.tme.1s, aes(x = .data[[ab1]], y = .data[[ab2]], col = density)) +
      geom_point() +
      scale_color_distiller(palette = "YlGnBu", direction = 1, type = "seq", name = "mean\ndensity") +
      theme_minimal() +
      theme(
        axis.text = element_text(size = 28, color = "black"),
        axis.title = element_text(size = 32, color = "black"),
        legend.title = element_text(size = 20, margin = margin(b = 10)),
        legend.text = element_text(size = 16),
        legend.key.size = unit(1.2, "cm"),
        legend.spacing.y = unit(1, "cm"),
        legend.key.height = unit(1.2, "cm"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(size = 1, color = "black")
      ) +
      xlab(xlab_text) +
      ylab(paste0(ab2, " (CN)")) +
      plot_annotation(
        title = ct,
        theme = theme(
          plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
          plot.background = element_blank()
        )
      ) + common_scale(0.42)

    out_path <- file.path(
      cycif_fig_dir, "spatial", "coexpression", "tme", "ct",
      paste0("scatter_", ab1, "_", ab2, "_", ct, ".png")
    )
    ggsave(out_path, p_main + theme(legend.position = "none"), width = 7, height = 7, units = "in", dpi = 300)

    scatters[[ct]][[paste0(ab1, "_", ab2)]] <- p_main
  }
}

save_legend_pdf(
  p_main,
  file.path(cycif_fig_dir, "spatial", "coexpression", "tme", "ct", "scatter_celltype_legend.pdf"),
  width = 4,
  height = 6
)

for (ab in cst7.abs[c(2, 3, 4, 9, 10)]) {
  scatters[[ab]] <- list()

  for (pair in ab_pairs) {
    ab1 <- pair[1]
    ab2 <- pair[2]

    xlab_text <- if (ab1 == "gH2AX") {
      expression(gamma * "H2AX (CN)")
    } else {
      paste0(ab1, " (CN)")
    }

    dt.tme.1s <- dt.tme.1 %>%
      mutate(ab_raw = .data[[ab]]) %>%
      mutate(
        density = {
          ab_vals <- ab_raw
          q <- quantile(ab_vals, probs = c(0.01, 0.99), na.rm = TRUE)
          pmin(pmax(ab_vals, q[1]), q[2])
        }
      ) %>%
      filter(gH2AX > -1 | pTBK1 > -1)

    coords <- as.matrix(dt.tme.1s[c(ab1, ab2)])
    knn <- get.knn(coords, k = 20)
    knn_indices <- cbind(1:nrow(coords), knn$nn.index)

    dt.tme.1s$density <- apply(knn_indices, 1, function(idx) {
      mean(dt.tme.1s$density[idx], na.rm = TRUE)
    })

    p_main <- ggplot(dt.tme.1s, aes(x = .data[[ab1]], y = .data[[ab2]], col = density)) +
      geom_point() +
      viridis::scale_color_viridis(guide = "none") +
      theme_minimal() +
      theme(
        axis.text = element_text(size = 28, color = "black"),
        axis.title = element_text(size = 32, color = "black"),
        legend.title = element_text(size = 20, margin = margin(b = 10)),
        legend.text = element_text(size = 16),
        legend.key.size = unit(1.2, "cm"),
        legend.spacing.y = unit(1, "cm"),
        legend.key.height = unit(1.2, "cm"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(size = 1, color = "black")
      ) +
      xlab(xlab_text) +
      ylab(paste0(ab2, " (CN)")) +
      plot_annotation(
        title = ab,
        theme = theme(
          plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
          plot.background = element_blank()
        )
      )

    out_path <- file.path(
      cycif_fig_dir, "spatial", "coexpression", "tme", "ab",
      paste0("scatter_", ab1, "_", ab2, "_", ab, ".png")
    )
    ggsave(out_path, p_main, width = 7, height = 7, units = "in", dpi = 300)

    scatters[[ab]][[paste0(ab1, "_", ab2)]] <- p_main
  }
}

p_main <- ggplot(dt.tme.1s, aes(x = !!sym(ab1), y = !!sym(ab2), col = density)) +
  geom_point() +
  viridis::scale_color_viridis() +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 20),
    axis.title = element_text(size = 24),
    legend.title = element_text(size = 20, margin = margin(b = 10)),
    legend.text = element_text(size = 16),
    legend.key.size = unit(1.2, "cm"),
    legend.spacing.y = unit(1, "cm"),
    legend.key.height = unit(1.2, "cm"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(size = 1)
  ) +
  xlab(paste0(ab1, " (log)")) +
  ylab(paste0(ab2, " (log)")) +
  plot_annotation(
    title = ab,
    theme = theme(
      plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
      plot.background = element_blank()
    )
  )
save_legend_pdf(
  p_main,
  file.path(cycif_fig_dir, "spatial", "coexpression", "tme", "ab", "scatter_ab_legend.pdf"),
  width = 4,
  height = 6
)
```

## ---- Macrophage Spatial Analysis ----

### [Fig 5H] Macrophage UMAP with Density of Each Cell Type

```{r}
for(ct in uniq.cts){
  frnn7.cmb.mac <- frnn7.cmb.mac %>%
    mutate(ct_raw = !!sym(ct)) %>%
    mutate(
      density = {
        ct_vals <- ct_raw
        q <- quantile(ct_vals, probs = c(0.01, 0.99), na.rm = TRUE)
        pmin(pmax(ct_vals, q[1]), q[2])
      }
    )

  coords <- as.matrix(frnn7.cmb.mac[, c("umap_x", "umap_y")])
  
  knn <- get.knn(coords, k = 20)
  
  knn_indices <- cbind(1:nrow(coords), knn$nn.index)
  
  frnn7.cmb.mac$density <- apply(knn_indices, 1, function(idx) {
    mean(frnn7.cmb.mac$density[idx], na.rm = TRUE)
  })
  
  p_main <- umap_base_num(frnn7.cmb.mac, aes(x = umap_x, y = umap_y, col = density)) + 
    scale_color_distiller(palette = "YlGnBu", direction = 1, type = "seq",name = "mean\ndensity") +
    umap_styling_1() +
    plot_annotation(
      title = ct,
      theme = theme(
        plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
        plot.background = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank()
      )
    ) + theme(legend.position = "none") + common_scale(max_limit=4)

  if(ct == uniq.cts[1]){
    p_full <- umap_base_num(frnn7.cmb.mac, aes(x = umap_x, y = umap_y, col = density)) +
      umap_styling() +
      scale_color_distiller(
        palette = "YlGnBu",
        limits = c(0,4),
        direction = 1,
        oob = scales::squish,
        name = "Density"
      )
    save_legend_pdf(p_full, file.path(cycif_fig_dir, "umap", "mac", "ct_density", "umap_cts_legend_mac.pdf"), width = 4, height = 6)
  }
  ggsave(file.path(cycif_fig_dir, "umap", "mac", "ct_density", paste0("umap__x_",ct,"_v2.png")),
     p_main, width = 7, height = 7, units = "in", dpi = 300)  
}

```

### [Fig 5I, 5J, 5D] Macrophage Cluster Enrichment

#### Enrichment heatmap

Two enrichment tests here, both Fisher's exact, feeding two different outputs: `df_enrichment_p`
tests whether each *neighboring* cell type (all of `uniq.cts`, not just the macrophage's own
subtype label) is present among a macrophage's neighbors more than expected in each UMAP
cluster -- `mat_plot`/`sig_labels` (full `uniq.cts` panel) from this test feed the barplots
below. `mac_enrich_tbl` separately tests each macrophage subtype's own *frequency* within each
cluster (the three subtypes' counts sum to the cluster's total macrophage count) -- its output
(`mat_freq_enrich_t`, limited to the 3 macrophage subtype columns) is what the heatmap plots.
```{r}
dir.create(file.path(cycif_fig_dir, "umap", "mac", "enrichment"), recursive = TRUE, showWarnings = FALSE)

df_long <- frnn7.cmb.mac %>%
  dplyr::select(cluster, all_of(uniq.cts)) %>%
  pivot_longer(-cluster, names_to = "cell_type", values_to = "cell_count") %>%
  mutate(present = as.integer(cell_count > 0))

df_enrichment_p <- df_long %>%
  group_by(cell_type) %>%
  group_modify(~{
    d <- .x
    map_dfr(unique(d$cluster), function(cl) {
      d$in_cluster <- d$cluster == cl
      tab <- table(d$present, d$in_cluster)

      if (all(dim(tab) == c(2, 2))) {
        ft <- fisher.test(tab, alternative = "greater")
        odds <- ft$estimate
        pval <- ft$p.value
      } else {
        odds <- NA
        pval <- NA
      }

      tibble(
        cluster = cl,
        p_value = pval,
        n_in = sum(d$in_cluster),
        n_out = sum(!d$in_cluster),
        effect_odds_ratio = odds
      )
    })
  }) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"),
         nlogp = -log10(p_value))

df_sig <- df_enrichment_p %>%
  mutate(nlogp = ifelse(p_adj < 0.05, nlogp, 0))

mat_plot <- df_sig %>%
  dplyr::select(cluster, cell_type, nlogp) %>%
  pivot_wider(names_from = cell_type, values_from = nlogp) %>%
  arrange(cluster) %>%
  column_to_rownames("cluster") %>%
  as.matrix()

keep_cols <- intersect(uniq.cts, colnames(mat_plot))
mat_plot <- mat_plot[, keep_cols, drop = FALSE]

cap <- 200
mat_plot <- pmax(pmin(mat_plot, cap), -cap)
sig_labels <- ifelse(mat_plot > 50, "*", "")

mac_freq_tbl <- frnn7.cmb.mac %>%
  group_by(cluster, cell_types) %>%
  summarise(n = n(), .groups = "drop") %>%
  complete(cluster, cell_types, fill = list(n = 0))

tot_cluster_mac <- mac_freq_tbl %>% group_by(cluster) %>% summarise(n_cluster = sum(n), .groups = "drop")
tot_type_mac    <- mac_freq_tbl %>% group_by(cell_types) %>% summarise(n_type = sum(n), .groups = "drop")
N_total_mac     <- sum(mac_freq_tbl$n)

mac_enrich_tbl <- mac_freq_tbl %>%
  left_join(tot_cluster_mac, by = "cluster") %>%
  left_join(tot_type_mac,    by = "cell_types") %>%
  mutate(
    a = n,
    b = n_cluster - a,
    c = n_type - a,
    d = N_total_mac - a - b - c
  ) %>%
  rowwise() %>%
  mutate(
    ft = list(fisher.test(matrix(c(a, b, c, d), nrow = 2, byrow = TRUE),
                          alternative = "greater")),
    p_greater = ft$p.value
  ) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p_greater, method = "BH"),
         nlogp = -log10(p_greater))

mac_enrich_sig <- mac_enrich_tbl %>%
  mutate(nlogp = ifelse(p_adj < 0.05, nlogp, 0),
         nlogp = pmin(nlogp, 200))

mat_freq_enrich <- mac_enrich_sig %>%
  dplyr::select(cluster, cell_types, nlogp) %>%
  pivot_wider(names_from = cell_types, values_from = nlogp) %>%
  arrange(cluster) %>%
  column_to_rownames("cluster") %>%
  as.matrix()

mac_cols <- intersect(c("Mac_CD68", "Mac_CD163", "Mac_CD68_CD163"), colnames(mat_freq_enrich))
mat_freq_enrich <- mat_freq_enrich[, mac_cols, drop = FALSE]
sig_labels_freq <- ifelse(mat_freq_enrich > 50, "*", "")

mat_freq_enrich_t <- t(mat_freq_enrich)
sig_labels_freq_t <- t(sig_labels_freq)

col_fun <- colorRamp2(c(0, 200), c("white", brewer.pal(9,"Reds")[7]))

pdf(file.path(cycif_fig_dir, "umap", "mac", "enrichment", "heatmap_mac_enrichment_cell_types.pdf"),
    width = 7, height = 2.5)
Heatmap(
  mat_freq_enrich_t,
  name = "-log10(FDR)",
  col = col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 10),
  heatmap_legend_param = list(
    title = expression(-log[10]~FDR),
    at = seq(0, 200, length = 5),
    labels = as.character(seq(0, 200, length = 5))
  ),
  cell_fun = function(j, i, x, y, width, height, fill) {
    if (sig_labels_freq_t[i, j] != "") {
      grid.text(sig_labels_freq_t[i, j], x, y,
                gp = gpar(fontsize = 20, fontface = "bold"))
    }
  },
  column_title = "Macrophage subtype enrichment by cluster"
)
dev.off()

```

#### Cell types enriched near Mac_1 / Mac_2
```{r}
mat1_df <- as.data.frame(t(mat_plot["Mac_1", , drop = FALSE])) %>%
  rownames_to_column("cluster") %>%
  rename(value = Mac_1) %>%
  arrange(desc(value))
mat1_df$cluster <- factor(mat1_df$cluster, levels = mat1_df$cluster)

p1 <- ggplot(mat1_df, aes(x = cluster, y = value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = -log10(0.05), color = 2) +
  annotate("text", x = Inf, y = -log10(0.05), label = "p-adj=0.05",
           color = 2, hjust = 1, vjust = -0.5, size = 6) +
  labs(x = NULL, y = "Mac_1 abundance (-log10(p))") +
  theme_minimal() +
  ggtitle("Cell types enriched near Mac_1") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 16),
    axis.title.y = element_text(size = 16)
  )

mat2_df <- as.data.frame(t(mat_plot["Mac_2", , drop = FALSE])) %>%
  rownames_to_column("cluster") %>%
  rename(value = Mac_2) %>%
  arrange(desc(value))
mat2_df$cluster <- factor(mat2_df$cluster, levels = mat2_df$cluster)

p2 <- ggplot(mat2_df, aes(x = cluster, y = value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = -log10(0.05), color = 2) +
  annotate("text", x = Inf, y = -log10(0.05), label = "p-adj=0.05",
           color = 2, hjust = 1, vjust = -0.5, size = 6) +
  labs(x = NULL, y = "Mac_2 abundance (-log10(p))") +
  theme_minimal() +
  ggtitle("Cell types enriched near Mac_2") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 16),
    axis.title.y = element_text(size = 16)
  )

ggsave(file.path(cycif_fig_dir, "umap", "mac", "enrichment", "barplot_cts_abundance_mac1.pdf"), p1,
       width = 6, height = 6, dpi = 300)

ggsave(file.path(cycif_fig_dir, "umap", "mac", "enrichment", "barplot_cts_abundance_mac2.pdf"), p2,
       width = 6, height = 6, dpi = 300)
```


### [Fig 6N right, EDF10I right] Macrophage Marker Relationships

#### PDL1 vs cCaspase3, in Macrophages
```{r}
scatters1 <- list()
for(ct in uniq.cts[4:6]){
  scatters1[[ct]] <- list()
  frnn7.cmb.mac <- frnn7.cmb.mac %>%
    mutate(ct_raw = !!sym(ct)) %>%
    mutate(
      density = {
        ct_vals <- ct_raw
        q <- quantile(ct_vals, probs = c(0.01, 0.99), na.rm = TRUE)
        pmin(pmax(ct_vals, q[1]), q[2])
      }
    )

  ab1 <- "PDL1.log"
  ab2 <- "cCaspase3.log"
  ab1s <- sub("\\.log","",ab1)
  ab2s <- sub("\\.log","",ab2)
  
  coords <- as.matrix(frnn7.cmb.mac[c(ab1,ab2)])

  knn <- get.knn(coords, k = 20)
  
  knn_indices <- cbind(1:nrow(coords), knn$nn.index)
  
  frnn7.cmb.mac$density <- apply(knn_indices, 1, function(idx) {
    mean(frnn7.cmb.mac$density[idx], na.rm = TRUE)
  })

  p_main <- ggplot(frnn7.cmb.mac , aes(x = !!sym(ab1), y = !!sym(ab2), col = density)) +
    geom_point() + 
    scale_color_distiller(palette = "YlGnBu", direction = 1, type = "seq",name = "mean\ndensity") +
    theme_minimal() + 
    theme(
      axis.text = element_text(size = 28,color="black"),
      axis.title = element_text(size = 32,color="black"),
      legend.title = element_text(size = 20, margin = margin(b = 10)),
      legend.text = element_text(size = 16),
      legend.key.size = unit(1.2, "cm"),
      legend.spacing.y = unit(1, "cm"),
      legend.key.height = unit(1.2, "cm"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(size=1,color="black")
    ) +
    geom_vline(xintercept = bound_coords$vline$x, linetype="dashed",color="black",linewidth=1) +
    geom_path(data = bound_coords$curve, aes(x = x, y = y), linetype="dashed",color="black",linewidth=1) +    
    xlab(paste0(ab1s," (log) in Mac")) + 
    ylab(paste0(ab2s," (log) in Mac")) +
    plot_annotation(
      title = ct,
      theme = theme(
        plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
        plot.background = element_blank()
      )
    ) 
  p_main <- p_main  + theme(legend.position="none") + common_scale(max_limit = 5)
  scatters1[[ct]]$PDL_CC3 <- p_main 
  ggsave(file.path(cycif_fig_dir, "spatial", "coexpression", "mac", paste0("scatter_",ab1s,"_",ab2s,"_",ct,".png")),
     p_main, width = 7, height = 7, units = "in", dpi = 300)
}
```

#### gH2AX vs pTBK1, in Macrophages
```{r}
scatters1 <- list()
for(ct in uniq.cts[4:6]){
  scatters1[[ct]] <- list()
  frnn7.cmb.mac <- frnn7.cmb.mac %>%
    mutate(ct_raw = !!sym(ct)) %>%
    mutate(
      density = {
        ct_vals <- ct_raw
        q <- quantile(ct_vals, probs = c(0.01, 0.99), na.rm = TRUE)
        pmin(pmax(ct_vals, q[1]), q[2])
      }
    )

  ab1 <- "gH2AX.log"
  ab2 <- "pTBK1.log"
  ab1s <- sub("\\.log","",ab1)
  ab2s <- sub("\\.log","",ab2)
  
  coords <- as.matrix(frnn7.cmb.mac[c(ab1,ab2)])

  knn <- get.knn(coords, k = 20)
  
  knn_indices <- cbind(1:nrow(coords), knn$nn.index)
  
  frnn7.cmb.mac$density <- apply(knn_indices, 1, function(idx) {
    mean(frnn7.cmb.mac$density[idx], na.rm = TRUE)
  })

  p_main <- ggplot(frnn7.cmb.mac , aes(x = .data[[ab1]], y = .data[[ab2]], col = density)) +
    geom_point() + 
    scale_color_distiller(palette = "YlGnBu", direction = 1, type = "seq",name = "mean\ndensity") +
    theme_minimal() + 
    theme(
      axis.text = element_text(size = 28,color="black"),
      axis.title = element_text(size = 32,color="black"),
      legend.title = element_text(size = 20, margin = margin(b = 10)),
      legend.text = element_text(size = 16),
      legend.key.size = unit(1.2, "cm"),
      legend.spacing.y = unit(1, "cm"),
      legend.key.height = unit(1.2, "cm"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(size=1,color="black")
    ) +
    coord_cartesian(
      xlim = rngs[[ab1]],
      ylim = rngs[[ab2]]
    ) +
    geom_vline(xintercept = boundaries2$v, linetype="dashed",color="black",linewidth=1) +
    geom_hline(yintercept = boundaries2$h, linetype="dashed",color="black",linewidth=1) +
    xlab(paste0(ab1s," (log) in Mac")) + 
    ylab(paste0(ab2s," (log) in Mac")) +
    plot_annotation(
      title = ct,
      theme = theme(
        plot.title = element_text(size = 50, hjust = 0.5, margin = margin(b = 20)),
        plot.background = element_blank()
      )
    ) 
  p_main <- p_main  + theme(legend.position="none") + common_scale(max_limit = 5)
  scatters1[[ct]]$gH2AX_pTBK1 <- p_main 
  ggsave(file.path(cycif_fig_dir, "spatial", "coexpression", "mac", paste0("scatter_",ab1s,"_",ab2s,"_",ct,".png")),
     p_main, width = 7, height = 7, units = "in", dpi = 300)
}
```

#### Base-R version (last cell type from the loop above, for a custom density-color legend)
```{r}
pal_fun <- colorRampPalette(RColorBrewer::brewer.pal(9, "YlGnBu"))
col_vec <- pal_fun(100)

density_scaled <- frnn7.cmb.mac$density
density_scaled <- (density_scaled - min(density_scaled, na.rm = TRUE)) /
                  (max(density_scaled, na.rm = TRUE) - min(density_scaled, na.rm = TRUE))

col_idx <- pmax(1, pmin(100, round(density_scaled * 100)))
pt_cols <- col_vec[col_idx]

plot(
  frnn7.cmb.mac[[ab1]], frnn7.cmb.mac[[ab2]],
  col = pt_cols, pch = 16, cex = 0.8,
  xlab = paste0(ab1s, " (log) in Mac"),
  ylab = paste0(ab2s, " (log) in Mac"),
  main = ct,
  cex.lab = 1.6,
  cex.axis = 1.3,
  cex.main = 2.2,
  las = 1
)

usr <- par("usr")
legend_x <- usr[2] + (usr[2] - usr[1]) * 0.02
legend_y <- seq(usr[3], usr[4], length.out = 100)

par(xpd = TRUE)
rect(
  xleft = legend_x,
  ybottom = legend_y[-100],
  xright = legend_x + (usr[2]-usr[1])*0.03,
  ytop = legend_y[-1],
  col = col_vec,
  border = NA
)

text(legend_x + (usr[2]-usr[1])*0.05, legend_y[1], "High\nDensity",
     adj = 0, cex = 1.2)
text(legend_x + (usr[2]-usr[1])*0.05, legend_y[100], "Low\nDensity",
     adj = 0, cex = 1.2)
par(xpd = FALSE)

```
## ---- T-Cell Spatial Analysis ----

### Define T-Cell Spatial Inputs
```{r}
frnn7.cmb.cd8t <- frnn7.cmb %>% 
  filter(cell_types %in% "CD8T") %>% 
  mutate(across(all_of(uniq.cts), as.numeric))

frnn7.cmb.cd4t <- frnn7.cmb %>% 
  filter(cell_types %in% "CD4T") %>% 
  mutate(across(all_of(uniq.cts), as.numeric))

frnn7.cmb.t_other <- frnn7.cmb %>% 
  filter(cell_types %in% "T_other") %>% 
  mutate(across(all_of(uniq.cts), as.numeric))
```

#### [EDF9G] Frequency of T Cell-Rich Clusters

`n` = total neighboring-cell count per row.
```{r}
n <- rowSums(as.matrix(frnn7.cmb[, uniq.cts]))

df_plot <- frnn7.cmb %>%
  filter(n >= 10 & is.sele.7) %>%
  mutate(across(all_of(uniq.cts), as.numeric)) %>%
  mutate(
    across(all_of(uniq.cts), ~ .x + runif(length(.x), min = -0.2, max = 0.2)),
    across(all_of(uniq.cts), ~ pmax(.x, 0))
  ) %>%
  rowwise() %>%
  mutate(
    total_cells = sum(c_across(all_of(uniq.cts)), na.rm = TRUE),
    CD8T_freq   = ifelse(total_cells > 0, CD8T / total_cells, 0),
    CD4T_freq   = ifelse(total_cells > 0, CD4T / total_cells, 0),
    Tother_freq = ifelse(total_cells > 0, T_other / total_cells, 0)
  ) %>%
  ungroup()

df_bar <- df_plot %>%
  group_by(cluster) %>%
  summarise(
    CD8T = mean(CD8T_freq, na.rm = TRUE),
    CD4T = mean(CD4T_freq, na.rm = TRUE),
    T_other = mean(Tother_freq, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(CD8T, CD4T, T_other),
               names_to = "CellType", values_to = "Frequency") %>%
  ungroup()

df_bar <- df_bar %>%
  mutate(cluster = factor(cluster, levels = unique(cluster)),
         CellType = factor(CellType, levels = rev(c("CD8T", "CD4T", "T_other"))))

p_bar <- ggplot(df_bar, aes(x = cluster, y = Frequency, fill = CellType)) +
  geom_bar(stat = "identity", width = 0.8, color = "black", size = 0.2) +
  scale_fill_manual(values = ct.cols7[c("CD8T", "CD4T", "T_other")]) +
  labs(
    x = "",
    y = "Mean Frequency (fraction of all cells)",
    fill = "Cell Type"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 12, color = "black"),
    axis.text.y = element_text(size = 12, color = "black"),
    axis.title = element_text(size = 14, color = "black"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )

ggsave(file.path(cycif_fig_dir, "umap", "tme", "expression_by_cluster", "tcell_freq_by_tme_cluster.pdf"),
       p_bar, width = 5, height = 6, units = "in", dpi = 300)
```

### [Fig 4K, 4J] T-Cell Density and Colocalization

#### PD1 vs neighboring T-cell density, per center/neighbor pair
```{r}
datasets <- list(
  CD8T   = frnn7.cmb.cd8t %>% filter(!is.na(cluster)),
  CD4T   = frnn7.cmb.cd4t %>% filter(!is.na(cluster)),
  T_other = frnn7.cmb.t_other %>% filter(!is.na(cluster))
)

cell_types <- c("CD8T", "CD4T", "T_other")

df <- bind_rows(lapply(names(datasets), function(obj_name) {
  obj <- datasets[[obj_name]]
  
  bind_rows(lapply(cell_types, function(ct) {
    x1 <- obj$PD1.log
    y1 <- obj[[ct]]
    cl <- obj$cluster
    
    if (obj_name == ct) {
      y1 <- y1 - 1
    }
    
    keep <- is.finite(x1) & is.finite(y1)
    
    data.frame(
      center = obj_name,
      neighbor = ct,
      PD1 = x1[keep],
      density = y1[keep],
      cluster = cl[keep]
    )
  }))
})) %>%
  mutate(
    center = factor(center, levels = c("CD8T", "CD4T", "T_other")),
    neighbor = factor(neighbor, levels = c("CD8T", "CD4T", "T_other"))
  )

stats_df <- df %>%
  group_by(center, neighbor) %>%
  summarize(
    slope = {
      fit <- try(lm(PD1 ~ density), silent = TRUE)
      if (inherits(fit, "try-error") || any(is.na(coef(fit)))) NA_real_
      else unname(coef(fit)[2])
    },
    r2 = {
      fit <- try(lm(PD1 ~ density), silent = TRUE)
      if (inherits(fit, "try-error") || any(is.na(coef(fit)))) NA_real_
      else summary(fit)$r.squared
    },
    pval = {
      fit <- try(lm(PD1 ~ density), silent = TRUE)
      if (inherits(fit, "try-error") || any(is.na(coef(fit)))) NA_real_
      else summary(fit)$coefficients[2, 4]
    },
    .groups = "drop"
  )


df_density <- df %>%
  group_by(center, neighbor) %>%
  mutate(
    ok = is.finite(PD1) & is.finite(density),
    density_col = NA_character_,
    density_col = replace(
      density_col, ok,
      grDevices::densCols(
        density[ok], PD1[ok],
        colramp = colorRampPalette(viridisLite::magma(256))
      )
    )
  ) %>%
  ungroup()

p_density <- ggplot(df_density, aes(density, PD1, colour = density_col)) +
  geom_point(size = .5) +
  scale_colour_identity() +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 0.8) +
  facet_grid(center ~ neighbor, switch = "y") +
  geom_text(
    data = stats_df,
    aes(
      x = Inf, y = -Inf,
      label = sprintf("R² = %.2f\np = %.3g\nslope = %.3f", r2, pval,slope)
    ),
    hjust = 1.1, vjust = -0.2, size = 4,
    inherit.aes = FALSE
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.background = element_blank(),
    strip.placement = "outside",
    strip.text.x = element_text(size = 12),
    strip.text.y.left = element_text(angle = 90, size = 12),
    plot.title = element_text(size = 14, hjust = 0.5),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 14),
    legend.position = "none"
  ) +
  labs(
    title = "Density (neighboring T cells/CN)",
    y = "PD1 expression in center of CN"
  )

ggsave(
  file.path(cycif_fig_dir, "spatial", "tcell_density", "pd1_tcell_density_slope", "correlation_tcell_pd1_density_v5.pdf"),
  p_density, width = 7, height = 7
)
```

#### Spatial co-occurrence heatmap across cell types
```{r}
df.cts <- frnn7.cmb %>%
  filter(n >= 10 & is.sele.7) %>%
  dplyr::select(all_of(uniq.cts)) %>%
  as.data.frame()

mat.cts <- as.matrix(df.cts)

row_sums <- rowSums(mat.cts, na.rm = TRUE)
mat.freq <- sweep(mat.cts, 1, row_sums, FUN = "/")
mat.freq[!is.finite(mat.freq)] <- 0

sp <- cor(mat.freq, method = "spearman", use = "pairwise.complete.obs")
sp_df <- as.matrix(sp)

mat_sig <- sp_df > 0.25
diag(mat_sig) <- FALSE
mat_sig[is.na(mat_sig)] <- FALSE

pdf(file.path(cycif_fig_dir, "spatial", "tcell_density", "pd1_tcell_density_slope", "heatmap_spearman_correlation_celltypes.pdf"),
    width = 6.5, height = 6)

ht <- Heatmap(
  sp_df,
  name = "cell type colocalization",
  col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_dend = TRUE,
  show_column_dend = TRUE,
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 10),
  heatmap_legend_param = list(
    title = "Spearman",
    at = seq(-1, 1, length.out = 5),
    labels = as.character(seq(-1, 1, length.out = 5))
  ),
  column_title = "Spatial co-occurrence of cell types",
  rect_gp = gpar(col = NA),
  cell_fun = function(j, i, x, y, w, h, fill) {
    if (mat_sig[i, j]) {
      grid.rect(
        x = x, y = y, width = w, height = h,
        gp = gpar(fill = NA, col = "black", lwd = 1.5)
      )
    }
  }
)

ComplexHeatmap::draw(ht, heatmap_legend_side = "right")
dev.off()
  
```

### Partial-residual grid function, feeds [EDF6E] below

For a chosen focus density, nets out the *other two* densities' fitted contribution from PD1
(using that bt3 group's own coefficients), then plots the residual against the focus density --
the slope of that residual scatter is the focus density's *adjusted* effect, net of the other
two. Built from `frnn7.cmb` directly (same source as `df1`), using all available cells per bt3
group rather than the 500-capped balanced sample -- the balancing only mattered for the single
pooled brms fit, not for separate per-group regressions.

```{r}
## coefs: bt3 x term coefficient table used to net out the two non-focus terms.
build_partial_residual_grid <- function(fit_result, center_type, coefs) {
  df_full <- fit_result$df_full
  terms <- c("CD8T", "CD4T", "T_other")

  grid_df <- purrr::map_dfr(terms, function(focus_term) {
    other_terms <- setdiff(terms, focus_term)
    df_full %>%
      left_join(
        coefs %>% filter(term %in% other_terms) %>%
          dplyr::select(bt3, term, slope) %>%
          pivot_wider(names_from = term, values_from = slope, names_prefix = "b_"),
        by = "bt3"
      ) %>%
      mutate(
        PD1_adj = PD1 - .data[[paste0("b_", other_terms[1])]] * .data[[other_terms[1]]] -
                        .data[[paste0("b_", other_terms[2])]] * .data[[other_terms[2]]],
        focus_density = .data[[focus_term]],
        focus_term = focus_term
      ) %>%
      dplyr::select(bt3, focus_term, focus_density, PD1_adj)
  }) %>%
    mutate(focus_term = factor(focus_term, levels = terms))

  ann_df <- coefs %>%
    rename(focus_term = term) %>%
    mutate(focus_term = factor(focus_term, levels = terms))

  p <- ggplot(grid_df, aes(focus_density, PD1_adj)) +
    geom_point(size = 0.4, alpha = 0.4) +
    geom_smooth(method = "lm", se = TRUE, color = "red", fill = "red", alpha = 0.15, linewidth = 0.8) +
    facet_grid(focus_term ~ bt3, switch = "y") +
    ## Two single-line layers (not one multi-line label) so both lines' right edges align.
    geom_text(
      data = ann_df,
      aes(x = Inf, y = -Inf, label = sprintf("slope = %.3f", slope)),
      hjust = 1.1, vjust = -1.6, size = 3.5, inherit.aes = FALSE
    ) +
    geom_text(
      data = ann_df,
      aes(x = Inf, y = -Inf, label = sprintf("95%% CrI [%.3f, %.3f]", ci_low, ci_high)),
      hjust = 1.1, vjust = -0.2, size = 3.5, inherit.aes = FALSE
    ) +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.x = element_text(size = 12),
      strip.text.y.left = element_text(angle = 90, size = 12),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14),
      legend.position = "none"
    ) +
    labs(x = "Density", y = paste0("PD-1 in ", center_type, "+ T (partial residual)"),
         title = paste0("PD1 in ", center_type, " ~ density, adjusted (partial residual), by bt3"))

  list(grid_df = grid_df, plot = p)
}
```

### Hierarchical (partial-pooling) alternative: brms with bt3 as a random-slope grouping factor, instead of six separate unpooled lmer fits per center

The `mv_fits`/`stats_df` pipeline above treats each bt3 group's coefficients as fully
independent -- WT_BX3's estimate reflects only its own 218 cells, with no borrowing of
strength from the other five groups, which is why its uncertainty dwarfs the others. This
fits one brms model per center (same structure as the primary `fit_full` model, just scoped
to one center's cells) with `bt3` as a random-slope grouping factor, then extracts each
bt3 group's *shrunk* coefficient via `coef()` -- population fixed effect + that group's
partially-pooled deviation, pulled toward the mean in proportion to how little data
supports it. Feeds the same slope-vs-mean-CD4T-density meta-regression as before.

```{r}
if(0){
  fit_hier_by_center <- function(center_type) {
    df_full <- frnn7.cmb %>%
      mutate(bt3 = factor(paste0(BRCA, "_", TimePoint),
                           levels = c("MUT_BS","MUT_BX2","MUT_BX3","WT_BS","WT_BX2","WT_BX3"))) %>%
      filter(n >= 10 & is.sele.7) %>%
      filter(cell_types == center_type) %>%
      dplyr::select(PD1, CD8T, CD4T, T_other, Patient.ID, bt3)
  
    if ("package:org.Hs.eg.db" %in% search()) detach("package:org.Hs.eg.db", unload = TRUE)
  
    fit <- brm(
      PD1 ~ CD8T + CD4T + T_other + (1 + CD8T + CD4T + T_other || bt3) + (1 | Patient.ID),
      data = df_full, family = student(), iter = 4000, chains = 4, cores = 4, seed = 123
    )
  
    list(df_full = df_full, fit = fit)
  }
  
  ## Fitting three Bayesian models -- this will take a while, same order of magnitude as fit_full earlier.
  hier.CD4 <- fit_hier_by_center("CD4T")
  hier.CD8 <- fit_hier_by_center("CD8T")
  hier.T_other <- fit_hier_by_center("T_other")
  save(hier.CD4, hier.CD8, hier.T_other, file = file.path(cycif_obj_dir, "cycif7_tcell_density_bayesian_fits.rda"))
}else{
  load(file.path(cycif_obj_dir,"cycif7_tcell_density_bayesian_fits.rda")) # hier.CD4, hier.CD8, hier.T_other
}
```

##### Convergence Check
```{r}
cat("CD4T-center model, max Rhat:", max(brms::rhat(hier.CD4$fit), na.rm = TRUE), "\n")
cat("CD8T-center model, max Rhat:", max(brms::rhat(hier.CD8$fit), na.rm = TRUE), "\n")
cat("T_other-center model, max Rhat:", max(brms::rhat(hier.T_other$fit), na.rm = TRUE), "\n")

np.CD4 <- brms::nuts_params(hier.CD4$fit)
np.CD8 <- brms::nuts_params(hier.CD8$fit)
np.T_other <- brms::nuts_params(hier.T_other$fit)
cat("CD4T-center divergences:", sum(np.CD4$Value[np.CD4$Parameter == "divergent__"]),
    "out of", brms::ndraws(hier.CD4$fit), "\n")
cat("CD8T-center divergences:", sum(np.CD8$Value[np.CD8$Parameter == "divergent__"]),
    "out of", brms::ndraws(hier.CD8$fit), "\n")
cat("T_other-center divergences:", sum(np.T_other$Value[np.T_other$Parameter == "divergent__"]),
    "out of", brms::ndraws(hier.T_other$fit), "\n")
```

Extracts each bt3 group's shrunk coefficient for each density term.
```{r}
extract_bt3_coefs <- function(fit_result, center_type) {
  cf <- coef(fit_result$fit)$bt3[, , c("CD8T", "CD4T", "T_other")]
  purrr::map_dfr(c("CD8T", "CD4T", "T_other"), function(term) {
    as.data.frame(cf[, , term]) %>%
      rownames_to_column("bt3") %>%
      mutate(term = term)
  }) %>%
    rename(slope = Estimate, ci_low = Q2.5, ci_high = Q97.5) %>%
    mutate(center = center_type)
}

hier_coefs.CD4 <- extract_bt3_coefs(hier.CD4, "CD4T")
hier_coefs.CD8 <- extract_bt3_coefs(hier.CD8, "CD8T")
hier_coefs.T_other <- extract_bt3_coefs(hier.T_other, "T_other")
```

### [EDF6E] Partial-residual grid, CD8T center, adjustment from the brm() hierarchical fit

For CD8T-center cells, nets out the other two densities' fitted contribution from PD1 using
`hier_coefs.CD8` (the brm() hierarchical fit's shrunk per-bt3 estimates), then plots the
residual against each focus density, individual points + red regression line, faceted bt3 x
term.

```{r}
grid.CD8.hier <- build_partial_residual_grid(hier.CD8, "CD8T", coefs = hier_coefs.CD8)
grid.CD8.hier$plot
ggsave(file.path(cycif_fig_dir, "spatial", "tcell_density", "pd1_tcell_density_slope", "correlation_tcell_pd1-cd8t_density_by_bt3_hier.pdf"),
       grid.CD8.hier$plot, width = 11, height = 6, units = "in", dpi = 300)
```

### [EDF6F] Boxplot: shrunk slopes across bt3, for PD1 in CD4T/CD8T ~ CD4T/CD8T density

Direct visualization of the finding this whole exploration converged on: the shrunk (hierarchical,
partially-pooled) per-bt3 slope for each of the four center x focus-density combinations,
boxplotted across the six bt3 groups with each group's own point and label shown -- CD4T-density's
slope should be visibly consistent across bt3 (tight box), while CD8T-density's varies a lot more
(wide box). `facet_wrap(~ center)`, each panel: paired Wilcoxon signed-rank tests (paired by
bt3) comparing the three neighbor terms pairwise, brackets/p-values overlaid.

```{r}
slope_box_df <- bind_rows(hier_coefs.CD4, hier_coefs.CD8, hier_coefs.T_other) %>%
  mutate(
    center = factor(center, levels = c("CD4T", "CD8T", "T_other")),
    term   = factor(term,   levels = c("CD4T", "CD8T", "T_other"))
  )

comparisons <- list(c("CD4T", "CD8T"), c("CD4T", "T_other"), c("CD8T", "T_other"))

wilcox_df <- slope_box_df %>%
  group_by(center) %>%
  group_modify(~ {
    map_dfr(comparisons, function(comp) {
      d <- .x %>% dplyr::select(bt3, term, slope) %>%
        pivot_wider(names_from = term, values_from = slope)
      w <- try(suppressWarnings(wilcox.test(d[[comp[1]]], d[[comp[2]]], paired = TRUE)), silent = TRUE)
      tibble(term1 = comp[1], term2 = comp[2],
             pval = if (inherits(w, "try-error")) NA_real_ else w$p.value)
    })
  }) %>%
  ungroup() %>%
  mutate(label = ifelse(is.na(pval), "", paste0("p=", signif(pval, 2))))

term_levels <- c("CD4T", "CD8T", "T_other")
x_pos <- setNames(seq_along(term_levels), term_levels)
y_base <- max(slope_box_df$slope, na.rm = TRUE)
y_offsets <- c("CD4T_CD8T" = y_base * 1.10, "CD8T_T_other" = y_base * 1.10,
               "CD4T_T_other" = y_base * 1.22)

ann_df <- wilcox_df %>%
  mutate(
    x1 = x_pos[term1] + 0.05,
    x2 = x_pos[term2] - 0.05,
    xmid = (x1 + x2) / 2,
    contrast = paste(term1, term2, sep = "_"),
    y = y_offsets[contrast]
  )

p_slope_box <- ggplot(slope_box_df, aes(term, slope)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey50") +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.3, color = "black") +
  geom_line(aes(group = bt3), color = "grey60", linewidth = 0.5) +
  geom_point(aes(color = bt3), size = 2.5) +
  geom_segment(data = ann_df, aes(x = x1, xend = x2, y = y, yend = y),
               inherit.aes = FALSE, color = "black", linewidth = 0.4) +
  geom_text(data = ann_df, aes(x = xmid, y = y * 1.02, label = label),
            inherit.aes = FALSE, size = 4, vjust = 0) +
  facet_wrap(~ center, nrow = 1) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  theme_bw(base_size = 14) +
  labs(x = "Neighboring T-cell type", y = "Slope", color = "bt3",
       title = "PD1 ~ density slope by bt3") +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

p_slope_box

ggsave(file.path(cycif_fig_dir, "spatial", "tcell_density", "pd1_tcell_density_slope", "slope_boxplot_by_bt3_3x3.pdf"),
       p_slope_box, width = 8, height = 6, dpi = 300)
```

### Mean CD4T / CD8T density per sample, by BRCA x TimePoint (plot_ct)

Explains what's driving the x-axis position in every meta-regression panel above: the actual
per-patient mean density values underlying "mean CD4T/CD8T density" per bt3 group, plotted with
the project's standard `plot_ct()` framework (per-patient trajectories across bt3, paired
LMER + emmeans contrasts). Density averaged across all QC-passing T cells per sample (not restricted to one center
type), matching the existing QC filters used throughout. Self-density only: CD4T density
among CD4T cells, CD8T density among CD8T cells.

```{r}
if (!exists("pt.cols")) pt.cols <- load_pt_colors()

mean_cd4t_density_by_sample <- frnn7.cmb %>%
  filter(n >= 10 & is.sele.7) %>%
  filter(cell_types == "CD4T") %>%
  group_by(sample) %>%
  summarize(mean_CD4T_density = mean(CD4T, na.rm = TRUE), .groups = "drop")

mean_cd8t_density_by_sample <- frnn7.cmb %>%
  filter(n >= 10 & is.sele.7) %>%
  filter(cell_types == "CD8T") %>%
  group_by(sample) %>%
  summarize(mean_CD8T_density = mean(CD8T, na.rm = TRUE), .groups = "drop")

mean_density_by_sample <- mean_cd4t_density_by_sample %>%
  full_join(mean_cd8t_density_by_sample, by = "sample") %>%
  left_join(pd7 %>% dplyr::select(id, BRCA, BOR, bt3, TimePoint, Patient.ID), by = c("sample" = "id"))

p_mean_cd4t_by_bt3 <- plot_ct(mean_density_by_sample, y_var = "mean_CD4T_density",
                                y_label = "Mean CD4T density per sample",
                                title = "Mean CD4T density by BRCA x TimePoint",
                                pt.cols = pt.cols)$plot
p_mean_cd4t_by_bt3

ggsave(file.path(cycif_fig_dir, "spatial", "tcell_density", "pd1_tcell_density_slope", "plot_ct_mean_cd4t_density_by_bt3.pdf"),
       p_mean_cd4t_by_bt3, width = 5, height = 7)

p_mean_cd8t_by_bt3 <- plot_ct(mean_density_by_sample, y_var = "mean_CD8T_density",
                                y_label = "Mean CD8T density per sample",
                                title = "Mean CD8T density by BRCA x TimePoint",
                                pt.cols = pt.cols)$plot
p_mean_cd8t_by_bt3

ggsave(file.path(cycif_fig_dir, "spatial", "tcell_density", "pd1_tcell_density_slope", "plot_ct_mean_cd8t_density_by_bt3.pdf"),
       p_mean_cd8t_by_bt3, width = 5, height = 7)
```

### [Fig 6I, 4L, 4M, EDF10A] T-Cell Regression and Expression Mapping

#### PD1 vs density, highlighting TME-11/TME-12
```{r}
df2 <- df %>%
  filter(center %in% c("CD4T","CD8T"), neighbor %in% c("CD4T","CD8T")) %>%
  mutate(
    cluster_id = str_replace_all(cluster, "TME_", "TME-"),
    cluster_id = factor(cluster_id, levels = paste0("TME-", 1:17)),
    
    cluster_plot = case_when(
      cluster_id %in% c("TME-11","TME-12") ~ as.character(cluster_id),
      TRUE ~ "others"
    )
  ) %>%
  arrange(cluster_plot != "others")   # others first (behind), highlighted last


ct.cols17.1 <- ct.cols17
names(ct.cols17.1) <- sub("_","-",names(ct.cols17))
plot_cols <- c(
  others = "grey70",
  unname(ct.cols17.1[c("TME-11", "TME-12")])
)
names(plot_cols)[2:3] <- c("TME-11", "TME-12")

p_density <- ggplot(
  df2,
  aes(density, PD1, colour = cluster_plot)
) +
  geom_point(size = 0.4) +
  scale_colour_manual(values = plot_cols) +
  facet_grid(center ~ neighbor, switch = "y") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.background = element_blank(),
    strip.placement = "outside",
    strip.text.x = element_text(size = 12),
    strip.text.y.left = element_text(angle = 90, size = 12),
    axis.text.x  = element_text(size = 16, color = "black"),
    axis.text.y  = element_text(size = 15, color = "black"),
    axis.title.x = element_text(size = 17, margin = margin(t = 2, r = 4), color = "black"),
    axis.title.y = element_text(size = 17, margin = margin(r = 4), color = "black"),
    plot.title   = element_text(size = 20, hjust = 0.5, color = "black"),
    legend.text  = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 15, color = "black"),
    strip.text   = element_text(size = 16, color = "black"),
    legend.key.size = unit(1.2, "cm"),
    legend.position = "right"
  ) +
  labs(
    title = "TME-11 and TME-12",
    x = "Density",
    y = "PD1",
    colour = "CN cluster"
  )

ggsave(
  file.path(cycif_fig_dir, "spatial", "tcell_density", "tme_clusters", "correlation_tcell_pd1_density_TME11_12.pdf"),
  p_density,
  width = 6.5,
  height = 5,
  units = "in",
  dpi = 300
)
```

#### Full brms model: PD1 ~ center x density, bt3 as random-slope grouping

No bt3-balancing cap: `fit_full`'s random-slope term handles per-group shrinkage directly.
`center` (3 levels) is too few for a random-slope variance component, so it's a fixed
interaction; `bt3` (6 levels) is its own non-nested random-slope grouping.
```{r}
df1 <- frnn7.cmb %>%
  mutate(bt3 = factor(paste0(BRCA, "_", TimePoint),
                       levels = c("MUT_BS","MUT_BX2","MUT_BX3","WT_BS","WT_BX2","WT_BX3"))) %>%
  filter(n >= 10 & is.sele.7) %>%
  filter(cell_types %in% c("CD8T","CD4T","T_other")) %>%
  mutate(cell_types = factor(cell_types, levels = c("CD8T","CD4T","T_other"))) %>%
  dplyr::select(PD1, CD8T, CD4T, T_other, center = cell_types, Patient.ID, bt3)

# org.Hs.eg.db masks dplyr::select(), which brms/rstan's internals rely on.
if ("package:org.Hs.eg.db" %in% search()) detach("package:org.Hs.eg.db", unload = TRUE)

fit_full <- brm(
  PD1 ~ 0 + center + center:CD8T + center:CD4T + center:T_other +
    (1 + CD8T + CD4T + T_other || bt3) + (1 | Patient.ID),
  data = df1, family = student(), iter = 4000, chains = 4, cores = 4, seed = 123,
  control = list(adapt_delta = 0.95)
)
```

##### Convergence Check (Rhat < 1.01 threshold)
```{r}
summary(fit_full)$fixed
cat("Max Rhat across all parameters:", max(brms::rhat(fit_full), na.rm = TRUE), "\n")
```

##### Divergent Transitions
```{r}
np_full <- brms::nuts_params(fit_full)
n_divergent <- sum(np_full$Value[np_full$Parameter == "divergent__"])
n_post_warmup <- sum(fit_full$fit@sim$n_save - fit_full$fit@sim$warmup2)
cat("Divergent transitions:", n_divergent, "out of", n_post_warmup, "post-warmup iterations\n")
```

##### Posterior Predictive Check

Overlays draws from the model's posterior predictive distribution on the observed PD1
distribution.
```{r}
p_ppc_full <- pp_check(fit_full, ndraws = 100) +
  labs(title = "Posterior predictive check: fit_full (PD1 ~ CD8T + CD4T + T_other)")
ggsave(file.path(cycif_fig_dir, "misc", "qc", "brm_tcell_pd1_pp_check.pdf"), p_ppc_full, width = 7, height = 5)
p_ppc_full
```

##### Extract Coefficients

Point estimate + 95% credible interval per center, per neighbor-type slope, parsed from
`fixef()` rows named `"center{level}:{term}"`.
```{r}
coef_df <- as.data.frame(fixef(fit_full)) %>%
  rownames_to_column("param") %>%
  filter(str_detect(param, ":")) %>%
  mutate(
    center   = str_remove(str_extract(param, "^center[^:]+"), "^center"),
    neighbor = str_remove(param, "^center[^:]+:")
  ) %>%
  rename(slope = Estimate, ci_low = Q2.5, ci_high = Q97.5) %>%
  mutate(
    center   = factor(center, levels = c("CD4T", "CD8T", "T_other")),
    neighbor = factor(neighbor, levels = c("CD4T", "CD8T", "T_other"))
  ) %>%
  select(center, neighbor, slope, Est.Error, ci_low, ci_high)

coef_df
```
### [Fig 4L] brm_tcell_pd1_coef_plot_combined

Pairwise p-values between the three neighbor-density coefficients, within each center only.
Uses posterior draws directly: for each pair, a two-sided Bayesian p-value analog (2x the
smaller tail probability of the draw-level difference being on one side of zero). One
`facet_wrap(~center)` plot with a single global `y_base`, so brackets align across panels.
```{r}
draws_full <- as_draws_df(fit_full)

comparisons <- list(c("CD4T", "CD8T"), c("CD4T", "T_other"), c("CD8T", "T_other"))

coef_pval_df <- purrr::map_dfr(levels(coef_df$center), function(ctr) {
  map_dfr(comparisons, function(comp) {
    col1 <- paste0("b_center", ctr, ":", comp[1])
    col2 <- paste0("b_center", ctr, ":", comp[2])
    diff <- draws_full[[col1]] - draws_full[[col2]]
    pval <- 2 * min(mean(diff > 0), mean(diff < 0))
    tibble(center = ctr, term1 = comp[1], term2 = comp[2], pval = pval)
  })
}) %>%
  mutate(center = factor(center, levels = levels(coef_df$center)),
         label = signif(pval, 2))

term_levels <- levels(coef_df$neighbor)
x_pos <- setNames(seq_along(term_levels), term_levels)
y_base <- max(coef_df$ci_high, na.rm = TRUE)
y_offsets <- c("CD4T_CD8T" = y_base * 1.15, "CD8T_T_other" = y_base * 1.15,
               "CD4T_T_other" = y_base * 1.35)

ann_df <- coef_pval_df %>%
  mutate(
    x1 = x_pos[term1] + 0.05,
    x2 = x_pos[term2] - 0.05,
    xmid = (x1 + x2) / 2,
    contrast = paste(term1, term2, sep = "_"),
    y = y_offsets[contrast]
  )

p_brm_combined <- ggplot(coef_df, aes(x = neighbor, y = slope, color = neighbor)) +
  geom_hline(yintercept = 0, linetype = 2, color = "gray50") +
  geom_pointrange(aes(ymin = ci_low, ymax = ci_high), size = 0.6, linewidth = 0.8) +
  geom_segment(data = ann_df, aes(x = x1, xend = x2, y = y, yend = y),
               inherit.aes = FALSE, color = "black", linewidth = 0.4) +
  geom_text(data = ann_df, aes(x = xmid, y = y * 1.02, label = label),
            inherit.aes = FALSE, size = 4, vjust = 0) +
  facet_wrap(~ center, nrow = 1) +
  coord_cartesian(ylim = c(min(coef_df$ci_low, na.rm = TRUE), y_base * 1.35 * 1.05)) +
  theme_bw(base_size = 14) +
  labs(
    title = "PD1 expression ~ neighboring T-cell density",
    # subtitle = "Point estimate + 95% CrI; brackets: pairwise posterior comparison within each center",
    x = "Neighboring T-cell type",
    y = "Posterior slope (fixed interaction coefficient)"
  ) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

p_brm_combined

ggsave(file.path(cycif_fig_dir, "spatial", "tcell_density", "pd1_tcell_density_slope", "brm_tcell_pd1_coef_plot_combined.pdf"),
       p_brm_combined, width = 5.5, height = 6, dpi = 300)
```

### [Fig 4M] correlation_tcell_pd1_density_ab_self (GrB)
```{r}
cst7.abs.log <- paste0(cst7.abs, ".log")

for(ab in cst7.abs.log[1]){
  lst.objs <- list()
  cell_types <- c("CD8T", "CD4T", "T_other")

  for (obj_name in names(datasets)) {
    obj <- datasets[[obj_name]]

    for (ct in cell_types) {
      key <- paste0(obj_name, ".", ct)


      x1 <- obj$PD1.log
      y1 <- obj[[ct]]
      z1 <- obj[[ab]]
      cl <- obj$cluster
      if (obj_name == ct) y1 <- y1 - 1

      set.seed(12345)
      x_scaled <- scale(x1)[, 1] + runif(length(x1), -0.5, 0.5)
      y_scaled <- scale(y1)[, 1]

      eps_x <- diff(range(x_scaled, na.rm = TRUE)) * 0.01
      eps_y <- diff(range(y_scaled, na.rm = TRUE)) * 0.01

      coords <- cbind(
        x_scaled + runif(length(x_scaled), -eps_x, eps_x),
        y_scaled + runif(length(y_scaled), -eps_y, eps_y)
      )

      knn <- get.knn(coords, k = 20)
      knn_indices <- cbind(1:nrow(coords), knn$nn.index)

      z1.smooth <- apply(knn_indices, 1, function(idx) {
        mean(z1[idx], na.rm = TRUE)
      })

      lst.objs[[key]] <- data.frame(
        center   = obj_name,
        neighbor = ct,
        PD1      = x1,
        density  = y1,
        ab       = z1.smooth,
        cluster  = cl
      )
    }
  }

  df <- do.call(rbind, lst.objs) %>%
    filter(center %in% c("CD8T","CD4T"),neighbor %in% c("CD8T","CD4T")) %>%
    mutate(center   = factor(center,   levels = c("CD8T", "CD4T")),
           neighbor = factor(neighbor, levels = c("CD8T", "CD4T")))

  q <- quantile(df$ab, c(0.01, 0.99), na.rm = TRUE)
  df$ab <- pmin(pmax(df$ab, q[1]), q[2])

  p_ab <- ggplot(df %>% arrange(ct), aes(density, PD1, colour = ab)) + # density_col
    geom_point(size = 0.5) +
    scale_color_viridis_c(
      option = "viridis",   # blue → green → yellow
      direction = 1,
      name = paste0(sub("\\.log","",ab), "\n(log exp)")
    ) +
    facet_grid(center ~ neighbor, switch = "y") +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.x = element_text(size = 12),
      strip.text.y.left = element_text(angle = 90, size = 12),
      axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 16,color = "black"),
      axis.text.y  = element_text(size = 15,color = "black"),
      axis.title.x = element_text(size = 17, margin = margin(t=2, r = 4),color = "black"),  # tighter gap to ticks
      axis.title.y = element_text(size = 17, margin = margin(r = 4),color = "black"),  # tighter gap to ticks
      plot.title   = element_text(size = 20, hjust = 0.5,color = "black"),
      legend.text  = element_text(size = 14,color = "black"),
      legend.title = element_text(size = 15,color = "black"),
      strip.text   = element_text(size = 16,color = "black"),
      legend.key.size = unit(1.2, "cm")
    ) +
    labs(x = "Number of neighboring T cells per CN", y = "PD1 expression in center T cells", title= paste0(sub("\\.log","",ab)," expression"))


  ggsave(file.path(cycif_fig_dir, "spatial", "tcell_density", "exps_in_self", paste0("correlation_tcell_pd1_density_",sub("\\.log","",ab),"_v2.pdf")),
         p_ab, width = 6, height = 5, units="in", dpi=300)
}
```
### [Fig 6J, EDF10A] correlation_tcell_pd1_density_ab_tme (PDL1, pTBK1, gH2AX, PD1)
```{r}
for(ab in cst7.abs[c(3,4,9,10)]){
  lst.objs <- list()
  cell_types <- c("CD8T", "CD4T", "T_other")

  for (obj_name in names(datasets)) {
    obj <- datasets[[obj_name]]

    for (ct in cell_types) {
      key <- paste0(obj_name, ".", ct)


      x1 <- obj$PD1.log
      y1 <- obj[[ct]]
      z1 <- obj[[ab]]
      cl <- obj$cluster
      if (obj_name == ct) y1 <- y1 - 1

      set.seed(12345)
      x_scaled <- scale(x1)[, 1] + runif(length(x1), -0.5, 0.5)
      y_scaled <- scale(y1)[, 1]

      eps_x <- diff(range(x_scaled, na.rm = TRUE)) * 0.01
      eps_y <- diff(range(y_scaled, na.rm = TRUE)) * 0.01

      coords <- cbind(
        x_scaled + runif(length(x_scaled), -eps_x, eps_x),
        y_scaled + runif(length(y_scaled), -eps_y, eps_y)
      )

      knn <- get.knn(coords, k = 20)
      knn_indices <- cbind(1:nrow(coords), knn$nn.index)

      z1.smooth <- apply(knn_indices, 1, function(idx) {
        mean(z1[idx], na.rm = TRUE)
      })

      lst.objs[[key]] <- data.frame(
        center   = obj_name,
        neighbor = ct,
        PD1      = x1,
        density  = y1,
        ab       = z1.smooth,
        cluster  = cl
      )
    }
  }

  df <- do.call(rbind, lst.objs) %>%
    filter(center %in% c("CD8T","CD4T"),neighbor %in% c("CD8T","CD4T")) %>%
    mutate(center   = factor(center,   levels = c("CD8T", "CD4T")),
           neighbor = factor(neighbor, levels = c("CD8T", "CD4T")))

  q <- quantile(df$ab, c(0.025, 0.975), na.rm = TRUE)
  df$ab <- pmin(pmax(df$ab, q[1]), q[2])

  p_ab <- ggplot(df %>% arrange(ct), aes(density, PD1, colour = ab)) + # density_col
    geom_point(size = 0.5) +
    scale_color_viridis_c(
      option = "viridis",   # blue → green → yellow
      direction = 1,
      name = paste0(ab, "\n(log exp)")
    ) +
    facet_grid(center ~ neighbor, switch = "y") +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.x = element_text(size = 12),
      strip.text.y.left = element_text(angle = 90, size = 12),
      axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 16,color = "black"),
      axis.text.y  = element_text(size = 15,color = "black"),
      axis.title.x = element_text(size = 17, margin = margin(t=2, r = 4),color = "black"),  # tighter gap to ticks
      axis.title.y = element_text(size = 17, margin = margin(r = 4),color = "black"),  # tighter gap to ticks
      plot.title   = element_text(size = 20, hjust = 0.5,color = "black"),
      legend.text  = element_text(size = 14,color = "black"),
      legend.title = element_text(size = 15,color = "black"),
      strip.text   = element_text(size = 16,color = "black"),
      legend.key.size = unit(1.2, "cm"),
      legend.position = "none"
    ) +
    labs(x = "Number of neighboring T cells per CN", y = "PD1 expression in center T cells", title= paste0(ab," expression (CN)"))


  ggsave(file.path(cycif_fig_dir, "spatial", "tcell_density", "exps_in_tme", paste0("correlation_tcell_pd1_density_",ab,"_v2.pdf")),
         p_ab, width = 5, height = 5, units="in", dpi=300)
}
```

## ---- TS Ratio Violin Plots ----

### Predicted Tumor/Stroma Classification

`is_pred_stroma` is a logistic regression predicting stroma membership from local cell-type
composition (`frnn7.cmb`'s own `uniq.cts` columns), trained on a size- and distance-stratified
subsample of baseline-sample cells, then applied to every cell in `frnn7.cmb`.
```{r}
is.ge10.7 <- rowSums(frnn7.rcn) >= 10

frnn7.cmb$dist <- unlist(sapply(lst.dists7, function(ld) ld$dist$dist))
frnn7.cmb <- frnn7.cmb %>% mutate(Tumor = Tumor_panCK + Tumor_neg)

train_pool <- frnn7.cmb %>%
  filter(is.ge10.7) %>%
  filter(grepl("-1$", sample)) %>%
  mutate(is_stroma = ifelse(dist > 0.5, 1, 0))

min_cells <- train_pool %>%
  count(sample) %>%
  pull(n) %>%
  sort() %>%
  `[`(3)

prop_within_min <- 0.9
num_within_target  <- round(min_cells * prop_within_min)
num_outside_target <- min_cells - num_within_target

set.seed(123)
train_set <- train_pool %>%
  group_by(sample) %>%
  group_modify(~ {
    df <- .x
    within_idx  <- which(df$dist >= -250 & df$dist <= 250)
    outside_idx <- which(df$dist < -250 | df$dist > 250)

    n_within_avail  <- length(within_idx)
    n_outside_avail <- length(outside_idx)

    n_outside_sel <- min(n_outside_avail, num_outside_target)
    n_within_sel  <- min(n_within_avail, num_within_target)

    if (n_outside_sel < num_outside_target) {
      additional_needed <- num_outside_target - n_outside_sel
      n_within_sel <- min(n_within_avail, n_within_sel + additional_needed)
    }

    total_selected <- n_within_sel + n_outside_sel
    if (total_selected < min_cells) {
      remaining <- min_cells - total_selected
      extra_outside <- min(n_outside_avail - n_outside_sel, remaining)
      n_outside_sel <- n_outside_sel + extra_outside
      remaining <- min_cells - (n_within_sel + n_outside_sel)
      if (remaining > 0) {
        n_within_sel <- min(n_within_avail, n_within_sel + remaining)
      }
    }

    selected_within  <- if (n_within_sel > 0)  sample(within_idx, n_within_sel)   else integer(0)
    selected_outside <- if (n_outside_sel > 0) sample(outside_idx, n_outside_sel) else integer(0)

    df$is_selected <- FALSE
    df$is_selected[c(selected_within, selected_outside)] <- TRUE
    df
  }) %>%
  ungroup() %>%
  filter(is_selected)

logit_model <- glm(
  is_stroma ~ Tumor + Fibro + CD8T + CD4T + T_other +
    Mac_CD68 + Mac_CD163 + Mac_CD68_CD163 + Immune_other,
  data = train_set, family = binomial
)

frnn7.cmb$predicted_prob_stroma <- predict(logit_model, newdata = frnn7.cmb, type = "response")
frnn7.cmb$is_pred_stroma <- as.numeric(frnn7.cmb$predicted_prob_stroma > 0.5)
```

### [EDF8A] Predicted Tumor/Stroma Classification -- Example ROIs

4 example ROIs (`1013-1`, `1013-2` = BS/BX2; `3002-1`, `3002-2` = BS/BX2). Per-sample
`wr <- this.cts %in% uniq.cts` filters `cs7[[nm]]`'s cells to the same cell-type set
`frnn7.cmb` was itself filtered to, so `tmp.all` (rows of `frnn7.cmb` for that sample) and the
`wr`-filtered per-cell vectors below are assumed row-order-aligned -- verify this alignment
holds before trusting the rendered output.
```{r}
this.nms <- c("1013-1","1013-2","3002-1","3002-2")
max.y <- cyApply(cs7[this.nms],function(cy)max(xys(cy)$Y_centroid))
coords <- list(`1013-1`=list(x=8660,y=4963),
             `1013-2`=list(x=5640,y=6724),
             `3002-1`=list(x=13326,y=15142),
             `3002-2`=list(x=11119,y=16249))
locs <- lapply(seq(coords),function(i)list(x=coords[[i]]$x,y=max.y[[i]]-coords[[i]]$y))
names(locs) <- names(coords)

library(sf)

make_square <- function(cx, cy, h) {
  coords <- matrix(c(
    cx - h, cy - h,
    cx + h, cy - h,
    cx + h, cy + h,
    cx - h, cy + h,
    cx - h, cy - h
  ), ncol = 2, byrow = TRUE)
  st_polygon(list(coords))
}

h <- 400  # ROI half-width

for(nm in this.nms){
  tmp.all <- frnn7.cmb %>% filter(sample == nm)
  is.str.1 <- factor(c("Tumor","Stroma")[tmp.all$is_pred_stroma + 1], levels = c("Tumor","Stroma"))

  x1 <- cs7[[nm]]

  this.cts <- cell_types(x1)$cell_types
  this.cts <- as.character(this.cts)
  this.cts[is.na(this.cts)] <- "NA"

  wr <- this.cts %in% uniq.cts

  xy <- xys(x1)
  xy$Y_centroid <- max(xy$Y_centroid) - xy$Y_centroid
  xy.sf <- st_as_sf(xy[wr,],coords=c("X_centroid","Y_centroid"))

  obj <- lst.dists7[[nm]]
  this.cts1 <- this.cts[wr]
  this.b <- do.call(rbind,obj$borders)

  cols <- c(Tumor=2, Stroma=4)

  xy1 <- locs[[nm]]
  squares <- st_sfc(make_square(xy1$x, xy1$y, h),crs = st_crs(xy.sf))
  sq_union <- st_union(squares)

  inside <- lengths(st_intersects(xy.sf, sq_union)) > 0
  xy_sub <- xy.sf[inside, ]
  col_sub <- cols[is.str.1][inside]
  cts_sub <- this.cts1[inside]
  ct_col_sub <- ct.cols7[cts_sub]

  poly_inside <- lengths(st_intersects(this.b[[1]], sq_union)) > 0
  b_clipped <- st_intersection(this.b[[1]], sq_union)

  if(!dir.exists(file.path(cycif_fig_dir, "spatial", "tumor_stroma", "classification", "logit"))){
    dir.create(file.path(cycif_fig_dir, "spatial", "tumor_stroma", "classification", "logit"),recursive=TRUE)
  }

  pdf(file.path(cycif_fig_dir, "spatial", "tumor_stroma", "classification", "logit", paste0("ts_",nm,"_roi_logit.pdf")), width=7, height=7)
  par(mar=c(1,1,1,1))
  plot(xy_sub, pch = 20, col = col_sub, asp = 1, cex=1, main = "", reset = FALSE)
  plot(squares, border = "black", add = TRUE, lwd = 2)
  plot(b_clipped, border="grey20", add=TRUE, lwd=2)
  dev.off()

  pdf(file.path(cycif_fig_dir, "spatial", "tumor_stroma", "classification", "logit", paste0("ts_",nm,"_roi_cts.pdf")), width=7, height=7)
  par(mar=c(1,1,1,1))
  plot(xy_sub, pch = 20, col = ct_col_sub, asp = 1, cex=1, main = "", reset = FALSE)
  plot(squares, border = "black", add = TRUE, lwd = 2)
  plot(b_clipped, border="grey20", add=TRUE, lwd=2)
  dev.off()
}
```

### [EDF8G] Tumor/Stroma Classification on Macrophage UMAP

```{r}
frnn7.cmb.mac.ts <- frnn7.cmb %>%
  filter(cell_types %in% c("Mac_CD68","Mac_CD163","Mac_CD68_CD163")) %>%
  filter(is.sele.mac) %>%
  cbind(dt.mac.1 %>% dplyr::select(umap_x,umap_y)) %>%
  mutate(cluster = dt.mac.1 %>% pull(cluster)) %>%
  mutate(is_stroma.1 = factor(c("Tumor","Stroma")[is_pred_stroma + 1], levels = c("Tumor","Stroma")))

p_full <- umap_base(frnn7.cmb.mac.ts, aes(x = umap_x, y = umap_y, col = is_stroma.1)) +
  umap_styling_num() +
  scale_color_manual(values=c(Tumor=2,Stroma=4))
p_main <- p_full + theme(legend.position="none")
ggsave(file.path(cycif_fig_dir, "spatial", "tumor_stroma", "classification", "umap_ts_class_mac.png"),
   p_main, width = 7, height = 7, units = "in", dpi = 300)
save_legend_pdf(p_full, file.path(cycif_fig_dir, "spatial", "tumor_stroma", "classification", "umap_ts_class_legend_mac.pdf"), width = 4, height = 6)
```

Per-macrophage-cluster Tumor/Stroma composition -- each cluster's Tumor+Stroma proportions
sum to 1.
```{r}
tab <- apply(table(frnn7.cmb.mac.ts$cluster, frnn7.cmb.mac.ts$is_stroma.1), 1, function(x) x/sum(x))

pdf(file.path(cycif_fig_dir, "spatial", "tumor_stroma", "classification", "ts_ratio_enrichment_in_niche_mac_barplot.pdf"), width = 5, height = 7)
par(mar = c(5, 5, 4, 10))
barplot(tab[, rev(seq_len(ncol(tab)))], beside = FALSE, col = c(2, 4),
        horiz = TRUE,
        main = "",
        xlab = "Tumor-Stroma proportion", ylab = "", las = 2)
legend(par()$usr[2] + .1, par()$usr[4] - 1, rownames(tab), fill = c(2, 4), xpd = TRUE, border = TRUE)
dev.off()
```

### [Fig 6C top] TME Cluster Composition by Tumor vs Stroma

Joins `frnn7.cmb$is_pred_stroma` (computed above) and `frnn7.cmb$cluster_tme` (the TME
cluster assignment from file 3) at the per-cell level, since both columns already live on
the same `frnn7.cmb` table at this point in the notebook.
```{r}
freq.ts <- frnn7.cmb %>%
  filter(!is.na(cluster_tme) & !is.na(is_pred_stroma)) %>%
  mutate(is_stroma = factor(is_pred_stroma, levels = c(0,1), labels = c("Tumor","Stroma"))) %>%
  group_by(is_stroma, cluster_tme) %>%
  summarize(n = n(), .groups = "drop") %>%
  group_by(cluster_tme) %>%
  mutate(freq = n / sum(n))

p <- ggplot(freq.ts %>% mutate(cluster_tme = factor(cluster_tme, levels = rev(paste0("TME_",1:17)))),
            aes(x = cluster_tme, y = freq, fill = is_stroma)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = c("Stroma" = 4, "Tumor" = 2), name = "is_stroma") +
  theme_minimal() +
  xlab("Cluster") +
  ylab("Frequency") +
  ggtitle("Cluster Frequencies by Tumor vs Stroma") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 10),
    legend.position="top",
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.line = element_blank(),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

ggsave(file.path(cycif_fig_dir, "umap", "tme", "expression_by_cluster", "umap_tme_cluster_barplot.pdf"), p, width = 3, height = 7, units = "in", dpi = 300)
```

### [EDF9E left] Tumor vs Stroma on TME UMAP

`dt.tme.1` corresponds row-for-row to `frnn7.cmb[is.sele.tme, ]`.
```{r}
is.stroma.tme <- factor(frnn7.cmb$is_pred_stroma[is.sele.tme], levels = c(0,1), labels = c("Tumor","Stroma"))

dt.tme.1 <- dt.tme.1 %>% mutate(is_stroma = is.stroma.tme)

p9_full <- umap_base(dt.tme.1, aes(x = umap_x, y = umap_y, col = is_stroma), point_size = 0.8) +
  scale_color_manual(values = c("Stroma" = 4, "Tumor" = 2), name = "is_stroma") +
  ggtitle("Tumor vs Stroma")
p9 <- p9_full + theme(legend.position = "none")
ggsave(file.path(cycif_fig_dir, "umap", "tme", "metadata", "umap_tme_stroma.png"), p9, width = 7, height = 7, units = "in", dpi = 300)
save_legend_pdf(p9_full, file.path(cycif_fig_dir, "umap", "tme", "metadata", "umap_tme_stroma_legend.pdf"), width = 4, height = 6)
```

### [EDF8B] Delaunay contact–based tumor/stroma mixing score

Per-cell `X_centroid`/`Y_centroid` coordinates aren't carried on `frnn7.cmb` itself, so
they're pulled from the per-sample `cs7[[sample_id]]` objects via `xys()`.

```{r}
library(deldir)

# Fixed edge-length cutoff (X_centroid/Y_centroid, and thus edge lengths, are in pixel units).
edge_len_max <- 20 / 0.65                     # 30.77 px (~20um); set NULL to disable
n_perm <- 0                                   # set >0 (e.g., 1000) to run permutations

sample_ids <- unique(frnn7.cmb$sample)

results_list <- list()

for(i in seq_along(sample_ids)){
  sample_id <- sample_ids[i]
  cat("Processing:", sample_id, "(", i, "of", length(sample_ids), ")\n")

  # --- subset to this sample ---
  df_this <- frnn7.cmb %>% filter(sample == sample_id)
  this.cs <- cs7[[sample_id]]
  this.cts <- cell_types(this.cs)
  this.xys <- xys(this.cs)

  cts.xys <- cbind(this.cts, this.xys) %>% filter(!is.na(cell_types))

  df_s <- df_this %>%
    mutate(type_bin = ifelse(is_pred_stroma == 1, "stroma", "tumor")) %>%
    cbind(data.frame(x = cts.xys$X_centroid, y = cts.xys$Y_centroid)) %>%
    select(x, y, type_bin)

  if (nrow(df_s) < 3) {
    results_list[[i]] <- tibble(
      sample = sample_id,
      n_cells = nrow(df_s),
      n_edges = NA,
      tumor_frac = NA,
      mix_raw = NA,
      exp_hetero = NA,
      mix_norm = NA,
      p_perm = NA
    )
    next
  }

  # --- Delaunay triangulation ---
  tri <- deldir(df_s$x, df_s$y)
  edges <- as_tibble(tri$delsgs[, c("ind1","ind2","x1","y1","x2","y2")]) %>%
    mutate(len = sqrt((x1 - x2)^2 + (y1 - y2)^2))

  if (!is.null(edge_len_max)) {
    edges <- filter(edges, len <= edge_len_max)
  }

  if (nrow(edges) == 0) {
    results_list[[i]] <- tibble(
      sample = sample_id,
      n_cells = nrow(df_s),
      n_edges = 0,
      tumor_frac = mean(df_s$type_bin == "tumor"),
      mix_raw = NA,
      exp_hetero = NA,
      mix_norm = NA,
      p_perm = NA
    )
    next
  }

  # --- classify edges ---
  type1  <- df_s$type_bin[edges$ind1]
  type2  <- df_s$type_bin[edges$ind2]
  hetero <- (type1 != type2)

  # --- mixing scores ---
  mix_raw <- mean(hetero, na.rm = TRUE)
  pT      <- mean(df_s$type_bin == "tumor", na.rm = TRUE)
  pS      <- 1 - pT
  exp_ht  <- 2 * pT * pS
  mix_norm <- if (exp_ht > 0) mix_raw / exp_ht else NA_real_

  # --- save results ---
  results_list[[i]] <- tibble(
    sample = sample_id,
    n_cells = nrow(df_s),
    n_edges = nrow(edges),
    tumor_frac = pT,
    mix_raw = mix_raw,
    exp_hetero = exp_ht,
    mix_norm = mix_norm
  )
}

results <- bind_rows(results_list)
r1 <- results %>%
  left_join(pd7 %>% select(id,sample, BRCA, BOR, bt3, TimePoint, Patient.ID), by = c("sample"="id"))
p <- plot_ct(r1,y_var = "mix_norm", y_label = "Delaunay mixing score (norm)",
  title = "Delaunay mixing score (norm)",
  pt.cols = pt.cols)$plot

ggsave(file.path(cycif_fig_dir, "spatial", "tumor_stroma", "delaunay", "delaunay_mixing_score_20um.pdf"),p,width=5,height=7)
```

### TS log-ratio data prep (per-cell-type, all timepoints)

```{r}
rts.all <- frnn7.cmb %>%
  filter(cell_types %in% uniq.cts[-(1:2)]) %>%
  mutate(is_stroma = is_pred_stroma) %>%
  filter(!is.na(is_stroma)) %>%
  group_by(sample) %>%
  summarize(npos=sum(is_stroma == 1),
            nneg=sum(is_stroma == 0),
            ntot=n()) %>%
  # Robustness floor: both the stroma count (npos) and the tumor count (nneg) must
  # independently be >= 10, not just their sum (ntot) -- a sample with e.g. npos=19, nneg=1
  # would pass an ntot>=10-only check but its ratio is driven by a single cell.
  mutate(ratio_all = ifelse(npos >= 10 & nneg >= 10, npos/nneg, NA_real_)) %>%
  dplyr::select(ratio_all) %>%
  unlist()
names(rts.all) <- nms7

rts <- list()
for(ct in uniq.cts[3:11]){
  tmp <- frnn7.cmb %>%
    filter(cell_types == ct) %>%
    mutate(is_stroma = is_pred_stroma) %>%
    filter(!is.na(is_stroma)) %>%
    group_by(sample) %>%
    summarize(npos=sum(is_stroma == 1),
              nneg=sum(is_stroma == 0),
              ntot=n()) %>%
    mutate(ratio = ifelse(npos >= 10 & nneg >= 10, npos/nneg, NA_real_)) %>%
    mutate(rts.all = rts.all[as.character(sample)]) %>%
    mutate(rratio = ratio/rts.all) %>%
    mutate(log_ratio = log2(rratio)) %>%
    mutate(cell_type=ct)
  rts[[ct]] <- tmp
}

log.rts <- do.call(rbind,rts) %>%
  dplyr::select(sample,cell_type,log_ratio,ntot) %>%
  # npos>=10 & nneg>=10 (enforced above, on both this cell type's split and the sample-wide
  # rts.all baseline) is the per-side guard against a ratio driven by a handful of cells.
  filter(!is.na(log_ratio)) %>%
  mutate(cell_type = factor(cell_type,levels=uniq.cts[3:11])) %>%
  left_join(pd7 %>% dplyr::select(id,Patient.ID,BRCA,PFS_days,bt3,BOR,TimePoint),by=c("sample"="id"))

log.rts$log_ratio[log.rts$log_ratio==Inf] <- 6
log.rts$log_ratio[log.rts$log_ratio==-Inf] <- -6
```

### [EDF8C] TS Ratio for Macrophages

```{r}
log.rts.mac <- log.rts %>%
  filter(cell_type %in% uniq.cts[7:9]) %>%
  mutate(cell_type=factor(cell_type,levels=uniq.cts[c(7,9,8)])) %>%
  filter(!is.na(cell_type)) %>%
  arrange(sample)

log.rts.mac.1 <- log.rts.mac %>% dplyr::select(-ntot) %>% spread(key = cell_type, value = log_ratio)

# 3 paired Wilcoxon signed-rank tests comparing TS log-ratio across the 3 macrophage subtypes
# (moved from paired t-test, no normality justification for a t-test here).
pvals <- rep(NA,3)
pairs3 <- list(c(1,2),c(1,3),c(2,3))
for(i in 1:3){
  idx <- uniq.cts[c(7,9,8)][pairs3[[i]]]
  p <- tryCatch(
    suppressWarnings(wilcox.test(log.rts.mac.1[[idx[1]]],log.rts.mac.1[[idx[2]]],paired=T)$p.value),
    error = function(e) NA_real_
  )
  pvals[i] <- signif(p,3)
}

# Bracket heights derived from the data range (points + per-sample connecting lines), not
# fixed constants -- adjacent pair (1,2)/(2,3) share the lowest tier, the full-span (1,3)
# pair sits one tier above so its bracket doesn't overlap the other two.
y_range <- range(log.rts.mac$log_ratio, na.rm = TRUE)
y_span  <- diff(y_range)
bracket_step <- 0.08 * y_span
bracket_y0   <- y_range[2] + 0.04 * y_span
tier <- c(0, 1, 0)  # (1,2), (1,3), (2,3)

df.pvals <- data.frame(
  x1_full = sapply(pairs3, `[`, 1),
  x2_full = sapply(pairs3, `[`, 2),
  y  = bracket_y0 + tier * bracket_step,
  pval = as.character(pvals)
) %>%
  mutate(
    xmid = (x1_full + x2_full) / 2,
    # Shorten each bracket by ~5% of its span (inset symmetrically).
    x1 = x1_full + 0.025 * (x2_full - x1_full),
    x2 = x2_full - 0.025 * (x2_full - x1_full)
  )

df.mid <- log.rts.mac %>%
  group_by(cell_type) %>%
  summarize(log_ratio=median(log_ratio,na.rm=T)) %>%
  mutate(sample="mid") %>%
  mutate(x=c(1,2,3))

p <- ggplot(log.rts.mac,aes(cell_type,log_ratio)) +
  geom_violin(aes(fill=cell_type)) +
  scale_fill_manual(values=ct.cols7w) +
  geom_point(size=.5,alpha=.8) +
  geom_point(data=df.mid,aes(x=x,y=log_ratio),size=12,shape=95) +
  geom_line(aes(group=sample),alpha=.5,color="grey30") +
  geom_hline(yintercept = 0) +
  ggtitle("log(Stroma/Tumor) ratio") +
  geom_segment(data=df.pvals,aes(x=x1,xend=x2,y=y,yend=y),inherit.aes=FALSE) +
  geom_text(data=df.pvals,aes(x=xmid,y=y,label=pval),vjust=-0.3,inherit.aes=FALSE)+
  coord_cartesian(ylim = c(y_range[1], bracket_y0 + (max(tier)+1) * bracket_step)) +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position="none") +
  xlab("") +
  ylab("log2 (TS ratio)")

out <- log.rts.mac %>% filter(cell_type == "Mac_CD68") %>% arrange(desc(log_ratio))

fn <- paste0("ts_ratio_mac.pdf")
pdf(file.path(cycif_fig_dir, "spatial", "tumor_stroma", "ratio", fn),width=3,height=7)
print(p)
dev.off()
```

### [EDF8E] ts_ratio_all_v3

TS log-ratio across all cell types (`uniq.cts[3:11]`, all time points), one-sample deviation
from 0 tested per cell type (Wilcoxon signed-rank, moved from one-sample t-test, no normality
justification for a t-test here), cell types ordered by median log-ratio.
```{r}
rts.p <- sapply(uniq.cts[3:11], function(uc) {
  tryCatch(
    suppressWarnings(wilcox.test(log.rts %>% filter(cell_type == uc) %>% pull(log_ratio), mu = 0, alternative = "two.sided")$p.value),
    error = function(e) NA_real_
  )
})
names(rts.p) <- uniq.cts[3:11]

mid.log.rts <- log.rts %>%
  group_by(cell_type) %>%
  summarise(log_ratio = median(log_ratio, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(pval = rts.p[as.character(cell_type)])

y.symbol <- max(log.rts$log_ratio, na.rm = TRUE) * 1.05

asterisks <- mid.log.rts %>%
  mutate(label = case_when(
      pval < 0.01 ~ "**",
      pval < 0.05 ~ "*",
      TRUE ~ ""),
    y = y.symbol)

sorted.cts <- sort(tapply(log.rts$log_ratio, log.rts$cell_type, median, na.rm = TRUE)) %>% names()

log.rts.all <- log.rts %>%
  mutate(cell_type = factor(cell_type, levels = sorted.cts)) %>%
  filter(!is.na(cell_type))

mid.log.rts <- mid.log.rts %>%
  mutate(cell_type = factor(cell_type, levels = sorted.cts))

asterisks <- asterisks %>%
  mutate(cell_type = factor(cell_type, levels = sorted.cts))

p <- ggplot(log.rts.all, aes(cell_type, log_ratio)) +
  geom_violin(aes(fill = cell_type), show.legend = FALSE) +
  scale_fill_manual(values = ct.cols7w) +
  geom_point(data = mid.log.rts, aes(x = cell_type, y = log_ratio), size = 12, shape = 95, color = "grey20") +
  geom_jitter(color = 1, width = 0.3, size = .8) +
  geom_hline(yintercept = 0) +
  geom_text(data = asterisks, aes(x = cell_type, y = y, label = label),
            inherit.aes = FALSE, size = 6, color = "black") +
  ggtitle("log TS-ratio of all cell types (all time points)") +
  theme_minimal(base_size = 16) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "top"
  ) +
  xlab("") +
  ylab("log2 TS-ratio")

pdf(file.path(cycif_fig_dir, "spatial", "tumor_stroma", "ratio", "ts_ratio_all_v3.pdf"), width = 6, height = 9)
print(p)
dev.off()
```
