---
author: "Kenichi Shimada"
date: "2023-03-10"
title: "data integration"
---

Harmonizes shared markers across IO360 and DSP protein data, subsets shared clinical
cohorts, computes cross-platform correlation summaries, compares CyCIF with other
technologies, and fits integrated CD8 signal models.

**Required run order: `1` -> `2` -> `7` (this notebook).**
(This notebook loads IO360 and DSP Protein checkpoints produced by `1` and `2`.)

## ---- Initial Setup ----

### Load Configuration
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Bootstrap
```{r bootstrap}
source(file.path(root_dir, "functions", "bootstrap.R"))
initialize_project()

if (!requireNamespace("CycifAnalyzeR", quietly = TRUE)) {
  remotes::install_github("<org>/CycifAnalyzeR")
}
library(CycifAnalyzeR)
source(file.path(root_dir,"functions","plotting.R"))
source(file.path(source_dir, "io.R"))
```

## ---- Define Shared Inputs for Retained Figures ----

### Load Expression Data

Both loaded from precomputed checkpoints (`io360_expression_input_data.rda`,
`dsp_protein_expression_input_data.rda`) rather than independently re-reading the raw
source files -- `df.dsp.all` already reflects the same patient-exclusion criteria applied
upstream when the checkpoint was built.
```{r}
load_io360_expression_data()
io360 <- df.io360

load(file.path(protein_obj_dir, "dsp_protein_expression_input_data.rda")) # df.dsp.all, all.prots
```

### Harmonize Shared Markers
```{r}
prot.lookup <- load_dsp_protein_lookup()
prot.lookup$SYMBOL[37] <- "TMEM173"

hugos <- unlist(strsplit(prot.lookup$SYMBOL, ";"))
gns.io <- colnames(io360)[-1]
all.comm <- intersect(hugos, gns.io)

syms <- prot.lookup$SYMBOL
lns <- sapply(strsplit(syms, ";"), length)
all.syms <- unlist(strsplit(syms, ";"))
df.syms <- data.frame(idx = rep(seq_along(syms), lns), symbol = all.syms)

all.comm.syms.idx <- df.syms %>%
  filter(symbol %in% all.comm) %>%
  pull(idx) %>%
  unique()
all.comm.syms <- prot.lookup$Protein[all.comm.syms.idx]
```

### Load Clinical Metadata and Subset Samples
```{r}
clin.all <- load_clinical_metadata("clinical_metadata.rds") %>%
  mutate(bt3 = paste0(BRCA, "_", TimePoint)) %>%
  mutate(bt3 = factor(bt3, levels = c("MUT_BS", "MUT_BX2", "MUT_BX3", "WT_BS", "WT_BX2", "WT_BX3")))

io.1 <- io360 %>%
  select(1, !!!syms(all.comm)) %>%
  mutate(PT = factor(PT)) %>%
  filter(PT %in% clin.all$PT)

prot.1 <- df.dsp.all %>%
  select(PT, segment, !!!syms(all.comm.syms)) %>%
  mutate(PT = factor(PT)) %>%
  filter(PT %in% clin.all$PT)

prot.1s <- prot.1 %>%
  filter(segment == "Stroma") %>%
  group_by(PT) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))

prot.1t <- prot.1 %>%
  filter(segment == "Tumor") %>%
  group_by(PT) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))

prot.1m <- prot.1 %>%
  group_by(PT) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

### Define Patient Colors
```{r}
pd <- clin.all
n.pts <- pd %>% pull(Patient.ID) %>% unique() %>% length()

if (n.pts > 12) {
  set.seed(123)
  col1 <- sample(colorRampPalette(RColorBrewer::brewer.pal(11, "Spectral"))(n.pts))
} else {
  col1 <- RColorBrewer::brewer.pal(length(unique(pd$Patient.ID)), "Set3")
}

names(col1) <- unique(pd$Patient.ID)
pt.cols <- unique(col1[pd$Patient.ID])
names(pt.cols) <- unique(pd$Patient.ID)
```

## ---- Compare CyCIF with Other Technologies ----

### Prepare CyCIF and Cross-Platform Inputs
```{r}
cs5 <- readRDS(file.path(cycif_obj_dir, "cs5.rds"))
cs7 <- readRDS(file.path(cycif_obj_dir, "cs7.rds"))

excluded_ids <- c("1001-3", "1019-3")
used.pt5 <- names(cs5)[!names(cs5) %in% excluded_ids]
used.pt7 <- names(cs7)[!names(cs7) %in% excluded_ids]
cs5 <- cs5[used.pt5]
cs7 <- cs7[used.pt7]

tab5 <- table(cell_types(cs5))
tab7 <- table(cell_types(cs7))

n5 <- apply(tab5, 1, function(x) x / sum(x))
n7 <- apply(tab7, 1, function(x) x / sum(x))
colnames(n5) <- used.pt5
colnames(n7) <- used.pt7

cy5 <- n5
cy5["Mac_CD68", ] <- cy5["Mac_CD68", ] + cy5["Mac_CD68_CD163", ]
cy5["Mac_CD163", ] <- cy5["Mac_CD163", ] + cy5["Mac_CD68_CD163", ]
cy5["T_other", ] <- cy5["CD4T", ] + cy5["T_other", ] + cy5["CD8T", ]
rownames(cy5)[6] <- "T"
cy5 <- cy5[c("CD8T", "T", "Tumor_panCK", "Mac_CD68", "Mac_CD163"), ]
```

## ---- Integrated CD8 Signal ----

### Prepare CD8 Inputs
```{r}
i1 <- io360 %>%
  left_join(clin.all, by = "PT") %>%
  filter(!is.na(BRCA)) %>%
  group_by(Patient.ID) %>%
  mutate(
    max_bn = max(as.numeric(TimePoint)),
    text = ifelse(as.numeric(TimePoint) == max_bn, as.character(Patient.ID), "")
  )

i3 <- prot.1m %>%
  left_join(clin.all, by = "PT") %>%
  filter(!is.na(BRCA)) %>%
  group_by(Patient.ID) %>%
  mutate(
    max_bn = max(as.numeric(TimePoint)),
    text = ifelse(as.numeric(TimePoint) == max_bn, as.character(Patient.ID), "")
  )

i3t <- prot.1t %>%
  left_join(clin.all, by = "PT") %>%
  filter(!is.na(BRCA)) %>%
  group_by(Patient.ID) %>%
  mutate(
    max_bn = max(as.numeric(TimePoint)),
    text = ifelse(as.numeric(TimePoint) == max_bn, as.character(Patient.ID), "")
  )

i3s <- prot.1s %>%
  left_join(clin.all, by = "PT") %>%
  filter(!is.na(BRCA)) %>%
  group_by(Patient.ID) %>%
  mutate(
    max_bn = max(as.numeric(TimePoint)),
    text = ifelse(as.numeric(TimePoint) == max_bn, as.character(Patient.ID), "")
  )

i4 <- t(cy5) %>%
  as.data.frame() %>%
  rownames_to_column("PT") %>%
  left_join(clin.all, by = "PT") %>%
  filter(!is.na(BRCA)) %>%
  group_by(Patient.ID) %>%
  mutate(
    max_bn = max(as.numeric(TimePoint)),
    text = ifelse(as.numeric(TimePoint) == max_bn, as.character(Patient.ID), "")
  )
```

### [Fig 4D] Platform-Specific CD8 Signals
```{r}
dir.create(file.path(integrated_fig_dir, "CD8T"), recursive = TRUE, showWarnings = FALSE)

p1 <- plot_ct(
  i1,
  y_var = "CD8A",
  y_label = "CD8A Expression",
  title = "CD8A (IO360)",
  ymin = NULL,
  pt.cols = pt.cols
)

p3 <- plot_ct(
  i3,
  y_var = "CD8",
  y_label = "CD8 Expression",
  title = "CD8 (Protein)",
  ymin = NULL,
  pt.cols = pt.cols
)

p4 <- plot_ct(
  i4,
  y_var = "CD8T",
  y_label = "CD8+ T cell Frequency",
  title = "CD8T (CyCIF)",
  ymin = NULL,
  pt.cols = pt.cols
)

pdf(file.path(integrated_fig_dir, "CD8T", "cd8t_io360_v2.pdf"), width = 5, height = 7)
print(p1$plot)
dev.off()

pdf(file.path(integrated_fig_dir, "CD8T", "cd8t_prot_v2.pdf"), width = 5, height = 7)
print(p3$plot)
dev.off()

pdf(file.path(integrated_fig_dir, "CD8T", "cd8t_cycif_v2.pdf"), width = 5, height = 7)
print(p4$plot)
dev.off()
```

### [Fig 4E] Integrated CD8 Model
```{r}
i1.0 <- i1 %>% ungroup() %>% select(PT, CD8A) %>% dplyr::rename(CD8A_IO360 = CD8A)
i4.0 <- i4 %>% ungroup() %>% select(PT, CD8T) %>% dplyr::rename(CD8T_CyCIF = CD8T)

i1.0$CD8A_IO360 <- scale(i1.0$CD8A_IO360, center = TRUE, scale = TRUE)
i4.0$CD8T_CyCIF <- scale(i4.0$CD8T_CyCIF, center = TRUE, scale = TRUE)

raw.exp <- clin.all %>%
  select(PT) %>% unique() %>%
  left_join(i1.0, by = "PT") %>%
  left_join(i4.0, by = "PT") %>%
  gather(key = "tech", value = "CD8", CD8A_IO360:CD8T_CyCIF) %>%
  mutate(
    tech = factor(tech, levels = c("CD8A_IO360", "CD8T_CyCIF")),
    PT = factor(PT)
  )

lm1 <- lmer(CD8 ~ PT + (1 | tech), data = raw.exp)
coefs <- summary(lm1)$coefficient[, "Estimate"]
coefs1 <- coefs[1] + c(`PT1001-1` = 0, coefs[-1])

df.coef <- clin.all %>%
  left_join(data.frame(PT = sub("PT", "", names(coefs1)), CD8 = coefs1), by = "PT") %>%
  filter(!is.na(CD8)) %>%
  group_by(Patient.ID) %>%
  mutate(
    max_bn = max(as.numeric(TimePoint)),
    text = ifelse(as.numeric(TimePoint) == max_bn, as.character(Patient.ID), "")
  ) %>%
  ungroup()

p <- plot_ct(
  df.coef,
  y_var = "CD8",
  y_label = "CD8 (normalized)",
  title = "CD8 (IO360 & CyCIF)",
  pt.cols = pt.cols
)

pdf(file.path(integrated_fig_dir, "CD8T", "cd8_integrated_v2.pdf"), width = 5, height = 7)
print(p)
dev.off()
```

### [EDF5D] Difference Between BX2 and BS, BX3 and BS
```{r}
i1d <- i1 %>%
  select(Patient.ID, TimePoint, CD8A) %>%
  spread(TimePoint, CD8A) %>%
  mutate(
    dBX2 = BX2 - BS,
    dBX3 = BX3 - BS
  ) %>%
  select(Patient.ID, dBX2, dBX3)

i3td <- i3t %>%
  select(Patient.ID, TimePoint, CD8) %>%
  spread(TimePoint, CD8) %>%
  mutate(
    dBX2 = BX2 - BS,
    dBX3 = BX3 - BS
  ) %>%
  select(Patient.ID, dBX2, dBX3)

i3sd <- i3s %>%
  select(Patient.ID, TimePoint, CD8) %>%
  spread(TimePoint, CD8) %>%
  mutate(
    dBX2 = BX2 - BS,
    dBX3 = BX3 - BS
  ) %>%
  select(Patient.ID, dBX2, dBX3)

i4d <- i4 %>%
  select(Patient.ID, TimePoint, CD8T) %>%
  spread(TimePoint, CD8T) %>%
  mutate(
    dBX2 = BX2 - BS,
    dBX3 = BX3 - BS
  ) %>%
  select(Patient.ID, dBX2, dBX3)

i5d <- df.coef %>%
  select(Patient.ID, TimePoint, CD8) %>%
  spread(TimePoint, CD8) %>%
  mutate(
    dBX2 = BX2 - BS,
    dBX3 = BX3 - BS
  ) %>%
  select(Patient.ID, dBX2, dBX3)

d.exp <- clin.all %>%
  select(Patient.ID) %>% unique() %>%
  left_join(i1d, by = "Patient.ID") %>%
  left_join(i3td, by = "Patient.ID") %>%
  left_join(i3sd, by = "Patient.ID") %>%
  left_join(i4d, by = "Patient.ID") %>%
  left_join(i5d, by = "Patient.ID") %>%
  mutate(Patient.ID = factor(Patient.ID))

colnames(d.exp) <- c(
  "Patient.ID",
  "dBX2_IO360", "dBX3_IO360",
  "dBX2_Prot_Tumor", "dBX3_Prot_Tumor",
  "dBX2_Prot_Stroma", "dBX3_Prot_Stroma",
  "dBX2_CyCIF", "dBX3_CyCIF",
  "dBX2_integrated", "dBX3_integrated"
)

lev.pts <- levels(d.exp$Patient.ID)
d.exp <- as.matrix(d.exp[-1])
rownames(d.exp) <- lev.pts

sds <- sapply(1:5, function(j) {
  j1 <- j * 2 + (-1:0)
  sd(d.exp[, j1], na.rm = TRUE)
})

nd.exp <- d.exp / array(rep(sds, each = 2)[col(d.exp)], dim = dim(d.exp))

pdf(file.path(integrated_fig_dir, "CD8T", "diff_BX2_BX3_all_technologies_v4.pdf"), width = 7, height = 7)
h3(
  nd.exp[rev(seq_len(nrow(nd.exp))), ],
  Colv = NA,
  Rowv = NA,
  balanceColor = TRUE,
  scale = "none",
  margins = c(10, 10),
  main = "diff in CD8 from BS"
)
dev.off()
```

### [EDF5C] Heatmap of BX2/BS and BX3/BS Differences
```{r}
d.bx2 <- nd.exp[, grep("BX2", colnames(nd.exp))[1:4]]
d.bx3 <- nd.exp[, grep("BX3", colnames(nd.exp))[1:4]]

cor2 <- cor(d.bx2, use = "pairwise.complete.obs")
cor3 <- cor(d.bx3, use = "pairwise.complete.obs")

rownames(cor2) <- colnames(cor2) <- rownames(cor3) <- colnames(cor3) <- c("IO360", "DSP, Tumor", "DSP, Stroma", "CyCIF")

shared_colors <- colorRampPalette(c("navy", "white", "firebrick3"))(101)
col_fun <- circlize::colorRamp2(seq(-1, 1, length.out = length(shared_colors)), shared_colors)
lgd_at <- c(-1, -0.5, 0, 0.5, 1)

pdf(file.path(integrated_fig_dir, "CD8T", "correlation_BX2_v5_ComplexHeatmap.pdf"), width = 7, height = 7)
ComplexHeatmap::draw(
  Heatmap(
    as.matrix(cor2),
    name = "Correlation",
    col = col_fun,
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_dend = TRUE,
    show_column_dend = TRUE,
    heatmap_legend_param = list(
      at = lgd_at,
      labels = lgd_at,
      title = "r",
      legend_height = unit(35, "mm")
    )
  )
)
dev.off()

pdf(file.path(integrated_fig_dir, "CD8T", "correlation_BX3_v5_ComplexHeatmap.pdf"), width = 7, height = 7)
ComplexHeatmap::draw(
  Heatmap(
    as.matrix(cor3),
    name = "Correlation",
    col = col_fun,
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_dend = TRUE,
    show_column_dend = TRUE,
    heatmap_legend_param = list(
      at = lgd_at,
      labels = lgd_at,
      title = "r",
      legend_height = unit(35, "mm")
    )
  )
)
dev.off()
```
